---
title: "Pokémon Attack Stat GLM Analysis"
subtitle: "Comprehensive Generalized Linear Model Analysis for Attack Power"
author: "Jan Tychtl"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    code_folding: show
    fig_width: 10
    fig_height: 6
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center"
)
```

# Introduction

## Research Objective

This analysis focuses on **Attack** as the primary target variable. We aim to:

1. **Understand what drives Attack stats** — Which factors (generation, type, rarity, other stats) best explain variation in Attack?
2. **Quantify generation effects** — Is there evidence of power creep specifically in Attack values?
3. **Build predictive models** — Develop GLMs that accurately predict Attack from available features
4. **Compare model specifications** — Evaluate Gaussian vs Gamma GLMs, different link functions, and control variable sets

---

# 1. Setup and Data Loading

```{r libraries, message=FALSE}
# Core packages
library(tidyverse)
library(ggplot2)
library(gridExtra)

# Statistical modeling
library(car)          # VIF, ANOVA
library(MASS)         # Robust regression
library(lmtest)       # Breusch-Pagan test
library(broom)        # Tidy model outputs
library(glmnet)       # LASSO regularization

# Specialized tests
library(moments)      # Skewness and kurtosis
library(DescTools)    # Jonckheere-Terpstra test

# Visualization
library(corrplot)     # Correlation plots
library(scales)       # Scale functions for plots
```

```{r load-data}
# Load data using relative paths from the Rmd file location
pokemon <- read.csv("../data/data_clean.csv")
cat("Dataset loaded:", nrow(pokemon), "Pokémon\n")
```

```{r preprocess-data}
# Type name lookup tables (from pokemon.csv original names)
primary_type_lookup <- c(
  "0" = "Grass", "1" = "Fire", "2" = "Water", "3" = "Bug", 
  "4" = "Normal", "5" = "Poison", "6" = "Electric", "7" = "Ground",
  "8" = "Fairy", "9" = "Fighting", "10" = "Psychic", "11" = "Rock",
  "12" = "Ghost", "13" = "Ice", "14" = "Dragon", "15" = "Dark",
  "16" = "Steel", "17" = "Flying"
)

secondary_type_lookup <- c(
  "0" = "Poison", "1" = "(None)", "2" = "Flying", "3" = "Dark",
  "4" = "Electric", "5" = "Ice", "6" = "Ground", "7" = "Fairy",
  "8" = "Grass", "9" = "Fighting", "10" = "Psychic", "11" = "Steel",
  "12" = "Fire", "13" = "Rock", "14" = "Water", "15" = "Dragon",
  "16" = "Ghost", "17" = "Bug", "18" = "Normal"
)

# Add readable type names BEFORE converting to factors
pokemon <- pokemon %>%
  mutate(
    type_name = primary_type_lookup[as.character(class_primary)],
    type_name_secondary = secondary_type_lookup[as.character(class_secondary)]
  )

# Convert factors
pokemon <- pokemon %>%
  mutate(
    gen = as.factor(gen),
    rarity = as.factor(rarity),
    class_primary = as.factor(class_primary),
    class_secondary = as.factor(class_secondary),
    has_mega_evolution = as.factor(has_mega_evolution),
    has_gigantamax = as.factor(has_gigantamax)
  )

# Define rarity levels (0 = normal, 1 = sublegendary, 2 = legendary, 3 = mythical)
levels(pokemon$rarity) <- c("Normal", "SubLegendary", "Legendary", "Mythical")

# Create derived features
pokemon <- pokemon %>%
  mutate(
    total_stats = hp + attack + defense + sp_attack + sp_defense + speed,
    physical_power = attack + defense,
    special_power = sp_attack + sp_defense,
    offensive_power = attack + sp_attack,
    defensive_power = defense + sp_defense,
    gen_numeric = as.numeric(as.character(gen))
  )

# ═══════════════════════════════════════════════════════════════════
# DEFINE PREDICTOR SETS (used throughout the analysis)
# ═══════════════════════════════════════════════════════════════════

# All predictors for full model (excluding attack which is the response)
all_predictors <- c(
  # Generation (main predictor of interest)
  "gen",
  # Combat stats
  "hp", "defense", "sp_attack", "sp_defense", "speed",
  # Pokémon characteristics
  "capture_rate", "rarity", "height_m", "weight_kg", "base_egg_steps",
  "num_abilities", "evo_length", "has_mega_evolution", "has_gigantamax",
  # Types
  "class_primary", "class_secondary",
  # Demographics
  "percent_male","percent_female",
  # Popularity
  "votes_first", "votes_top_6",
  # Type effectiveness (18 types)
  "against_normal", "against_fire", "against_water", "against_electric",
  "against_grass", "against_ice", "against_fighting", "against_poison",
  "against_ground", "against_flying", "against_psychict", "against_bug",
  "against_rock", "against_ghost", "against_dragon", "against_dark",
  "against_steel", "against_fairy"
)

# Predictors without generation (for null model)
predictors_no_gen <- setdiff(all_predictors, "gen")

# Create formula helper function
make_formula <- function(response, predictors) {
  as.formula(paste(response, "~", paste(predictors, collapse = " + ")))
}

# Display predictor count
cat("Total predictors:", length(all_predictors), "\n")

cat("Dataset size:", nrow(pokemon), "Pokémon\n")
```

---

# 2. Exploratory Analysis of Attack

## 2.1 Attack Distribution

```{r attack-distribution, fig.width=12, fig.height=5}
# Basic statistics
attack_stats <- data.frame(
  Statistic = c("Mean", "Median", "SD", "Min", "Max", "Skewness", "Kurtosis"),
  Value = c(
    round(mean(pokemon$attack), 2),
    round(median(pokemon$attack), 2),
    round(sd(pokemon$attack), 2),
    min(pokemon$attack),
    max(pokemon$attack),
    round(skewness(pokemon$attack), 3),
    round(kurtosis(pokemon$attack), 3)
  )
)

print(attack_stats, row.names = FALSE)

# Histogram with density overlay
p1 <- ggplot(pokemon, aes(x = attack)) +
  geom_histogram(aes(y = after_stat(density)), bins = 35, 
                 fill = "#E76F51", alpha = 0.7, color = "white") +
  geom_density(linewidth = 1.2, color = "#264653") +
  stat_function(fun = dnorm, 
                args = list(mean = mean(pokemon$attack), sd = sd(pokemon$attack)),
                linewidth = 1.2, color = "#2A9D8F", linetype = "dashed") +
  geom_vline(xintercept = mean(pokemon$attack), color = "blue", linewidth = 1) +
  geom_vline(xintercept = median(pokemon$attack), color = "darkgreen", 
             linewidth = 1, linetype = "dotted") +
  labs(title = "Distribution of Attack",
       subtitle = paste0("Mean = ", round(mean(pokemon$attack), 1),
                        ", Median = ", round(median(pokemon$attack), 1),
                        ", Skewness = ", round(skewness(pokemon$attack), 2)),
       x = "Attack", y = "Density") +
  theme_minimal()

# Q-Q plot
p2 <- ggplot(pokemon, aes(sample = attack)) +
  stat_qq(color = "#E76F51", alpha = 0.6, size = 2) +
  stat_qq_line(color = "#264653", linewidth = 1.2) +
  labs(title = "Q-Q Plot: Attack",
       subtitle = "Checking normality assumption",
       x = "Theoretical Quantiles", y = "Sample Quantiles") +
  theme_minimal()

grid.arrange(p1, p2, ncol = 2)
```

```{r shapiro-test}
# Shapiro-Wilk normality test
shapiro_result <- shapiro.test(pokemon$attack)

data.frame(
  Statistic = c("W", "p-value", "Skewness"),
  Value = c(round(shapiro_result$statistic, 4), 
            format.pval(shapiro_result$p.value, digits = 4),
            round(skewness(pokemon$attack), 2))
)
```

## 2.2 Attack by Generation

```{r attack-by-generation, fig.width=12, fig.height=6}
# Summary statistics by generation
gen_summary <- pokemon %>%
  group_by(gen) %>%
  summarise(
    n = n(),
    mean_attack = mean(attack),
    median_attack = median(attack),
    sd_attack = sd(attack),
    min_attack = min(attack),
    max_attack = max(attack),
    se = sd(attack) / sqrt(n())
  ) %>%
  mutate(across(where(is.numeric) & !matches("^n$"), ~round(., 2)))

print(gen_summary, n = Inf)

# Boxplot with means
p_box <- ggplot(pokemon, aes(x = gen, y = attack, fill = gen)) +
  geom_boxplot(alpha = 0.7, outlier.color = "#E76F51", outlier.size = 2) +
  stat_summary(fun = mean, geom = "point", shape = 18, size = 4, color = "black") +
  geom_hline(yintercept = mean(pokemon$attack), linetype = "dashed", 
             color = "red", linewidth = 1) +
  scale_fill_viridis_d(option = "plasma") +
  labs(title = "Attack Distribution by Generation",
       subtitle = "Diamonds = generation means; Red line = overall mean",
       x = "Generation", y = "Attack") +
  theme_minimal() +
  theme(legend.position = "none")

# Mean plot with error bars
p_means <- ggplot(gen_summary, aes(x = gen, y = mean_attack, fill = gen)) +
  geom_bar(stat = "identity", alpha = 0.8) +
  geom_errorbar(aes(ymin = mean_attack - 1.96*se, ymax = mean_attack + 1.96*se), 
                width = 0.3, linewidth = 0.8) +
  geom_hline(yintercept = gen_summary$mean_attack[1], linetype = "dashed", 
             color = "red", linewidth = 1) +
  scale_fill_viridis_d(option = "plasma") +
  labs(title = "Mean Attack by Generation",
       subtitle = "Error bars = 95% CI; Red line = Gen 1 baseline",
       x = "Generation", y = "Mean Attack") +
  theme_minimal() +
  theme(legend.position = "none")

grid.arrange(p_box, p_means, ncol = 2)
```

## 2.3 Attack by Rarity

```{r attack-by-rarity, fig.width=10, fig.height=5}
# Summary by rarity
rarity_summary <- pokemon %>%
  group_by(rarity) %>%
  summarise(
    n = n(),
    mean_attack = round(mean(attack), 2),
    sd_attack = round(sd(attack), 2),
    se = round(sd(attack) / sqrt(n()), 2)
  )

print(rarity_summary)

ggplot(pokemon, aes(x = rarity, y = attack, fill = rarity)) +
  geom_boxplot(alpha = 0.7) +
  stat_summary(fun = mean, geom = "point", shape = 18, size = 4, color = "black") +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Attack Distribution by Rarity",
       subtitle = "Legendary and Mythical Pokémon typically have higher Attack",
       x = "Rarity", y = "Attack") +
  theme_minimal() +
  theme(legend.position = "none")
```

## 2.4 Available Variables Overview

```{r available-variables}
# List all available columns
all_vars <- data.frame(
  Variable = names(pokemon),
  Class = sapply(pokemon, class),
  NonNA = sapply(pokemon, function(x) sum(!is.na(x))),
  Example = sapply(pokemon, function(x) paste0(head(x, 2), collapse = ", "))
)
print(all_vars, row.names = FALSE)
```

## 2.5 Attack by Type (Primary and Secondary)

```{r attack-by-secondary-type, fig.width=14, fig.height=5}
# Secondary type distribution
secondary_summary <- pokemon %>%
  group_by(class_secondary) %>%
  summarise(
    n = n(),
    mean_attack = round(mean(attack), 1)
  ) %>%
  arrange(desc(mean_attack))

print(secondary_summary, n = Inf)
```

## 2.6 Attack by Primary Type

```{r attack-by-type, fig.width=14, fig.height=6}
# Summary by primary type (using readable names)
type_summary_eda <- pokemon %>%
  group_by(type_name) %>%
  summarise(
    n = n(),
    mean_attack = round(mean(attack), 1),
    sd_attack = round(sd(attack), 1),
    median_attack = median(attack)
  ) %>%
  arrange(desc(mean_attack))

print(type_summary_eda, n = Inf)

# Boxplot by type
ggplot(pokemon, aes(x = reorder(type_name, attack, FUN = median), 
                    y = attack, fill = type_name)) +
  geom_boxplot(alpha = 0.7) +
  stat_summary(fun = mean, geom = "point", shape = 18, size = 3, color = "black") +
  geom_hline(yintercept = mean(pokemon$attack), linetype = "dashed", 
             color = "red", linewidth = 1) +
  labs(title = "Attack Distribution by Primary Type",
       subtitle = "Ordered by median Attack; Red line = overall mean",
       x = "Primary Type", y = "Attack") +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1))
```

---

# 3. GLM Prerequisites for Attack Models

To use Generalized Linear Models correctly, we must determine:

1. What **TYPE** of variables are we modeling?
2. What **DISTRIBUTION** does the response follow?
3. How does **VARIANCE** relate to the **MEAN**?
4. What **LINK FUNCTION** is appropriate?

## 3.1 Variable Types

```{r variable-types}
# Response variable
response_info <- data.frame(
  Aspect = c("Variable", "Type", "Nature", "Range", "Bounded"),
  Value = c("Attack", "Numeric (continuous)", "Combat statistic", 
            paste0(min(pokemon$attack), " - ", max(pokemon$attack)),
            "Yes (positive integers)")
)
cat("═══ RESPONSE VARIABLE ═══\n")
print(response_info, row.names = FALSE)

# All predictor variables organized by category
cat("\n═══ PREDICTOR VARIABLES ═══\n")
cat("\n--- Main Predictor (Generation) ---\n")
gen_info <- data.frame(
  Variable = "gen",
  Type = "Categorical (factor)",
  Range_Levels = "9 levels (1-9)",
  Description = "Generation introduced"
)
print(gen_info, row.names = FALSE)

cat("\n--- Combat Stats ---\n")
combat_stats <- data.frame(
  Variable = c("hp", "defense", "sp_attack", "sp_defense", "speed"),
  Type = rep("Continuous", 5),
  Range = c(
    paste0(min(pokemon$hp), "-", max(pokemon$hp)),
    paste0(min(pokemon$defense), "-", max(pokemon$defense)),
    paste0(min(pokemon$sp_attack), "-", max(pokemon$sp_attack)),
    paste0(min(pokemon$sp_defense), "-", max(pokemon$sp_defense)),
    paste0(min(pokemon$speed), "-", max(pokemon$speed))
  )
)
print(combat_stats, row.names = FALSE)

cat("\n--- Pokémon Characteristics ---\n")
char_vars <- data.frame(
  Variable = c("capture_rate", "height_m", "weight_kg", "base_egg_steps", 
               "num_abilities", "evo_length", "percent_male"),
  Type = rep("Continuous", 7),
  Range = c(
    paste0(min(pokemon$capture_rate), "-", max(pokemon$capture_rate)),
    paste0(round(min(pokemon$height_m, na.rm=T), 1), "-", round(max(pokemon$height_m, na.rm=T), 1)),
    paste0(round(min(pokemon$weight_kg, na.rm=T), 1), "-", round(max(pokemon$weight_kg, na.rm=T), 1)),
    paste0(min(pokemon$base_egg_steps, na.rm=T), "-", max(pokemon$base_egg_steps, na.rm=T)),
    paste0(min(pokemon$num_abilities, na.rm=T), "-", max(pokemon$num_abilities, na.rm=T)),
    paste0(min(pokemon$evo_length, na.rm=T), "-", max(pokemon$evo_length, na.rm=T)),
    paste0(min(pokemon$percent_male, na.rm=T), "-", max(pokemon$percent_male, na.rm=T))
  )
)
print(char_vars, row.names = FALSE)

cat("\n--- Categorical Variables ---\n")
cat_vars <- data.frame(
  Variable = c("rarity", "class_primary", "class_secondary", 
               "has_mega_evolution", "has_gigantamax"),
  Type = rep("Categorical (factor)", 5),
  Levels = c(
    paste0(length(unique(pokemon$rarity)), " levels"),
    paste0(length(unique(pokemon$class_primary)), " types"),
    paste0(length(unique(pokemon$class_secondary)), " types"),
    "2 levels (0/1)",
    "2 levels (0/1)"
  )
)
print(cat_vars, row.names = FALSE)

cat("\n--- Popularity (Reddit Survey) ---\n")
pop_vars <- data.frame(
  Variable = c("votes_first", "votes_top_6"),
  Type = rep("Continuous", 2),
  Range = c(
    paste0(min(pokemon$votes_first, na.rm=T), "-", max(pokemon$votes_first, na.rm=T)),
    paste0(min(pokemon$votes_top_6, na.rm=T), "-", max(pokemon$votes_top_6, na.rm=T))
  )
)
print(pop_vars, row.names = FALSE)

cat("\n--- Type Effectiveness (18 variables) ---\n")
against_example <- data.frame(
  Variables = "against_normal, against_fire, ... against_fairy",
  Type = "Continuous",
  Range = "0.25 - 4.0 (damage multiplier)",
  Count = "18 variables"
)
print(against_example, row.names = FALSE)

cat("\n═══ TOTAL PREDICTORS:", length(all_predictors), "═══\n")
```

**Variable Summary:**

| Category | Count | Type |
|----------|-------|------|
| Generation | 1 | Categorical (9 levels) |
| Combat Stats | 5 | Continuous |
| Characteristics | 7 | Continuous |
| Categorical | 5 | Factor |
| Popularity | 2 | Continuous |
| Type Effectiveness | 18 | Continuous |
| **Total** | **38** | Mixed |

**Conclusion:** Attack is a **continuous, positive-valued** response variable. We use 38 predictors including generation (categorical factor), combat stats, physical characteristics, types, popularity metrics, and type effectiveness multipliers.

## 3.2 Distribution of Response Variable

The distribution analysis from Section 2.1 shows:

- **Skewness:** `r round(skewness(pokemon$attack), 2)` (slight positive skew, within acceptable range |skew| < 1)
- **Shapiro-Wilk:** Rejects normality, but with n = `r nrow(pokemon)`, test is overly sensitive
- **Visual assessment:** Approximately normal with slight right tail

```{r distribution-conclusion}
dist_assessment <- data.frame(
  Criterion = c("Skewness", "Kurtosis", "Visual (Q-Q)", "Recommendation"),
  Finding = c(
    paste0(round(skewness(pokemon$attack), 2), " (acceptable: |skew| < 1)"),
    paste0(round(kurtosis(pokemon$attack), 2), " (normal = 3)"),
    "Minor deviation in tails only",
    "Gaussian GLM appropriate; Gamma GLM as robustness check"
  )
)
print(dist_assessment, row.names = FALSE)
```

**Conclusion:** Attack is **approximately normally distributed**. A **Gaussian (Normal) distribution** is appropriate for the primary model. We use **Gamma distribution** as a robustness check since Attack is strictly positive.

## 3.3 Variance-Mean Relationship

For GLM family selection, we examine how variance relates to mean:

- **Gaussian (Normal):** Variance is constant (doesn't depend on mean)
- **Poisson:** Variance equals mean
- **Gamma:** Variance proportional to mean²

```{r variance-analysis, fig.width=14, fig.height=10}
# Fit simple model to check residual variance
model_simple <- lm(attack ~ gen, data = pokemon)

# Variance by generation
var_by_gen <- pokemon %>%
  mutate(residuals = residuals(model_simple)) %>%
  group_by(gen) %>%
  summarise(
    n = n(),
    mean_attack = mean(attack),
    var_residuals = var(residuals),
    sd_residuals = sd(residuals)
  )

# Print table
var_by_gen_display <- var_by_gen
var_by_gen_display[, -1] <- round(var_by_gen_display[, -1], 2)
print(var_by_gen_display)

# ═══════════════════════════════════════════════════════════════════
# VISUALIZATION 1: Variance vs Mean by Generation
# ═══════════════════════════════════════════════════════════════════
p1 <- ggplot(var_by_gen, aes(x = mean_attack, y = var_residuals)) +
  geom_point(aes(size = n, color = gen), alpha = 0.8) +
  geom_smooth(method = "lm", se = TRUE, color = "#E76F51", linetype = "dashed") +
  geom_text(aes(label = paste0("Gen ", gen)), vjust = -1.2, size = 3.5) +
  scale_size_continuous(range = c(4, 12), name = "Sample Size") +
  scale_color_viridis_d(option = "plasma", guide = "none") +
  labs(title = "Variance vs Mean Relationship",
       subtitle = "Each point = one generation; Flat line suggests constant variance (homoscedasticity)",
       x = "Mean Attack by Generation",
       y = "Residual Variance") +
  theme_minimal() +
  theme(legend.position = "right")

# ═══════════════════════════════════════════════════════════════════
# VISUALIZATION 2: Residual Boxplots by Generation
# ═══════════════════════════════════════════════════════════════════
pokemon_with_resid <- pokemon %>%
  mutate(residuals = residuals(model_simple))

p2 <- ggplot(pokemon_with_resid, aes(x = gen, y = residuals, fill = gen)) +
  geom_boxplot(alpha = 0.7, outlier.color = "#E76F51", outlier.size = 2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red", linewidth = 1) +
  scale_fill_viridis_d(option = "plasma", guide = "none") +
  labs(title = "Residual Distribution by Generation",
       subtitle = "Similar box widths suggest homoscedasticity; Red line = zero",
       x = "Generation", y = "Residuals") +
  theme_minimal()

# ═══════════════════════════════════════════════════════════════════
# VISUALIZATION 3: SD Comparison Bar Chart
# ═══════════════════════════════════════════════════════════════════
overall_sd <- sd(pokemon$attack)
p3 <- ggplot(var_by_gen, aes(x = gen, y = sd_residuals, fill = gen)) +
  geom_bar(stat = "identity", alpha = 0.8, color = "white") +
  geom_hline(yintercept = mean(var_by_gen$sd_residuals), linetype = "dashed", 
             color = "#E76F51", linewidth = 1.2) +
  geom_text(aes(label = round(sd_residuals, 1)), vjust = -0.5, size = 3.5) +
  scale_fill_viridis_d(option = "plasma", guide = "none") +
  labs(title = "Residual Standard Deviation by Generation",
       subtitle = paste0("Red line = average SD (", round(mean(var_by_gen$sd_residuals), 1), 
                        "); Ratio max/min = ", 
                        round(max(var_by_gen$sd_residuals)/min(var_by_gen$sd_residuals), 2)),
       x = "Generation", y = "Residual SD") +
  theme_minimal()

# ═══════════════════════════════════════════════════════════════════
# VISUALIZATION 4: Variance Ratio Assessment
# ═══════════════════════════════════════════════════════════════════
var_ratio <- max(var_by_gen$var_residuals) / min(var_by_gen$var_residuals)
sd_ratio <- max(var_by_gen$sd_residuals) / min(var_by_gen$sd_residuals)

assessment_df <- data.frame(
  Metric = c("Max Variance", "Min Variance", "Variance Ratio", 
             "Max SD", "Min SD", "SD Ratio", "Assessment"),
  Value = c(round(max(var_by_gen$var_residuals), 1),
            round(min(var_by_gen$var_residuals), 1),
            round(var_ratio, 2),
            round(max(var_by_gen$sd_residuals), 1),
            round(min(var_by_gen$sd_residuals), 1),
            round(sd_ratio, 2),
            ifelse(sd_ratio < 2, "✓ Acceptable (ratio < 2)", 
                   ifelse(sd_ratio < 3, "⚠ Borderline (ratio 2-3)", 
                          "✗ Heteroscedastic (ratio > 3)")))
)

p4 <- ggplot(assessment_df, aes(x = Metric, y = 1)) +
  geom_tile(aes(fill = Metric == "Assessment"), color = "white", width = 0.9, height = 0.5) +
  geom_text(aes(label = Value), size = 4, fontface = "bold") +
  scale_fill_manual(values = c("TRUE" = "#2A9D8F", "FALSE" = "#E9C46A"), guide = "none") +
  coord_flip() +
  labs(title = "Variance Homogeneity Assessment",
       subtitle = "Rule of thumb: SD ratio < 2 indicates acceptable homoscedasticity") +
  theme_minimal() +
  theme(axis.text.x = element_blank(),
        axis.ticks = element_blank(),
        axis.title = element_blank(),
        panel.grid = element_blank())

# Display plots individually
print(p1)
print(p2)
print(p3)
print(p4)
```

**Interpretation:**

- **Gaussian GLM is appropriate if:** Variance is roughly constant across different mean values (flat relationship in Plot 1, similar box widths in Plot 2)
- **SD Ratio < 2:** Generally indicates acceptable homoscedasticity for regression
- If heteroscedasticity is detected, we can use robust standard errors or Gamma GLM as alternative

## 3.4 Breusch-Pagan Heteroscedasticity Test

```{r bp-test}
# Simple model
bp_simple <- bptest(model_simple)

# Full model with ALL available predictors
model_full <- lm(make_formula("attack", all_predictors), data = pokemon)
bp_full <- bptest(model_full)

bp_results <- data.frame(
  Model = c("Simple (attack ~ gen)", "Full (with controls)"),
  BP_Statistic = c(bp_simple$statistic, bp_full$statistic),
  p_value = c(bp_simple$p.value, bp_full$p.value),
  Result = c(
    ifelse(bp_simple$p.value > 0.05, "Homoscedastic", "Heteroscedastic"),
    ifelse(bp_full$p.value > 0.05, "Homoscedastic", "Heteroscedastic")
  )
)
bp_results[, 2:3] <- round(bp_results[, 2:3], 4)
print(bp_results, row.names = FALSE)
```

### Full Model Variance Diagnostics

```{r full-model-variance, fig.width=14, fig.height=6}
# Get residuals and fitted values from full model
full_model_diag <- data.frame(
  fitted = fitted(model_full),
  residuals = residuals(model_full),
  std_resid = rstandard(model_full),
  sqrt_std_resid = sqrt(abs(rstandard(model_full)))
)

# ═══════════════════════════════════════════════════════════════════
# PLOT 1: Residuals vs Fitted (Full Model)
# ═══════════════════════════════════════════════════════════════════
p_resid_full <- ggplot(full_model_diag, aes(x = fitted, y = residuals)) +
  geom_point(alpha = 0.4, color = "#2A9D8F", size = 2) +
  geom_smooth(method = "loess", se = TRUE, color = "#E76F51", linewidth = 1.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black", linewidth = 1) +
  labs(title = "Residuals vs Fitted Values (Full Model)",
       subtitle = "Random scatter = good fit; Pattern = model misspecification; Funnel shape = heteroscedasticity",
       x = "Fitted Values", y = "Residuals") +
  theme_minimal() +
  theme(plot.title = element_text(size = 14, face = "bold"))

print(p_resid_full)
```

```{r full-model-scale-location, fig.width=14, fig.height=6}
# ═══════════════════════════════════════════════════════════════════
# PLOT 2: Scale-Location Plot (Full Model)
# ═══════════════════════════════════════════════════════════════════
p_scale_full <- ggplot(full_model_diag, aes(x = fitted, y = sqrt_std_resid)) +
  geom_point(alpha = 0.4, color = "#264653", size = 2) +
  geom_smooth(method = "loess", se = TRUE, color = "#E76F51", linewidth = 1.5) +
  geom_hline(yintercept = mean(full_model_diag$sqrt_std_resid), 
             linetype = "dashed", color = "#2A9D8F", linewidth = 1) +
  labs(title = "Scale-Location Plot (Full Model)",
       subtitle = "Flat line = constant variance (homoscedastic); Upward slope = variance increases with fitted values",
       x = "Fitted Values", y = "√|Standardized Residuals|") +
  theme_minimal() +
  theme(plot.title = element_text(size = 14, face = "bold"))

print(p_scale_full)
```

```{r full-model-variance-bins, fig.width=14, fig.height=6}
# ═══════════════════════════════════════════════════════════════════
# PLOT 3: Variance by Fitted Value Bins (Full Model)
# ═══════════════════════════════════════════════════════════════════
# Create bins of fitted values
full_model_diag$fitted_bin <- cut(full_model_diag$fitted, breaks = 10, labels = FALSE)

var_by_bin <- full_model_diag %>%
  group_by(fitted_bin) %>%
  summarise(
    n = n(),
    mean_fitted = mean(fitted),
    var_residuals = var(residuals),
    sd_residuals = sd(residuals)
  ) %>%
  filter(!is.na(fitted_bin))

p_var_bins <- ggplot(var_by_bin, aes(x = mean_fitted, y = sd_residuals)) +
  geom_point(aes(size = n), color = "#E76F51", alpha = 0.8) +
  geom_smooth(method = "lm", se = TRUE, color = "#264653", linetype = "dashed") +
  geom_text(aes(label = paste0("n=", n)), vjust = -1.5, size = 3) +
  scale_size_continuous(range = c(4, 12), name = "Sample Size") +
  labs(title = "Residual SD by Fitted Value Range (Full Model)",
       subtitle = "Each point = one decile of fitted values; Flat trend = homoscedastic",
       x = "Mean Fitted Value (by decile)", 
       y = "Residual Standard Deviation") +
  theme_minimal() +
  theme(plot.title = element_text(size = 14, face = "bold"),
        legend.position = "right")

print(p_var_bins)
```

```{r full-model-variance-comparison, fig.width=14, fig.height=6}
# ═══════════════════════════════════════════════════════════════════
# PLOT 4: Compare Simple vs Full Model Residual Spread
# ═══════════════════════════════════════════════════════════════════
comparison_data <- data.frame(
  Model = rep(c("Simple (attack ~ gen)", "Full (38 predictors)"), 
              c(nrow(pokemon), nrow(pokemon))),
  Residuals = c(residuals(model_simple), residuals(model_full)),
  Fitted = c(fitted(model_simple), fitted(model_full))
)

p_compare <- ggplot(comparison_data, aes(x = Fitted, y = Residuals, color = Model)) +
  geom_point(alpha = 0.3, size = 1.5) +
  geom_smooth(method = "loess", se = FALSE, linewidth = 1.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  scale_color_manual(values = c("Simple (attack ~ gen)" = "#E9C46A", 
                                "Full (38 predictors)" = "#2A9D8F")) +
  facet_wrap(~Model, scales = "free_x") +
  labs(title = "Residual Comparison: Simple vs Full Model",
       subtitle = paste0("Simple model SD = ", round(sd(residuals(model_simple)), 1),
                        " | Full model SD = ", round(sd(residuals(model_full)), 1),
                        " | Reduction = ", round((1 - sd(residuals(model_full))/sd(residuals(model_simple)))*100, 1), "%"),
       x = "Fitted Values", y = "Residuals") +
  theme_minimal() +
  theme(plot.title = element_text(size = 14, face = "bold"),
        legend.position = "none",
        strip.text = element_text(size = 12, face = "bold"))

print(p_compare)
```

```{r log-transformation-check, fig.width=14, fig.height=10}
# ═══════════════════════════════════════════════════════════════════
# LOG TRANSFORMATION CHECK: Does log(Attack) fix heteroscedasticity?
# ═══════════════════════════════════════════════════════════════════

# Fit model with log-transformed response
pokemon$log_attack <- log(pokemon$attack)
model_full_log <- lm(make_formula("log_attack", all_predictors), data = pokemon)

# Get diagnostics for log model
log_model_diag <- data.frame(
  fitted = fitted(model_full_log),
  residuals = residuals(model_full_log),
  std_resid = rstandard(model_full_log),
  sqrt_std_resid = sqrt(abs(rstandard(model_full_log)))
)

# Breusch-Pagan test for log model
bp_log <- bptest(model_full_log)

# ═══════════════════════════════════════════════════════════════════
# PLOT 1: Side-by-side Scale-Location comparison
# ═══════════════════════════════════════════════════════════════════
p_scale_original <- ggplot(full_model_diag, aes(x = fitted, y = sqrt_std_resid)) +
  geom_point(alpha = 0.4, color = "#E76F51", size = 2) +
  geom_smooth(method = "loess", se = TRUE, color = "#264653", linewidth = 1.5) +
  geom_hline(yintercept = mean(full_model_diag$sqrt_std_resid), 
             linetype = "dashed", color = "black") +
  labs(title = "Original: Attack",
       subtitle = paste0("BP test p = ", format.pval(bp_full$p.value, digits = 3)),
       x = "Fitted Values", y = "√|Standardized Residuals|") +
  theme_minimal() +
  theme(plot.title = element_text(size = 12, face = "bold"))

p_scale_log <- ggplot(log_model_diag, aes(x = fitted, y = sqrt_std_resid)) +
  geom_point(alpha = 0.4, color = "#2A9D8F", size = 2) +
  geom_smooth(method = "loess", se = TRUE, color = "#264653", linewidth = 1.5) +
  geom_hline(yintercept = mean(log_model_diag$sqrt_std_resid), 
             linetype = "dashed", color = "black") +
  labs(title = "Log-Transformed: log(Attack)",
       subtitle = paste0("BP test p = ", format.pval(bp_log$p.value, digits = 3)),
       x = "Fitted Values", y = "√|Standardized Residuals|") +
  theme_minimal() +
  theme(plot.title = element_text(size = 12, face = "bold"))

# ═══════════════════════════════════════════════════════════════════
# PLOT 2: Side-by-side Residuals vs Fitted
# ═══════════════════════════════════════════════════════════════════
p_resid_original <- ggplot(full_model_diag, aes(x = fitted, y = std_resid)) +
  geom_point(alpha = 0.4, color = "#E76F51", size = 2) +
  geom_smooth(method = "loess", se = TRUE, color = "#264653", linewidth = 1.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  labs(title = "Original: Attack",
       x = "Fitted Values", y = "Standardized Residuals") +
  theme_minimal() +
  theme(plot.title = element_text(size = 12, face = "bold"))

p_resid_log <- ggplot(log_model_diag, aes(x = fitted, y = std_resid)) +
  geom_point(alpha = 0.4, color = "#2A9D8F", size = 2) +
  geom_smooth(method = "loess", se = TRUE, color = "#264653", linewidth = 1.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  labs(title = "Log-Transformed: log(Attack)",
       x = "Fitted Values", y = "Standardized Residuals") +
  theme_minimal() +
  theme(plot.title = element_text(size = 12, face = "bold"))

# Combine plots
grid.arrange(
  p_scale_original, p_scale_log,
  p_resid_original, p_resid_log,
  ncol = 2,
  top = "Heteroscedasticity Comparison: Original vs Log-Transformed Attack"
)
```

```{r log-transform-qq, fig.width=14, fig.height=5}
# ═══════════════════════════════════════════════════════════════════
# Q-Q Plot Comparison
# ═══════════════════════════════════════════════════════════════════
p_qq_original <- ggplot(full_model_diag, aes(sample = std_resid)) +
  stat_qq(color = "#E76F51", alpha = 0.6, size = 2) +
  stat_qq_line(color = "#264653", linewidth = 1.2) +
  labs(title = "Q-Q Plot: Original Model",
       subtitle = "Residuals from Attack ~ predictors",
       x = "Theoretical Quantiles", y = "Sample Quantiles") +
  theme_minimal() +
  theme(plot.title = element_text(size = 12, face = "bold"))

p_qq_log <- ggplot(log_model_diag, aes(sample = std_resid)) +
  stat_qq(color = "#2A9D8F", alpha = 0.6, size = 2) +
  stat_qq_line(color = "#264653", linewidth = 1.2) +
  labs(title = "Q-Q Plot: Log-Transformed Model",
       subtitle = "Residuals from log(Attack) ~ predictors",
       x = "Theoretical Quantiles", y = "Sample Quantiles") +
  theme_minimal() +
  theme(plot.title = element_text(size = 12, face = "bold"))

grid.arrange(p_qq_original, p_qq_log, ncol = 2)
```

```{r sqrt-transformation-check, fig.width=14, fig.height=6}
# ═══════════════════════════════════════════════════════════════════
# SQRT TRANSFORMATION CHECK
# ═══════════════════════════════════════════════════════════════════

# Fit model with sqrt-transformed response
pokemon$sqrt_attack <- sqrt(pokemon$attack)
model_full_sqrt <- lm(make_formula("sqrt_attack", all_predictors), data = pokemon)

# Get diagnostics for sqrt model
sqrt_model_diag <- data.frame(
  fitted = fitted(model_full_sqrt),
  residuals = residuals(model_full_sqrt),
  std_resid = rstandard(model_full_sqrt),
  sqrt_std_resid = sqrt(abs(rstandard(model_full_sqrt)))
)

# Breusch-Pagan test for sqrt model
bp_sqrt <- bptest(model_full_sqrt)

# Scale-Location plots for all three
p_scale_sqrt <- ggplot(sqrt_model_diag, aes(x = fitted, y = sqrt_std_resid)) +
  geom_point(alpha = 0.4, color = "#9B5DE5", size = 2) +
  geom_smooth(method = "loess", se = TRUE, color = "#264653", linewidth = 1.5) +
  geom_hline(yintercept = mean(sqrt_model_diag$sqrt_std_resid), 
             linetype = "dashed", color = "black") +
  labs(title = "Sqrt-Transformed: √Attack",
       subtitle = paste0("BP test p = ", format.pval(bp_sqrt$p.value, digits = 3)),
       x = "Fitted Values", y = "√|Standardized Residuals|") +
  theme_minimal() +
  theme(plot.title = element_text(size = 12, face = "bold"))

# Compare all three scale-location plots
grid.arrange(
  p_scale_original, p_scale_log, p_scale_sqrt,
  ncol = 3,
  top = "Scale-Location Comparison: Original vs Log vs Sqrt Transformation"
)
```

```{r all-transforms-comparison, fig.width=14, fig.height=5}
# ═══════════════════════════════════════════════════════════════════
# Q-Q Plots for all three transformations
# ═══════════════════════════════════════════════════════════════════
p_qq_sqrt <- ggplot(sqrt_model_diag, aes(sample = std_resid)) +
  stat_qq(color = "#9B5DE5", alpha = 0.6, size = 2) +
  stat_qq_line(color = "#264653", linewidth = 1.2) +
  labs(title = "Q-Q Plot: Sqrt Model",
       subtitle = "Residuals from √Attack ~ predictors",
       x = "Theoretical Quantiles", y = "Sample Quantiles") +
  theme_minimal() +
  theme(plot.title = element_text(size = 12, face = "bold"))

grid.arrange(p_qq_original, p_qq_log, p_qq_sqrt, ncol = 3)
```

```{r transformation-summary}
# ═══════════════════════════════════════════════════════════════════
# COMPREHENSIVE SUMMARY: All Transformations
# ═══════════════════════════════════════════════════════════════════
cat(paste0(
  "\n═══════════════════════════════════════════════════════════════════\n",
  "         TRANSFORMATION COMPARISON SUMMARY\n",
  "═══════════════════════════════════════════════════════════════════\n\n"
))

comparison_table <- data.frame(
  Metric = c("Response Variable", "BP Statistic", "BP p-value", 
             "Heteroscedastic?", "Residual SD", "R²"),
  Original = c(
    "Attack",
    round(bp_full$statistic, 2),
    format.pval(bp_full$p.value, digits = 4),
    ifelse(bp_full$p.value < 0.05, "Yes ⚠", "No ✓"),
    round(sd(residuals(model_full)), 3),
    round(summary(model_full)$r.squared, 4)
  ),
  Log = c(
    "log(Attack)",
    round(bp_log$statistic, 2),
    format.pval(bp_log$p.value, digits = 4),
    ifelse(bp_log$p.value < 0.05, "Yes ⚠", "No ✓"),
    round(sd(residuals(model_full_log)), 4),
    round(summary(model_full_log)$r.squared, 4)
  ),
  Sqrt = c(
    "√Attack",
    round(bp_sqrt$statistic, 2),
    format.pval(bp_sqrt$p.value, digits = 4),
    ifelse(bp_sqrt$p.value < 0.05, "Yes ⚠", "No ✓"),
    round(sd(residuals(model_full_sqrt)), 4),
    round(summary(model_full_sqrt)$r.squared, 4)
  )
)
print(comparison_table, row.names = FALSE)

# Determine best transformation
p_values <- c(Original = bp_full$p.value, Log = bp_log$p.value, Sqrt = bp_sqrt$p.value)
best_transform <- names(which.max(p_values))

if(max(p_values) > 0.05) {
  rec_text <- paste0(
    "✓ The ", best_transform, " transformation achieves HOMOSCEDASTICITY (p > 0.05)\n",
    "  → Recommended for valid standard errors\n"
  )
} else {
  rec_text <- paste0(
    "⚠ None of the transformations fully resolve heteroscedasticity\n",
    "  → Consider using:\n",
    "    1. Robust/sandwich standard errors\n",
    "    2. Gamma GLM with log link\n",
    "    3. Weighted least squares\n"
  )
}

cat(paste0(
  "\n═══════════════════════════════════════════════════════════════════\n",
  "                    RECOMMENDATION\n",
  "═══════════════════════════════════════════════════════════════════\n\n",
  "BP p-values (higher = more homoscedastic):\n",
  "  Original:  ", format.pval(bp_full$p.value, digits = 4), "\n",
  "  Log:       ", format.pval(bp_log$p.value, digits = 4), "\n",
  "  Sqrt:      ", format.pval(bp_sqrt$p.value, digits = 4), "\n\n",
  "Best transformation: ", best_transform, "\n",
  "  (Highest BP p-value = least heteroscedasticity)\n\n",
  rec_text, "\n",
  "--- Coefficient Interpretation ---\n",
  "Original: β = change in Attack points\n",
  "Log:      β ≈ % change in Attack (multiply by 100)\n",
  "Sqrt:     β = change in √Attack units (harder to interpret)\n"
))
```

```{r heteroscedasticity-assessment}
# Summary assessment
sd_ratio_full <- max(var_by_bin$sd_residuals) / min(var_by_bin$sd_residuals)

if(bp_full$p.value < 0.05) {
  rec_hetero <- paste0(
    "Recommendation:\n",
    "  The full model shows heteroscedasticity. Consider:\n",
    "  1. Using robust standard errors (sandwich estimator)\n",
    "  2. Gamma GLM as alternative (already included as robustness check)\n",
    "  3. Results remain valid for coefficient estimates, but SEs may be affected\n"
  )
} else {
  rec_hetero <- paste0(
    "Recommendation:\n",
    "  The full model shows acceptable homoscedasticity.\n",
    "  Standard GLM inference is valid.\n"
  )
}

cat(paste0(
  "\n═══════════════════════════════════════════════════════════════════\n",
  "                 HETEROSCEDASTICITY ASSESSMENT\n",
  "═══════════════════════════════════════════════════════════════════\n\n",
  "Breusch-Pagan Test Results:\n",
  "  Simple model: p = ", round(bp_simple$p.value, 4), " ",
    ifelse(bp_simple$p.value > 0.05, "→ Homoscedastic ✓", "→ Heteroscedastic ⚠"), "\n",
  "  Full model:   p = ", round(bp_full$p.value, 4), " ",
    ifelse(bp_full$p.value > 0.05, "→ Homoscedastic ✓", "→ Heteroscedastic ⚠"), "\n\n",
  "Full Model Variance Check:\n",
  "  SD Range: ", round(min(var_by_bin$sd_residuals), 1), " - ", 
    round(max(var_by_bin$sd_residuals), 1), "\n",
  "  SD Ratio (max/min): ", round(sd_ratio_full, 2), "\n",
  "  Assessment: ", ifelse(sd_ratio_full < 2, "Acceptable (ratio < 2) ✓", 
                            ifelse(sd_ratio_full < 3, "Borderline (ratio 2-3) ⚠",
                                   "Problematic (ratio > 3) ✗")), "\n\n",
  rec_hetero
))
```

## 3.5 Multicollinearity Check (VIF)

```{r vif-check}
# VIF for full model
vif_values <- car::vif(model_full)

if(is.matrix(vif_values)) {
  vif_df <- data.frame(
    Variable = rownames(vif_values),
    GVIF = round(vif_values[, 1], 2),
    Df = vif_values[, 2],
    `GVIF_adj` = round(vif_values[, 3], 2)
  )
  print(vif_df, row.names = FALSE)
  max_vif <- max(vif_values[, 3])
} else {
  vif_df <- data.frame(Variable = names(vif_values), VIF = round(vif_values, 2))
  print(vif_df, row.names = FALSE)
  max_vif <- max(vif_values)
}

```

## 3.6 Link Function Selection

The link function connects the linear predictor to the expected value of the response.

```{r link-function-analysis, fig.width=12, fig.height=5}
# Compare identity vs log link predictions
model_identity <- glm(attack ~ gen + hp + defense, data = pokemon, 
                      family = gaussian(link = "identity"))
model_log <- glm(attack ~ gen + hp + defense, data = pokemon, 
                 family = gaussian(link = "log"))

# Prediction comparison
link_comparison <- data.frame(
  Actual = pokemon$attack,
  Identity_Pred = fitted(model_identity),
  Log_Pred = fitted(model_log)
)

rmse_identity <- sqrt(mean((link_comparison$Actual - link_comparison$Identity_Pred)^2))
rmse_log <- sqrt(mean((link_comparison$Actual - link_comparison$Log_Pred)^2))

link_choice <- data.frame(
  Link = c("Identity", "Log"),
  Formula = c("E(Y) = Xβ", "log(E(Y)) = Xβ"),
  RMSE = c(round(rmse_identity, 2), round(rmse_log, 2)),
  Interpretation = c("β = change in Attack points", "exp(β) = multiplicative change")
)
print(link_choice, row.names = FALSE)

# Visual comparison
p_id <- ggplot(link_comparison, aes(x = Actual, y = Identity_Pred)) +
  geom_point(alpha = 0.4, color = "#2A9D8F") +
  geom_abline(slope = 1, intercept = 0, color = "red", linewidth = 1.2, linetype = "dashed") +
  labs(title = "Identity Link", subtitle = paste0("RMSE = ", round(rmse_identity, 1)),
       x = "Actual", y = "Predicted") +
  theme_minimal()

p_log <- ggplot(link_comparison, aes(x = Actual, y = Log_Pred)) +
  geom_point(alpha = 0.4, color = "#E76F51") +
  geom_abline(slope = 1, intercept = 0, color = "red", linewidth = 1.2, linetype = "dashed") +
  labs(title = "Log Link", subtitle = paste0("RMSE = ", round(rmse_log, 1)),
       x = "Actual", y = "Predicted") +
  theme_minimal()

grid.arrange(p_id, p_log, ncol = 2, top = "Link Function Comparison")
```

**Conclusion:** We choose the **Identity link** because:

1. **Interpretability:** Coefficients represent direct change in Attack points
2. **No boundary issues:** Attack range (5-185) is well above zero
3. **Theoretical basis:** Power creep is expected to be additive rather than multiplicative
4. **Similar fit:** Both links perform comparably

## 3.7 Diagnostic Plots

```{r diagnostic-plots, fig.width=12, fig.height=10}
par(mfrow = c(2, 2), mar = c(4, 4, 3, 2))
plot(model_full, which = 1:4)
```

```{r custom-diagnostics, fig.width=12, fig.height=5}
# Custom diagnostic plots
diag_data <- data.frame(
  fitted = fitted(model_full),
  residuals = residuals(model_full),
  std_resid = rstandard(model_full),
  sqrt_std_resid = sqrt(abs(rstandard(model_full)))
)

p_resid_fitted <- ggplot(diag_data, aes(x = fitted, y = residuals)) +
  geom_point(alpha = 0.4, color = "#2A9D8F") +
  geom_smooth(method = "loess", se = TRUE, color = "#E76F51", linewidth = 1.2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  labs(title = "Residuals vs Fitted Values",
       subtitle = "Random scatter around 0 indicates good model specification",
       x = "Fitted Values", y = "Residuals") +
  theme_minimal()

p_scale_loc <- ggplot(diag_data, aes(x = fitted, y = sqrt_std_resid)) +
  geom_point(alpha = 0.4, color = "#264653") +
  geom_smooth(method = "loess", se = TRUE, color = "#E76F51", linewidth = 1.2) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "gray50") +
  labs(title = "Scale-Location Plot",
       subtitle = "Flat line indicates constant variance",
       x = "Fitted Values", y = "√|Standardized Residuals|") +
  theme_minimal()

grid.arrange(p_resid_fitted, p_scale_loc, ncol = 2)
```

## 3.8 Prerequisites Summary

```{r prerequisites-summary}
prereq_summary <- data.frame(
  Question = c(
    "1. Variable Types",
    "2. Distribution",
    "3. Variance-Mean",
    "4. Link Function"
  ),
  Finding = c(
    "Attack = continuous positive; Gen = categorical factor",
    "Approximately normal (skew = 0.28, acceptable)",
    "Homoscedastic in simple model; mild heteroscedasticity in full model",
    "Identity link (additive interpretation, no boundary issues)"
  ),
  GLM_Choice = c(
    "Gaussian family appropriate",
    "Gaussian primary; Gamma for robustness",
    "Standard errors valid; robust checks included",
    "Identity link for Gaussian; Log link for Gamma"
  )
)
print(prereq_summary, row.names = FALSE)
```

**Final GLM Specification:**

- **Primary Model:** `glm(attack ~ gen + controls, family = gaussian(link = "identity"))`
- **Robustness Check:** `glm(attack ~ gen + controls, family = Gamma(link = "log"))`

---

# 4. Attack GLM Models

## 4.1 Model 1: Simple Generation Effect

Before fitting the model, let's visualize what Attack looks like across generations to understand what the model will capture:

```{r simple-model-explanation, fig.width=14, fig.height=10}
# ═══════════════════════════════════════════════════════════════════
# VISUALIZING ATTACK BY GENERATION (What the Simple Model Captures)
# ═══════════════════════════════════════════════════════════════════

# Calculate generation means and Gen 1 reference
gen1_mean <- mean(pokemon$attack[pokemon$gen == "1"])

gen_stats <- pokemon %>%
  group_by(gen) %>%
  summarise(
    n = n(),
    mean_attack = mean(attack),
    median_attack = median(attack),
    sd_attack = sd(attack),
    se = sd(attack) / sqrt(n()),
    diff_from_gen1 = mean(attack) - gen1_mean
  ) %>%
  mutate(
    significant = abs(diff_from_gen1) / se > 1.96,  # Approximate significance
    label = paste0(ifelse(diff_from_gen1 >= 0, "+", ""), round(diff_from_gen1, 1))
  )

# PLOT 1: Boxplot with individual points and Gen 1 reference
p1 <- ggplot(pokemon, aes(x = gen, y = attack)) +
  geom_boxplot(aes(fill = gen), alpha = 0.6, outlier.shape = NA) +
  geom_jitter(aes(color = gen), width = 0.2, alpha = 0.3, size = 1) +
  geom_hline(yintercept = gen1_mean, linetype = "dashed", 
             color = "red", linewidth = 1.2) +
  stat_summary(fun = mean, geom = "point", shape = 18, size = 5, color = "black") +
  annotate("text", x = 9.5, y = gen1_mean + 5, label = paste0("Gen 1 Mean\n(", round(gen1_mean, 1), ")"),
           color = "red", fontface = "bold", size = 3.5) +
  scale_fill_viridis_d(option = "plasma", guide = "none") +
  scale_color_viridis_d(option = "plasma", guide = "none") +
  labs(title = "Attack Distribution by Generation",
       subtitle = "Each point = one Pokémon; Diamonds = generation means; Red line = Gen 1 mean (reference)",
       x = "Generation", y = "Attack") +
  theme_minimal() +
  theme(plot.title = element_text(size = 14, face = "bold"))

# PLOT 2: Bar chart showing difference from Gen 1 (= GLM coefficients)
p2 <- ggplot(gen_stats, aes(x = gen, y = diff_from_gen1, fill = diff_from_gen1 > 0)) +
  geom_bar(stat = "identity", alpha = 0.8, color = "black") +
  geom_errorbar(aes(ymin = diff_from_gen1 - 1.96*se, ymax = diff_from_gen1 + 1.96*se),
                width = 0.3, linewidth = 0.8) +
  geom_hline(yintercept = 0, linewidth = 1.2, color = "black") +
  geom_text(aes(label = label, y = diff_from_gen1 + sign(diff_from_gen1) * 3), 
            size = 4, fontface = "bold") +
  scale_fill_manual(values = c("TRUE" = "#2A9D8F", "FALSE" = "#E76F51"),
                    labels = c("TRUE" = "Higher than Gen 1", "FALSE" = "Lower than Gen 1"),
                    name = "") +
  labs(title = "How Each Generation Compares to Gen 1",
       subtitle = "These bars ARE the GLM coefficients (difference from Gen 1 mean); Error bars = 95% CI",
       x = "Generation", y = "Difference from Gen 1 (Attack Points)") +
  theme_minimal() +
  theme(plot.title = element_text(size = 14, face = "bold"),
        legend.position = "bottom")

# PLOT 3: Mean trajectory with trend line
p3 <- ggplot(gen_stats, aes(x = as.numeric(as.character(gen)), y = mean_attack)) +
  geom_line(linewidth = 1.2, color = "#264653") +
  geom_point(aes(size = n), color = "#E76F51", alpha = 0.8) +
  geom_smooth(method = "lm", se = TRUE, color = "#2A9D8F", 
              linetype = "dashed", linewidth = 1) +
  geom_hline(yintercept = gen1_mean, linetype = "dotted", 
             color = "red", linewidth = 1) +
  scale_x_continuous(breaks = 1:9) +
  scale_size_continuous(range = c(4, 10), name = "n Pokémon") +
  labs(title = "Attack Mean Trajectory Across Generations",
       subtitle = "Dashed line = linear trend; Dotted red = Gen 1 baseline; Point size = sample size",
       x = "Generation", y = "Mean Attack") +
  theme_minimal() +
  theme(plot.title = element_text(size = 14, face = "bold"))

# PLOT 4: Summary table as visualization
gen_table <- gen_stats %>%
  dplyr::select(gen, n, mean_attack, diff_from_gen1) %>%
  mutate(
    mean_attack = round(mean_attack, 1),
    diff_from_gen1 = round(diff_from_gen1, 1)
  )

p4 <- ggplot(gen_table, aes(x = gen, y = 1)) +
  geom_tile(aes(fill = diff_from_gen1), color = "white", width = 0.9, height = 0.8) +
  geom_text(aes(label = paste0("n=", n, "\nMean=", mean_attack, "\nΔ=", 
                                ifelse(diff_from_gen1 >= 0, "+", ""), diff_from_gen1)),
            size = 3.5, fontface = "bold") +
  scale_fill_gradient2(low = "#E76F51", mid = "white", high = "#2A9D8F", 
                       midpoint = 0, name = "Diff from Gen 1") +
  labs(title = "Generation Summary Statistics",
       subtitle = "n = sample size; Mean = average Attack; Δ = difference from Gen 1") +
  theme_minimal() +
  theme(axis.text.y = element_blank(),
        axis.ticks = element_blank(),
        axis.title = element_blank(),
        panel.grid = element_blank(),
        plot.title = element_text(size = 14, face = "bold"))

# Display plots
grid.arrange(p1, p2, ncol = 2)
```

```{r simple-model-trajectory, fig.width=12, fig.height=5}
grid.arrange(p3, p4, ncol = 2)
```

**Reading the Plots:**

- **Top Left (Boxplot):** Shows the full distribution of Attack for each generation. The red dashed line is the Gen 1 mean (our reference). Generations above this line have positive coefficients.
- **Top Right (Bar Chart):** These bars directly represent what the GLM estimates — the difference between each generation's mean Attack and Gen 1. Bars above zero = higher Attack than Gen 1.
- **Bottom Left (Trajectory):** Connects the means to show the overall pattern. The upward dashed trend line suggests power creep.
- **Bottom Right (Summary):** Quick reference for sample sizes and key statistics per generation.

Now let's fit the simple model:

```{r model1-simple}
# Model 1: Attack as function of generation only
model1 <- glm(attack ~ gen, data = pokemon, family = gaussian(link = "identity"))

summary(model1)

# Extract and format coefficients
model1_coefs <- tidy(model1) %>%
  mutate(
    estimate = round(estimate, 3),
    std.error = round(std.error, 3),
    statistic = round(statistic, 3),
    p.value = format.pval(p.value, digits = 3)
  )

# Generation coefficients
model1_coefs %>% filter(str_detect(term, "gen"))
```

**Interpretation:** The intercept (`r round(coef(model1)[1], 1)`) is the mean Attack for Generation 1. Each generation coefficient represents the difference from this baseline — exactly what we visualized in the bar chart above.

## 4.2 Model 2: With Combat Stat Controls

```{r model2-controls}
# Model 2: Add other combat stats as controls
model2 <- glm(attack ~ gen + hp + defense + sp_attack + sp_defense + speed, 
              data = pokemon, family = gaussian(link = "identity"))

summary(model2)

model2_coefs <- tidy(model2) %>%
  mutate(
    estimate = round(estimate, 3),
    std.error = round(std.error, 3),
    statistic = round(statistic, 3),
    p.value = format.pval(p.value, digits = 3),
    significant = ifelse(as.numeric(p.value) < 0.05, "*", "")
  )

model2_coefs
```

## 4.3 Model 3: Full Model with All Available Predictors

```{r model3-full}
# Model 3: Full model with ALL available predictors
model3 <- glm(make_formula("attack", all_predictors), 
              data = pokemon, family = gaussian(link = "identity"))

summary(model3)

# Extract significant predictors
model3_coefs <- tidy(model3) %>%
  mutate(significant = p.value < 0.1) %>%
  arrange(p.value)

# Show all significant predictors
model3_coefs %>%
  filter(significant) %>%
  mutate(
    estimate = round(estimate, 3),
    std.error = round(std.error, 3),
    p.value = format.pval(p.value, digits = 3)
  ) %>%
  dplyr::select(term, estimate, std.error, p.value)
```

## 4.4 Model 4: Gamma GLM (Robustness Check)

```{r model4-gamma}
# Model 4: Gamma GLM with log link (same predictors as Model 3)
model4 <- glm(make_formula("attack", all_predictors), 
              data = pokemon, family = Gamma(link = "log"))

summary(model4)

# Convert coefficients to percentage change
gamma_coefs <- tidy(model4) %>%
  filter(str_detect(term, "gen")) %>%
  mutate(
    generation = as.numeric(str_extract(term, "\\d+")),
    pct_change = round((exp(estimate) - 1) * 100, 2),
    estimate = round(estimate, 4),
    p.value = format.pval(p.value, digits = 3)
  ) %>%
  dplyr::select(generation, estimate, pct_change, p.value)

gamma_coefs
```

## 4.5 Model Comparison

```{r model-comparison}
# Compare all models
model_comparison <- data.frame(
  Model = c("M1: gen only", "M2: + combat stats", "M3: + rarity", "M4: Gamma GLM"),
  Predictors = c("gen", "gen + 5 stats", "gen + 5 stats + rarity + capture_rate", 
                 "Same as M3 (Gamma)"),
  AIC = c(AIC(model1), AIC(model2), AIC(model3), AIC(model4)),
  BIC = c(BIC(model1), BIC(model2), BIC(model3), BIC(model4)),
  Deviance = c(model1$deviance, model2$deviance, model3$deviance, model4$deviance),
  Pseudo_R2 = c(
    1 - model1$deviance/model1$null.deviance,
    1 - model2$deviance/model2$null.deviance,
    1 - model3$deviance/model3$null.deviance,
    1 - model4$deviance/model4$null.deviance
  )
)

model_comparison[, 3:6] <- round(model_comparison[, 3:6], 2)

print(model_comparison, row.names = FALSE)
```

```{r model-comparison-plot, fig.width=10, fig.height=5}
# Visual comparison
model_comparison_long <- model_comparison %>%
  pivot_longer(cols = c(AIC, BIC), names_to = "Criterion", values_to = "Value")

ggplot(model_comparison_long, aes(x = Model, y = Value, fill = Criterion)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8), alpha = 0.8) +
  scale_fill_manual(values = c("AIC" = "#E76F51", "BIC" = "#2A9D8F")) +
  labs(title = "Model Comparison: AIC and BIC",
       subtitle = "Lower values indicate better model fit",
       x = "Model", y = "Information Criterion Value") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 15, hjust = 1))
```

## 4.6 LASSO Regularization: Feature Selection

### What is LASSO and Why Use It?

**LASSO** (Least Absolute Shrinkage and Selection Operator) is a regularization technique that adds a penalty to the regression model. This penalty forces the model to "pay a cost" for each predictor it uses. The key insight is that LASSO shrinks unimportant coefficients all the way to **exactly zero**, effectively removing them from the model.

**Why is this useful for power creep analysis?**

In standard regression (OLS/GLM), we include all 38 predictors and each gets a coefficient. But which predictors actually *matter*? LASSO answers this objectively by forcing predictors to compete:

- **Strong predictors** keep their coefficients (though slightly shrunk)
- **Weak predictors** get shrunk to zero and are eliminated
- **The λ (lambda) parameter** controls how harsh the penalty is — higher λ = fewer survivors

**The critical question:** If we force all 38 predictors to compete, does **Generation** survive? If generation coefficients shrink to zero, it means other variables (combat stats, type, rarity) explain Attack better than generation does. If generation survives, it provides unique predictive value that other variables cannot explain.

**Two common λ choices:**
- **λ.min:** Minimizes cross-validation error (keeps more predictors)
- **λ.1se:** One standard error rule (more parsimonious, stricter selection)

```{r lasso-preparation}
# Prepare data for LASSO (requires matrix format)
# Create model matrix with all predictors (excluding intercept)
X <- model.matrix(make_formula("attack", all_predictors), data = pokemon)[, -1]
y <- pokemon$attack

cat("Feature matrix dimensions:", dim(X)[1], "observations x", dim(X)[2], "features\n")
```

```{r lasso-cv, fig.width=12, fig.height=5}
# Cross-validated LASSO to find optimal lambda
set.seed(42)
cv_lasso <- cv.glmnet(X, y, alpha = 1, nfolds = 10)

# Plot cross-validation curve
plot(cv_lasso, main = "LASSO Cross-Validation: Finding Optimal Lambda")

# Extract optimal lambdas
lambda_min <- cv_lasso$lambda.min
lambda_1se <- cv_lasso$lambda.1se

cat(paste0(
  "\n═══════════════════════════════════════════════════════════════════\n",
  "                 LASSO CROSS-VALIDATION RESULTS\n",
  "═══════════════════════════════════════════════════════════════════\n\n",
  "Lambda (min MSE):    ", round(lambda_min, 4), "\n",
  "Lambda (1-SE rule):  ", round(lambda_1se, 4), "\n",
  "  → 1-SE rule gives a more parsimonious model\n"
))
```

```{r lasso-coefficients, fig.width=12, fig.height=8}
# Fit final LASSO models
lasso_min <- glmnet(X, y, alpha = 1, lambda = lambda_min)
lasso_1se <- glmnet(X, y, alpha = 1, lambda = lambda_1se)

# Extract coefficients at lambda.min
coefs_min <- as.matrix(coef(lasso_min))
coefs_1se <- as.matrix(coef(lasso_1se))

# ═══════════════════════════════════════════════════════════════════
# CREATE READABLE VARIABLE NAMES (map factor levels back to names)
# ═══════════════════════════════════════════════════════════════════

# Type name lookup tables (derived from pokemon.csv mapping)
primary_type_names <- c(
  "0" = "Grass", "1" = "Fire", "2" = "Water", "3" = "Bug", 
  "4" = "Normal", "5" = "Poison", "6" = "Electric", "7" = "Ground",
  "8" = "Fairy", "9" = "Fighting", "10" = "Psychic", "11" = "Rock",
  "12" = "Ghost", "13" = "Ice", "14" = "Dragon", "15" = "Dark",
  "16" = "Steel", "17" = "Flying"
)

secondary_type_names <- c(
  "0" = "Poison", "1" = "(None)", "2" = "Flying", "3" = "Dark",
  "4" = "Electric", "5" = "Ice", "6" = "Ground", "7" = "Fairy",
  "8" = "Grass", "9" = "Fighting", "10" = "Psychic", "11" = "Steel",
  "12" = "Fire", "13" = "Rock", "14" = "Water", "15" = "Dragon",
  "16" = "Ghost", "17" = "Bug", "18" = "Normal"
)

rarity_names <- c("0" = "Normal", "1" = "SubLegendary", "2" = "Legendary", "3" = "Mythical")

# Function to convert coded variable names to readable names
make_readable <- function(var_name) {
  # Generation: gen2 -> "Gen 2", gen9 -> "Gen 9"
  if(str_detect(var_name, "^gen\\d+$")) {
    return(paste0("Gen ", str_extract(var_name, "\\d+")))
  }
  # Primary type: class_primary9 -> "Type (1°): Fighting"
  if(str_detect(var_name, "^class_primary")) {
    idx <- str_extract(var_name, "\\d+")
    if(!is.na(idx) && idx %in% names(primary_type_names)) {
      return(paste0("Type (1°): ", primary_type_names[idx]))
    }
  }
  # Secondary type: class_secondary14 -> "Type (2°): Water"
  if(str_detect(var_name, "^class_secondary")) {
    idx <- str_extract(var_name, "\\d+")
    if(!is.na(idx) && idx %in% names(secondary_type_names)) {
      return(paste0("Type (2°): ", secondary_type_names[idx]))
    }
  }
  # Rarity: rarity1 -> "Rarity: SubLegendary"
  if(str_detect(var_name, "^rarity\\d+$")) {
    idx <- str_extract(var_name, "\\d+")
    if(!is.na(idx) && idx %in% names(rarity_names)) {
      return(paste0("Rarity: ", rarity_names[idx]))
    }
  }
  # Mega/Gigantamax
  if(var_name == "has_mega_evolution1") return("Has Mega Evolution")
  if(var_name == "has_gigantamax1") return("Has Gigantamax")
  # Against types: against_fire -> "vs Fire"
  if(str_detect(var_name, "^against_")) {
    type_name <- str_replace(var_name, "against_", "")
    return(paste0("vs ", str_to_title(type_name)))
  }
  # Default: return as-is but with nicer formatting
  return(str_to_title(str_replace_all(var_name, "_", " ")))
}

# Create coefficient dataframe with readable names
lasso_coefs <- data.frame(
  Variable_Code = rownames(coefs_min),
  Coef_LambdaMin = round(as.vector(coefs_min), 4),
  Coef_Lambda1SE = round(as.vector(coefs_1se), 4)
) %>%
  filter(Variable_Code != "(Intercept)") %>%
  mutate(
    Variable = sapply(Variable_Code, make_readable),
    Selected_Min = Coef_LambdaMin != 0,
    Selected_1SE = Coef_Lambda1SE != 0
  ) %>%
  arrange(desc(abs(Coef_LambdaMin)))

# Summary statistics
n_selected_min <- sum(lasso_coefs$Selected_Min)
n_selected_1se <- sum(lasso_coefs$Selected_1SE)
n_total <- nrow(lasso_coefs)

cat(paste0(
  "\n═══════════════════════════════════════════════════════════════════\n",
  "                 FEATURE SELECTION SUMMARY\n",
  "═══════════════════════════════════════════════════════════════════\n\n",
  "Total features:              ", n_total, "\n",
  "Selected (lambda.min):       ", n_selected_min, " (", round(100*n_selected_min/n_total, 1), "%)\n",
  "Selected (lambda.1se):       ", n_selected_1se, " (", round(100*n_selected_1se/n_total, 1), "%)\n",
  "\n--- Top 15 Features (by |coefficient| at lambda.min) ---\n"
))
print(head(lasso_coefs %>% 
             filter(Selected_Min) %>%
             arrange(desc(abs(Coef_LambdaMin))) %>%
             dplyr::select(Variable, Coef_LambdaMin, Selected_1SE), 15), 
      row.names = FALSE)

# Also show top features at lambda.1se (stricter selection)
lasso_coefs_1se <- lasso_coefs %>%
  filter(Selected_1SE) %>%
  arrange(desc(abs(Coef_Lambda1SE)))

cat(paste0(
  "\n--- Top 15 Features (by |coefficient| at lambda.1se - stricter) ---\n"
))
print(head(lasso_coefs_1se %>%
             dplyr::select(Variable, Coef_Lambda1SE), 15), 
      row.names = FALSE)

# Show the mapping for reference
cat(paste0(
  "\n--- Variable Name Key ---\n",
  "Type (1°) = Primary Type | Type (2°) = Secondary Type\n",
  "vs [Type] = Damage multiplier against that type\n",
  "Gen X = Generation X (vs Gen 1 baseline)\n"
))
```

```{r lasso-generation-check}
# KEY QUESTION: Does Generation survive LASSO?
gen_vars <- lasso_coefs %>% filter(str_detect(Variable_Code, "^gen"))

cat(paste0(
  "\n═══════════════════════════════════════════════════════════════════\n",
  "       CRITICAL: DOES GENERATION SURVIVE FEATURE SELECTION?\n",
  "═══════════════════════════════════════════════════════════════════\n\n",
  "Generation coefficients in LASSO:\n\n"
))
print(gen_vars %>% dplyr::select(Variable, Coef_LambdaMin, Coef_Lambda1SE, Selected_Min, Selected_1SE), 
      row.names = FALSE)

n_gen_min <- sum(gen_vars$Selected_Min)
n_gen_1se <- sum(gen_vars$Selected_1SE)

if(n_gen_1se > 0) {
  cat(paste0(
    "\n--- Summary ---\n",
    "Generation levels retained (lambda.min): ", n_gen_min, " / 8\n",
    "Generation levels retained (lambda.1se): ", n_gen_1se, " / 8\n\n",
    "✓ GENERATION SURVIVES feature selection!\n",
    "  → ", n_gen_1se, " generation level(s) have non-zero effect even in sparse model\n",
    "  → This supports the power creep hypothesis\n"
  ))
} else if(n_gen_min > 0) {
  cat(paste0(
    "\n--- Summary ---\n",
    "Generation levels retained (lambda.min): ", n_gen_min, " / 8\n",
    "Generation levels retained (lambda.1se): ", n_gen_1se, " / 8\n\n",
    "⚠ GENERATION is marginally important\n",
    "  → Retained at lambda.min but dropped at lambda.1se\n",
    "  → Effect exists but is weaker than other predictors\n"
  ))
} else {
  cat(paste0(
    "\n--- Summary ---\n",
    "Generation levels retained (lambda.min): ", n_gen_min, " / 8\n",
    "Generation levels retained (lambda.1se): ", n_gen_1se, " / 8\n\n",
    "✗ GENERATION does NOT survive feature selection\n",
    "  → All generation coefficients shrunk to zero\n",
    "  → Other predictors explain Attack better than generation\n"
  ))
}
```

```{r lasso-coefficient-plot, fig.width=14, fig.height=10}
# Visualize non-zero coefficients with readable names
lasso_nonzero <- lasso_coefs %>%
  filter(Selected_Min) %>%
  mutate(
    is_gen = str_detect(Variable_Code, "^gen"),
    Category = case_when(
      str_detect(Variable_Code, "^gen") ~ "Generation",
      str_detect(Variable_Code, "^against") ~ "Type Effectiveness",
      str_detect(Variable_Code, "^class") ~ "Type",
      Variable_Code %in% c("hp", "defense", "sp_attack", "sp_defense", "speed") ~ "Combat Stats",
      str_detect(Variable_Code, "^rarity") ~ "Rarity",
      TRUE ~ "Other"
    )
  )

ggplot(lasso_nonzero, aes(x = reorder(Variable, Coef_LambdaMin), 
                           y = Coef_LambdaMin, fill = Category)) +
  geom_bar(stat = "identity", alpha = 0.8) +
  geom_hline(yintercept = 0, color = "black", linewidth = 1) +
  coord_flip() +
  scale_fill_manual(values = c(
    "Generation" = "#E76F51",
    "Combat Stats" = "#2A9D8F",
    "Type Effectiveness" = "#E9C46A",
    "Type" = "#9B5DE5",
    "Rarity" = "#264653",
    "Other" = "#F4A261"
  )) +
  labs(title = "LASSO Selected Features (lambda.min)",
       subtitle = paste0(n_selected_min, " features retained out of ", n_total, 
                        "; Generation variables highlighted"),
       x = "Variable", y = "LASSO Coefficient") +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 8),
        legend.position = "right")
```

```{r lasso-regularization-path, fig.width=12, fig.height=6}
# Plot regularization path
lasso_full <- glmnet(X, y, alpha = 1)

# Extract generation variable indices
gen_indices <- which(str_detect(colnames(X), "^gen"))

# Custom plot with generation highlighted
plot(lasso_full, xvar = "lambda", label = FALSE,
     main = "LASSO Regularization Path")
abline(v = log(lambda_min), lty = 2, col = "blue")
abline(v = log(lambda_1se), lty = 2, col = "red")
legend("topright", legend = c("lambda.min", "lambda.1se"), 
       lty = 2, col = c("blue", "red"))
```

```{r lasso-importance-ranking}
# Create importance ranking
importance_df <- lasso_coefs %>%
  filter(Selected_Min) %>%
  mutate(
    Abs_Coef = abs(Coef_LambdaMin),
    Rank = rank(-Abs_Coef)
  ) %>%
  arrange(Rank) %>%
  head(20)

cat(paste0(
  "\n═══════════════════════════════════════════════════════════════════\n",
  "              TOP 20 MOST IMPORTANT PREDICTORS (LASSO)\n",
  "═══════════════════════════════════════════════════════════════════\n\n"
))
print(importance_df %>% dplyr::select(Rank, Variable, Coef_LambdaMin), row.names = FALSE)

# Find where generation ranks
gen_ranks <- importance_df %>% 
  filter(str_detect(Variable_Code, "^gen")) %>%
  pull(Rank)

if(length(gen_ranks) > 0) {
  cat(paste0("\n→ Generation variables rank: ", paste(gen_ranks, collapse = ", "), 
      " out of ", n_selected_min, " selected features\n"))
}
```

---

# 5. Hypothesis Testing: Generation Effect

## 5.1 Likelihood Ratio Test

```{r lr-test}
# Null model (without generation) - same predictors as model3 minus gen
model_null <- glm(make_formula("attack", predictors_no_gen), 
                  data = pokemon, family = gaussian(link = "identity"))

# LR test: Does generation matter?
lr_test <- anova(model_null, model3, test = "Chisq")

print(lr_test)
lr_p <- lr_test$`Pr(>Chi)`[2]
```

## 5.2 ANOVA Type III

```{r anova-typeiii}
# Type III ANOVA to assess each predictor's contribution
# Use lm instead of glm for proper ANOVA with Sum of Squares
model3_lm <- lm(make_formula("attack", all_predictors), data = pokemon)
anova_results <- Anova(model3_lm, type = "III")

print(anova_results)

# Calculate partial eta-squared
# Get the residual SS (last row)
residual_ss <- anova_results$`Sum Sq`[nrow(anova_results)]

eta_sq <- data.frame(
  Predictor = rownames(anova_results),
  SS = anova_results$`Sum Sq`
) %>%
  filter(Predictor != "Residuals") %>%
  mutate(
    Partial_Eta_Sq = SS / (SS + residual_ss)
  ) %>%
  arrange(desc(Partial_Eta_Sq)) %>%
  mutate(across(where(is.numeric), ~round(., 4)))

eta_sq
```

## 5.3 Jonckheere-Terpstra Test for Monotonic Trend

The Jonckheere-Terpstra test specifically tests for an **ordered alternative hypothesis** — whether Attack systematically increases (or decreases) across generations. This is more powerful than ANOVA when we expect a directional "creep."

```{r jonckheere-test}
# Jonckheere-Terpstra test for monotonic trend
# H0: No trend across ordered groups
# H1: Monotonic increasing (or decreasing) trend

# Prepare data - ensure generation is ordered
pokemon_ordered <- pokemon %>%
  arrange(gen_numeric)

# Perform Jonckheere-Terpstra test
jt_result <- JonckheereTerpstraTest(
  x = pokemon_ordered$attack,
  g = pokemon_ordered$gen_numeric,
  alternative = "increasing"  # Test for increasing trend (power creep)
)

if(jt_result$p.value < 0.05) {
  conclusion <- paste0(
    "✓ SIGNIFICANT MONOTONIC TREND DETECTED (p < 0.05)\n",
    "  → Attack values show a systematic increase across generations\n",
    "  → This directly supports the 'power creep' hypothesis\n"
  )
} else if(jt_result$p.value < 0.10) {
  conclusion <- paste0(
    "⚠ MARGINALLY SIGNIFICANT TREND (0.05 < p < 0.10)\n",
    "  → Weak evidence for increasing trend\n"
  )
} else {
  conclusion <- paste0(
    "✗ NO SIGNIFICANT MONOTONIC TREND (p ≥ 0.10)\n",
    "  → Attack does not systematically increase across generations\n"
  )
}

cat(paste0(
  "\n═══════════════════════════════════════════════════════════════════\n",
  "         JONCKHEERE-TERPSTRA TEST FOR MONOTONIC TREND\n",
  "═══════════════════════════════════════════════════════════════════\n\n",
  "Test Hypothesis:\n",
  "  H₀: Attack distribution is the same across all generations\n",
  "  H₁: Attack shows a monotonic INCREASING trend (Gen 1 → Gen 9)\n\n",
  "Results:\n",
  "  JT Statistic:  ", jt_result$statistic, "\n",
  "  p-value:       ", format.pval(jt_result$p.value, digits = 4), "\n",
  "  Alternative:   ", jt_result$alternative, "\n\n",
  conclusion
))
```

```{r jt-visualization, fig.width=12, fig.height=5}
# Visualize the trend with regression line
gen_means_ordered <- pokemon %>%
  group_by(gen_numeric) %>%
  summarise(
    mean_attack = mean(attack),
    se = sd(attack) / sqrt(n()),
    n = n()
  )

# Linear trend
trend_model <- lm(mean_attack ~ gen_numeric, data = gen_means_ordered)
trend_coef <- coef(trend_model)[2]
trend_r2 <- summary(trend_model)$r.squared

ggplot(gen_means_ordered, aes(x = gen_numeric, y = mean_attack)) +
  geom_point(aes(size = n), color = "#E76F51", alpha = 0.8) +
  geom_errorbar(aes(ymin = mean_attack - 1.96*se, ymax = mean_attack + 1.96*se),
                width = 0.2, color = "#264653") +
  geom_smooth(method = "lm", se = TRUE, color = "#2A9D8F", linewidth = 1.5) +
  scale_size_continuous(range = c(4, 12), name = "Sample Size") +
  scale_x_continuous(breaks = 1:9) +
  labs(title = "Attack Trend Across Generations",
       subtitle = paste0("JT test p = ", format.pval(jt_result$p.value, digits = 3),
                        " | Linear trend: +", round(trend_coef, 2), " Attack/generation",
                        " | R² = ", round(trend_r2, 3)),
       x = "Generation", y = "Mean Attack (± 95% CI)") +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 12, face = "bold"))
```

```{r jt-two-sided}
# Also test two-sided for completeness
jt_twosided <- JonckheereTerpstraTest(
  x = pokemon_ordered$attack,
  g = pokemon_ordered$gen_numeric,
  alternative = "two.sided"
)

# Summary table
jt_summary <- data.frame(
  Test = c("One-sided (increasing)", "Two-sided"),
  JT_Statistic = c(jt_result$statistic, jt_twosided$statistic),
  p_value = c(jt_result$p.value, jt_twosided$p.value),
  Conclusion = c(
    ifelse(jt_result$p.value < 0.05, "Significant ✓", "Not significant"),
    ifelse(jt_twosided$p.value < 0.05, "Significant ✓", "Not significant")
  )
)
jt_summary$p_value <- format.pval(jt_summary$p_value, digits = 4)
print(jt_summary, row.names = FALSE)
```

## 5.4 Generation Coefficients Visualization

```{r gen-coef-viz, fig.width=10, fig.height=6}
# Extract generation coefficients from full model
gen_coefs <- tidy(model3) %>%
  filter(str_detect(term, "^gen")) %>%
  mutate(
    generation = as.numeric(str_extract(term, "\\d+")),
    ci_lower = estimate - 1.96 * std.error,
    ci_upper = estimate + 1.96 * std.error,
    significant = p.value < 0.05
  )

# Add Gen 1 reference
gen_coefs_plot <- bind_rows(
  data.frame(term = "gen1", estimate = 0, std.error = 0, statistic = NA, 
             p.value = NA, generation = 1, ci_lower = 0, ci_upper = 0,
             significant = NA),
  gen_coefs
)

ggplot(gen_coefs_plot, aes(x = factor(generation), y = estimate, fill = significant)) +
  geom_bar(stat = "identity", alpha = 0.8, color = "black") +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.3, linewidth = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red", linewidth = 1.2) +
  scale_fill_manual(values = c("TRUE" = "#2A9D8F", "FALSE" = "#E9C46A"),
                    na.value = "#264653",
                    labels = c("TRUE" = "p < 0.05", "FALSE" = "p ≥ 0.05"),
                    na.translate = TRUE) +
  labs(title = "Generation Effects on Attack (Full Model)",
       subtitle = "Coefficients represent difference from Gen 1; Error bars = 95% CI",
       x = "Generation", y = "Coefficient (Attack Points)",
       fill = "Significance") +
  theme_minimal() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(size = 12, face = "bold"))
```

---

# 6. Robustness Checks

## 6.1 Robust Regression (Huber M-estimation)

```{r robust-regression}
# Robust regression using Huber M-estimation (full model)
model_robust <- rlm(make_formula("attack", all_predictors), 
                    data = pokemon, method = "M")

# Compare OLS vs Robust coefficients
ols_coefs <- coef(model3)
robust_coefs <- coef(model_robust)

comparison <- data.frame(
  Term = names(ols_coefs),
  OLS = round(ols_coefs, 3),
  Robust = round(robust_coefs, 3),
  Difference = round(ols_coefs - robust_coefs, 3),
  Pct_Diff = round(100 * (ols_coefs - robust_coefs) / abs(ols_coefs), 1)
) %>%
  filter(str_detect(Term, "^gen"))

print(comparison, row.names = FALSE)

max_pct_diff <- max(abs(comparison$Pct_Diff), na.rm = TRUE)
```

## 6.2 Robustness Comparison Plot

```{r robustness-plot, fig.width=12, fig.height=6}
# Compare OLS vs Robust regression
robustness_data <- bind_rows(
  gen_coefs %>% 
    dplyr::select(generation, estimate) %>% 
    mutate(Model = "GLM (OLS)"),
  data.frame(generation = 2:9, estimate = robust_coefs[grep("^gen", names(robust_coefs))],
             Model = "Robust (Huber)")
)

# Add Gen 1 reference
robustness_data <- bind_rows(
  data.frame(generation = 1, estimate = 0, Model = c("GLM (OLS)", "Robust (Huber)")),
  robustness_data
)

ggplot(robustness_data, aes(x = factor(generation), y = estimate, fill = Model)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8), alpha = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red", linewidth = 1) +
  scale_fill_manual(values = c("GLM (OLS)" = "#2A9D8F", "Robust (Huber)" = "#E76F51")) +
  labs(title = "Robustness Check: OLS vs Robust Regression",
       subtitle = "Similar coefficients indicate outliers are not driving results",
       x = "Generation", y = "Coefficient (Attack Points)") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

---

# 7. Secondary Analyses: Subgroup Trends

## 7.1 Type-Specific Power Creep Analysis

Do all Pokémon types experience equal power creep, or are some types driving the trend?

```{r type-trends-data}
# Calculate mean Attack by Type and Generation (type_name already created in preprocessing)
type_gen_trends <- pokemon %>%
  group_by(type_name, gen_numeric) %>%
  summarise(
    n = n(),
    mean_attack = mean(attack),
    sd_attack = sd(attack),
    se = sd(attack) / sqrt(n()),
    .groups = "drop"
  )

# Count types and their coverage across generations
type_coverage <- type_gen_trends %>%
  group_by(type_name) %>%
  summarise(
    n_gens = n_distinct(gen_numeric),
    total_n = sum(n)
  ) %>%
  arrange(desc(total_n))

cat("All types and their data coverage:\n")
print(type_coverage, n = Inf)
```

```{r type-trends-regression}
# Fit linear trend for each type (all types included)
type_trends <- pokemon %>%
  group_by(type_name) %>%
  do({
    model <- lm(attack ~ gen_numeric, data = .)
    data.frame(
      slope = coef(model)[2],
      intercept = coef(model)[1],
      p_value = summary(model)$coefficients[2, 4],
      r_squared = summary(model)$r.squared,
      n = nrow(.)
    )
  }) %>%
  ungroup() %>%
  mutate(
    significant = p_value < 0.05,
    direction = case_when(
      slope > 0 & significant ~ "Increasing ↑",
      slope < 0 & significant ~ "Decreasing ↓",
      TRUE ~ "No trend"
    )
  ) %>%
  arrange(desc(slope))

cat(paste0(
  "\n═══════════════════════════════════════════════════════════════════\n",
  "              TYPE-SPECIFIC ATTACK TRENDS\n",
  "═══════════════════════════════════════════════════════════════════\n\n",
  "Trend = Change in Attack per generation (slope of linear regression)\n\n"
))
print(type_trends %>% 
        mutate(across(c(slope, intercept, r_squared), ~round(., 3)),
               p_value = format.pval(p_value, digits = 3)) %>%
        dplyr::select(type_name, slope, p_value, direction, n), 
      n = Inf)

# Summary
n_increasing <- sum(type_trends$direction == "Increasing ↑")
n_decreasing <- sum(type_trends$direction == "Decreasing ↓")
n_none <- sum(type_trends$direction == "No trend")

cat(paste0(
  "\n--- Summary ---\n",
  "Types with significant INCREASING trend: ", n_increasing, "\n",
  "Types with significant DECREASING trend: ", n_decreasing, "\n",
  "Types with NO significant trend:         ", n_none, "\n"
))
```

```{r type-trends-plot, fig.width=14, fig.height=8}
# Plot trends for SIGNIFICANT types only
significant_types <- type_trends %>%
  filter(direction != "No trend") %>%
  pull(type_name)

n_sig <- length(significant_types)

if(n_sig > 0) {
  type_plot_data <- pokemon %>%
    filter(type_name %in% significant_types) %>%
    group_by(type_name, gen_numeric) %>%
    summarise(mean_attack = mean(attack), n = n(), .groups = "drop") %>%
    # Add direction info for coloring
    left_join(type_trends %>% dplyr::select(type_name, direction, slope, p_value), by = "type_name")
  
  ggplot(type_plot_data, aes(x = gen_numeric, y = mean_attack, color = direction)) +
    geom_line(linewidth = 1.2, alpha = 0.8) +
    geom_point(aes(size = n), alpha = 0.8) +
    geom_smooth(method = "lm", se = TRUE, linetype = "dashed", linewidth = 1, alpha = 0.2) +
    scale_x_continuous(breaks = 1:9) +
    scale_size_continuous(range = c(2, 6), name = "n Pokémon") +
    scale_color_manual(values = c("Increasing ↑" = "#2A9D8F", "Decreasing ↓" = "#E76F51"),
                       name = "Trend Direction") +
    facet_wrap(~type_name, ncol = 3, scales = "free_y") +
    labs(title = paste0("Types with SIGNIFICANT Attack Trends (", n_sig, " types)"),
         subtitle = "Only showing types with p < 0.05; Shaded area = 95% CI; Green = increasing, Red = decreasing",
         x = "Generation", y = "Mean Attack") +
    theme_minimal() +
    theme(legend.position = "bottom",
          strip.text = element_text(face = "bold", size = 11),
          axis.text.x = element_text(size = 9))
} else {
  cat("No types show statistically significant trends at p < 0.05\n")
}
```

```{r type-trends-plot-marginal, fig.width=14, fig.height=8}
# Plot trends for MARGINALLY SIGNIFICANT types (0.05 ≤ p < 0.1)
marginal_types <- type_trends %>%
  filter(p_value >= 0.05 & p_value < 0.1) %>%
  mutate(
    direction_marginal = case_when(
      slope > 0 ~ "Increasing (marginal)",
      slope < 0 ~ "Decreasing (marginal)",
      TRUE ~ "No trend"
    )
  )

n_marginal <- nrow(marginal_types)

if(n_marginal > 0) {
  type_plot_marginal <- pokemon %>%
    filter(type_name %in% marginal_types$type_name) %>%
    group_by(type_name, gen_numeric) %>%
    summarise(mean_attack = mean(attack), n = n(), .groups = "drop") %>%
    left_join(marginal_types %>% dplyr::select(type_name, direction_marginal, slope, p_value), 
              by = "type_name")
  
  ggplot(type_plot_marginal, aes(x = gen_numeric, y = mean_attack, color = direction_marginal)) +
    geom_line(linewidth = 1.2, alpha = 0.8) +
    geom_point(aes(size = n), alpha = 0.8) +
    geom_smooth(method = "lm", se = TRUE, linetype = "dashed", linewidth = 1, alpha = 0.2) +
    scale_x_continuous(breaks = 1:9) +
    scale_size_continuous(range = c(2, 6), name = "n Pokémon") +
    scale_color_manual(values = c("Increasing (marginal)" = "#F4A261", 
                                   "Decreasing (marginal)" = "#E9C46A"),
                       name = "Trend Direction") +
    facet_wrap(~type_name, ncol = 3, scales = "free_y") +
    labs(title = paste0("Types with MARGINALLY Significant Trends (", n_marginal, " types)"),
         subtitle = "Types with 0.05 ≤ p < 0.1; Orange = increasing, Yellow = decreasing; Worth monitoring",
         x = "Generation", y = "Mean Attack") +
    theme_minimal() +
    theme(legend.position = "bottom",
          strip.text = element_text(face = "bold", size = 11),
          axis.text.x = element_text(size = 9))
} else {
  cat("No types show marginally significant trends (0.05 ≤ p < 0.1)\n")
}
```

```{r type-trends-plot-nonsig-strict, fig.width=14, fig.height=12}
# Plot trends for NON-SIGNIFICANT types (p >= 0.1)
nonsig_strict_types <- type_trends %>%
  filter(p_value >= 0.1) %>%
  pull(type_name)

n_nonsig_strict <- length(nonsig_strict_types)

if(n_nonsig_strict > 0) {
  type_plot_nonsig_strict <- pokemon %>%
    filter(type_name %in% nonsig_strict_types) %>%
    group_by(type_name, gen_numeric) %>%
    summarise(mean_attack = mean(attack), n = n(), .groups = "drop") %>%
    left_join(type_trends %>% dplyr::select(type_name, slope, p_value), by = "type_name")
  
  ggplot(type_plot_nonsig_strict, aes(x = gen_numeric, y = mean_attack)) +
    geom_line(linewidth = 1, alpha = 0.7, color = "#6c757d") +
    geom_point(aes(size = n), alpha = 0.7, color = "#6c757d") +
    geom_smooth(method = "lm", se = TRUE, linetype = "dashed", 
                linewidth = 0.8, color = "#adb5bd", fill = "#dee2e6", alpha = 0.3) +
    scale_x_continuous(breaks = 1:9) +
    scale_size_continuous(range = c(2, 5), name = "n Pokémon") +
    facet_wrap(~type_name, ncol = 4, scales = "free_y") +
    labs(title = paste0("Types with NO Significant Trend (", n_nonsig_strict, " types)"),
         subtitle = "These types show no change in Attack across generations (p ≥ 0.1)",
         x = "Generation", y = "Mean Attack") +
    theme_minimal() +
    theme(legend.position = "bottom",
          strip.text = element_text(face = "bold", size = 10),
          axis.text.x = element_text(size = 8))
} else {
  cat("All types show at least marginally significant trends!\n")
}
```

```{r type-trends-heatmap, fig.width=14, fig.height=10}
# Heatmap of Attack by Type and Generation (all types)
# Color = mean Attack, Dot size = sample size
heatmap_data <- pokemon %>%
  group_by(type_name, gen_numeric) %>%
  summarise(mean_attack = mean(attack), n = n(), .groups = "drop")

ggplot(heatmap_data, aes(x = factor(gen_numeric), y = type_name)) +
  # Background tiles for color
  geom_tile(aes(fill = mean_attack), color = "white", linewidth = 0.5) +
  # Dots sized by sample count
  geom_point(aes(size = n), color = "white", alpha = 0.8) +
  # Text labels for mean attack values
  geom_text(aes(label = round(mean_attack, 0)), color = "black", size = 2.5, 
            fontface = "bold", vjust = -1.5) +
  # Scales
  scale_fill_viridis_c(option = "plasma", name = "Mean Attack") +
  scale_size_continuous(range = c(1, 12), name = "Sample Size (n)", 
                        breaks = c(5, 15, 30, 50)) +
  labs(title = "Attack Heatmap: Type × Generation",
       subtitle = "Color = Mean Attack | Dot size = Number of Pokémon | Larger dots = more reliable estimates",
       x = "Generation", y = "Primary Type") +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 10, face = "bold"),
        axis.text.x = element_text(size = 10, face = "bold"),
        panel.grid = element_blank(),
        legend.position = "right")
```

```{r type-trends-heatmap-alt, fig.width=14, fig.height=10}
# Alternative: Bubble chart version (cleaner for seeing sample sizes)
ggplot(heatmap_data, aes(x = factor(gen_numeric), y = type_name)) +
  geom_point(aes(size = n, fill = mean_attack), shape = 21, color = "black", alpha = 0.85) +
  scale_fill_viridis_c(option = "plasma", name = "Mean Attack") +
  scale_size_continuous(range = c(2, 15), name = "Sample Size (n)",
                        breaks = c(5, 10, 20, 40)) +
  labs(title = "Attack by Type × Generation (Bubble Chart)",
       subtitle = "Bubble size = sample count | Color = mean Attack | Missing bubbles = no Pokémon of that type in that generation",
       x = "Generation", y = "Primary Type") +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 10, face = "bold"),
        axis.text.x = element_text(size = 10, face = "bold"),
        panel.grid.major = element_line(color = "gray90"),
        legend.position = "right")
```

```{r type-jt-tests}
# Jonckheere-Terpstra test for each type (all types)
type_jt_results <- pokemon %>%
  group_by(type_name) %>%
  do({
    ordered_data <- arrange(., gen_numeric)
    jt <- JonckheereTerpstraTest(
      x = ordered_data$attack,
      g = ordered_data$gen_numeric,
      alternative = "increasing"
    )
    data.frame(
      JT_statistic = jt$statistic,
      p_value = jt$p.value,
      n = nrow(.)
    )
  }) %>%
  ungroup() %>%
  mutate(
    significant = p_value < 0.05,
    conclusion = ifelse(significant, "Power Creep ✓", "No Creep")
  ) %>%
  arrange(p_value)

cat(paste0(
  "\n═══════════════════════════════════════════════════════════════════\n",
  "         JONCKHEERE-TERPSTRA TEST BY TYPE\n",
  "═══════════════════════════════════════════════════════════════════\n\n"
))
print(type_jt_results %>%
        mutate(p_value = format.pval(p_value, digits = 3)), 
      n = Inf)

cat(paste0("\n→ Types with significant monotonic trend: ", 
    sum(type_jt_results$significant), " / ", nrow(type_jt_results), "\n"))
```

## 7.2 Rarity-Specific Power Creep Analysis

Is power creep occurring in competitive (Normal) Pokémon or mainly in Legendaries?

```{r rarity-trends-data}
# Calculate trends by rarity
rarity_gen_trends <- pokemon %>%
  group_by(rarity, gen_numeric) %>%
  summarise(
    n = n(),
    mean_attack = mean(attack),
    sd_attack = sd(attack),
    se = sd(attack) / sqrt(n()),
    .groups = "drop"
  )

print(rarity_gen_trends, n = Inf)
```

```{r rarity-trends-regression}
# Fit linear trend for each rarity
rarity_trends <- pokemon %>%
  group_by(rarity) %>%
  do({
    model <- lm(attack ~ gen_numeric, data = .)
    data.frame(
      slope = coef(model)[2],
      intercept = coef(model)[1],
      p_value = summary(model)$coefficients[2, 4],
      r_squared = summary(model)$r.squared,
      n = nrow(.)
    )
  }) %>%
  ungroup() %>%
  mutate(
    significant = p_value < 0.05,
    direction = case_when(
      slope > 0 & significant ~ "Increasing ↑",
      slope < 0 & significant ~ "Decreasing ↓",
      TRUE ~ "No trend"
    ),
    attack_per_gen = round(slope, 2)
  )

cat(paste0(
  "\n═══════════════════════════════════════════════════════════════════\n",
  "              RARITY-SPECIFIC ATTACK TRENDS\n",
  "═══════════════════════════════════════════════════════════════════\n\n"
))
print(rarity_trends %>%
        mutate(across(c(slope, intercept, r_squared), ~round(., 3)),
               p_value = format.pval(p_value, digits = 3)) %>%
        dplyr::select(rarity, slope, p_value, direction, n),
      row.names = FALSE)

# Interpretation
interp_lines <- sapply(1:nrow(rarity_trends), function(i) {
  r <- rarity_trends[i, ]
  paste0(as.character(r$rarity), ": ", 
         ifelse(r$significant, 
                paste0("Attack changes by ", r$attack_per_gen, " points per generation"),
                "No significant trend"))
})
cat(paste0("\n--- Interpretation ---\n", paste(interp_lines, collapse = "\n"), "\n"))
```

```{r rarity-trends-plot, fig.width=12, fig.height=6}
# Plot by rarity
ggplot(rarity_gen_trends, aes(x = gen_numeric, y = mean_attack, color = rarity)) +
  geom_line(linewidth = 1.5) +
  geom_point(aes(size = n), alpha = 0.8) +
  geom_ribbon(aes(ymin = mean_attack - 1.96*se, ymax = mean_attack + 1.96*se, 
                  fill = rarity), alpha = 0.2, color = NA) +
  geom_smooth(method = "lm", se = FALSE, linetype = "dashed", linewidth = 0.8) +
  scale_x_continuous(breaks = 1:9) +
  scale_color_brewer(palette = "Set2") +
  scale_fill_brewer(palette = "Set2", guide = "none") +
  scale_size_continuous(range = c(3, 8), name = "n Pokémon") +
  labs(title = "Attack Trends by Rarity Category",
       subtitle = "Shaded area = 95% CI; Dashed line = linear trend",
       x = "Generation", y = "Mean Attack",
       color = "Rarity") +
  theme_minimal() +
  theme(legend.position = "right")
```

```{r rarity-jt-tests}
# Jonckheere-Terpstra test by rarity
rarity_jt_results <- pokemon %>%
  group_by(rarity) %>%
  do({
    ordered_data <- arrange(., gen_numeric)
    jt <- JonckheereTerpstraTest(
      x = ordered_data$attack,
      g = ordered_data$gen_numeric,
      alternative = "increasing"
    )
    data.frame(
      JT_statistic = jt$statistic,
      p_value = jt$p.value,
      n = nrow(.)
    )
  }) %>%
  ungroup() %>%
  mutate(
    significant = p_value < 0.05,
    conclusion = ifelse(significant, "Power Creep ✓", "No Creep")
  )

cat(paste0(
  "\n═══════════════════════════════════════════════════════════════════\n",
  "         JONCKHEERE-TERPSTRA TEST BY RARITY\n",
  "═══════════════════════════════════════════════════════════════════\n\n"
))
print(rarity_jt_results %>%
        mutate(p_value = format.pval(p_value, digits = 4)),
      row.names = FALSE)

# Key finding
normal_jt <- rarity_jt_results %>% filter(rarity == "Normal")
if(normal_jt$significant) {
  key_finding <- paste0(
    "✓ POWER CREEP affects NORMAL Pokémon\n",
    "  → This has direct competitive implications!\n"
  )
} else {
  key_finding <- paste0(
    "✗ Power creep NOT significant in Normal Pokémon\n",
    "  → Creep may be driven by Legendaries/Mythicals\n"
  )
}
cat(paste0(
  "\n--- KEY FINDING: Normal (Competitive) Pokémon ---\n",
  "JT statistic: ", normal_jt$JT_statistic, "\n",
  "p-value:      ", format.pval(normal_jt$p_value, digits = 4), "\n",
  key_finding
))
```

```{r rarity-faceted-boxplot, fig.width=14, fig.height=8}
# Faceted boxplots
ggplot(pokemon, aes(x = factor(gen_numeric), y = attack, fill = rarity)) +
  geom_boxplot(alpha = 0.7, outlier.size = 1) +
  stat_summary(fun = mean, geom = "point", shape = 18, size = 3, color = "black") +
  scale_fill_brewer(palette = "Set2") +
  facet_wrap(~rarity, ncol = 2, scales = "free_y") +
  labs(title = "Attack Distribution by Generation and Rarity",
       subtitle = "Diamonds = means; Compare across generations within each rarity",
       x = "Generation", y = "Attack") +
  theme_minimal() +
  theme(legend.position = "none",
        strip.text = element_text(face = "bold", size = 12))
```

## 7.3 Secondary Analysis Summary

```{r secondary-summary}
# Type summary
top_creep_types <- type_trends %>% 
  filter(direction == "Increasing ↑") %>%
  head(3) %>%
  pull(type_name)

type_summary <- paste0(
  "TYPE-SPECIFIC TRENDS:\n",
  "  • Types with significant power creep: ", n_increasing, "\n",
  if(length(top_creep_types) > 0) paste0("  • Strongest creep in: ", paste(top_creep_types, collapse = ", "), "\n") else ""
)

# Rarity summary
rarity_lines <- sapply(1:nrow(rarity_jt_results), function(i) {
  r <- rarity_jt_results[i, ]
  paste0("  • ", as.character(r$rarity), ": ", r$conclusion, 
         " (p = ", format.pval(r$p_value, digits = 3), ")")
})
rarity_summary <- paste0("RARITY-SPECIFIC TRENDS:\n", paste(rarity_lines, collapse = "\n"), "\n")

# Overall conclusion
if(normal_jt$significant) {
  conclusion <- paste0(
    "CONCLUSION:\n",
    "  Power creep affects Normal (competitive) Pokémon,\n",
    "  meaning it has real implications for game balance.\n"
  )
} else {
  conclusion <- paste0(
    "CONCLUSION:\n",
    "  Power creep may be concentrated in Legendary/Mythical Pokémon,\n",
    "  with less impact on competitive play.\n"
  )
}

cat(paste0(
  "\n═══════════════════════════════════════════════════════════════════\n",
  "           SECONDARY ANALYSES SUMMARY\n",
  "═══════════════════════════════════════════════════════════════════\n\n",
  type_summary, "\n",
  rarity_summary, "\n",
  conclusion
))
```

---

# 8. Effect Size Analysis

```{r effect-sizes}
# Calculate Cohen's d for each generation vs Gen 1
gen1_attack <- pokemon$attack[pokemon$gen == "1"]

cohens_d <- data.frame(
  Generation = 2:9,
  Cohens_d = sapply(2:9, function(g) {
    gen_attack <- pokemon$attack[pokemon$gen == as.character(g)]
    mean_diff <- mean(gen_attack) - mean(gen1_attack)
    pooled_sd <- sqrt(((length(gen1_attack)-1)*var(gen1_attack) + 
                        (length(gen_attack)-1)*var(gen_attack)) / 
                       (length(gen1_attack) + length(gen_attack) - 2))
    return(mean_diff / pooled_sd)
  })
) %>%
  mutate(
    Magnitude = case_when(
      abs(Cohens_d) < 0.2 ~ "Negligible",
      abs(Cohens_d) < 0.5 ~ "Small",
      abs(Cohens_d) < 0.8 ~ "Medium",
      TRUE ~ "Large"
    ),
    Cohens_d = round(Cohens_d, 3)
  )

cohens_d
```

```{r effect-size-plot, fig.width=10, fig.height=5}
# Add Gen 1 reference
cohens_d_plot <- bind_rows(
  data.frame(Generation = 1, Cohens_d = 0, Magnitude = "Reference"),
  cohens_d
)

ggplot(cohens_d_plot, aes(x = factor(Generation), y = Cohens_d, fill = Magnitude)) +
  geom_bar(stat = "identity", alpha = 0.8, color = "black") +
  geom_hline(yintercept = c(-0.2, 0.2), linetype = "dotted", color = "gray50") +
  geom_hline(yintercept = c(-0.5, 0.5), linetype = "dashed", color = "gray50") +
  geom_hline(yintercept = 0, color = "red", linewidth = 1) +
  scale_fill_manual(values = c("Reference" = "#264653", "Negligible" = "#E9C46A", 
                               "Small" = "#F4A261", "Medium" = "#E76F51", 
                               "Large" = "#9B5DE5")) +
  labs(title = "Practical Significance: Cohen's d by Generation",
       subtitle = "Comparing each generation's Attack to Generation 1",
       x = "Generation", y = "Cohen's d",
       fill = "Effect Size") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

---

# 9. Prediction Analysis

```{r prediction-performance, fig.width=12, fig.height=5}
# Predictions from the full model
pokemon$predicted_attack <- predict(model3, type = "response")
pokemon$residual <- pokemon$attack - pokemon$predicted_attack

# Prediction metrics
rmse <- sqrt(mean(pokemon$residual^2))
mae <- mean(abs(pokemon$residual))
r_squared <- 1 - sum(pokemon$residual^2) / sum((pokemon$attack - mean(pokemon$attack))^2)

data.frame(
  Metric = c("RMSE", "MAE", "R²", "Mean Attack", "RMSE % of Mean"),
  Value = c(round(rmse, 2), round(mae, 2), round(r_squared, 4), 
            round(mean(pokemon$attack), 2), round(100 * rmse / mean(pokemon$attack), 1))
)

# Predicted vs Actual plot
p_pred <- ggplot(pokemon, aes(x = attack, y = predicted_attack, color = rarity)) +
  geom_point(alpha = 0.6, size = 2) +
  geom_abline(slope = 1, intercept = 0, color = "red", linewidth = 1.2, linetype = "dashed") +
  scale_color_brewer(palette = "Set2") +
  labs(title = "Predicted vs Actual Attack",
       subtitle = paste0("R² = ", round(r_squared, 3), "; RMSE = ", round(rmse, 1)),
       x = "Actual Attack", y = "Predicted Attack") +
  theme_minimal()

# Residual distribution
p_resid <- ggplot(pokemon, aes(x = residual)) +
  geom_histogram(bins = 35, fill = "#2A9D8F", alpha = 0.7, color = "white") +
  geom_vline(xintercept = 0, color = "red", linewidth = 1) +
  labs(title = "Distribution of Residuals",
       subtitle = paste0("Mean = ", round(mean(pokemon$residual), 2), 
                        ", SD = ", round(sd(pokemon$residual), 2)),
       x = "Residual (Actual - Predicted)", y = "Count") +
  theme_minimal()

grid.arrange(p_pred, p_resid, ncol = 2)
```

---

# 10. Conclusions

```{r final-summary}
# Build summary statistics
lr_sig <- lr_test$`Pr(>Chi)`[2] < 0.05
n_pos_sig <- sum(gen_coefs$estimate > 0 & gen_coefs$p.value < 0.05, na.rm = TRUE)
n_neg_sig <- sum(gen_coefs$estimate < 0 & gen_coefs$p.value < 0.05, na.rm = TRUE)
max_d <- max(abs(cohens_d$Cohens_d))
max_d_gen <- cohens_d$Generation[which.max(abs(cohens_d$Cohens_d))]
max_d_mag <- cohens_d$Magnitude[which.max(abs(cohens_d$Cohens_d))]

# Summary table
summary_df <- data.frame(
  Category = c("Target Variable", "", "", "",
               "Best Model", "", "",
               "Generation Effect", "", "",
               "Effect Size", ""),
  Metric = c("Mean", "SD", "Range", "",
             "Model", "R²", "RMSE",
             "LR Test Significant?", "Gens higher than Gen 1", "Gens lower than Gen 1",
             "Max Cohen's d", "Magnitude"),
  Value = c(round(mean(pokemon$attack), 1), round(sd(pokemon$attack), 1),
            paste0(min(pokemon$attack), "-", max(pokemon$attack)), "",
            "Full Gaussian GLM", round(r_squared, 4), round(rmse, 1),
            ifelse(lr_sig, "Yes", "No"), n_pos_sig, n_neg_sig,
            round(max_d, 2), paste0("Gen ", max_d_gen, " (", max_d_mag, ")"))
)
print(summary_df, row.names = FALSE)
```

## Key Findings

```{r conclusions-data, echo=FALSE}
# Gather key statistics for conclusions
lasso_gen_survived <- sum(gen_vars$Selected_1SE)
jt_sig <- jt_result$p.value < 0.05
normal_creep <- rarity_jt_results %>% filter(rarity == "Normal") %>% pull(significant)
```

### 1. Evidence of Power Creep in Attack

**Yes, there is statistically significant evidence of power creep in Attack stats.**

#### GLM Evidence:
In the simple model (Attack ~ Generation), later generations show significantly higher Attack. When controlling for all 38 predictors in the full model, some generation effects remain significant, suggesting that part of the "power creep" is explained by other factors (type composition, legendary distribution, etc.).

#### Monotonic Trend (Jonckheere-Terpstra Test):
The J-T test specifically tests whether Attack *systematically increases* across generations (not just differs). This directly addresses the "creep" hypothesis.

- **JT Statistic:** `r jt_result$statistic`
- **p-value:** `r format.pval(jt_result$p.value, digits = 4)`
- **Conclusion:** `r ifelse(jt_sig, "✓ Significant monotonic increasing trend detected", "✗ No significant monotonic trend")`

### 2. LASSO Feature Selection: Does Generation Matter?

LASSO regularization objectively determines which predictors are most important when they compete.

| Model | Features Selected | Generation Retained? |
|-------|-------------------|---------------------|
| Lambda.min | `r n_selected_min` / `r n_total` | `r ifelse(n_gen_min > 0, paste0("Yes (", n_gen_min, " levels)"), "No")` |
| Lambda.1SE (sparse) | `r n_selected_1se` / `r n_total` | `r ifelse(n_gen_1se > 0, paste0("Yes (", n_gen_1se, " levels)"), "No")` |

**Key Insight:** `r ifelse(n_gen_1se > 0, "Generation SURVIVES automatic feature selection, supporting its importance as a predictor", "Generation is dropped in the sparse model, suggesting other predictors explain Attack better")`

### 3. Model Performance

- **R² = `r round(r_squared * 100, 1)`%** — The full model explains approximately 60% of Attack variance
- **RMSE = `r round(rmse, 1)` points** — Average prediction error is about 19 Attack points

### 4. Most Important Predictors (LASSO Ranking)

Based on LASSO regularization, the strongest predictors are combat stats (Defense, HP, Speed), followed by type characteristics and rarity. Generation effects are `r ifelse(length(gen_ranks) > 0, paste0("ranked ", paste(gen_ranks, collapse = ", "), " out of ", n_selected_min, " selected features"), "not among the top predictors")`.

### 5. Secondary Analyses: Who Gets Power Creep?

#### Type-Specific Trends:
- **`r n_increasing`** types show significant increasing Attack trends
- **`r n_none`** types show no significant trend
- Power creep is **not uniform** across all Pokémon types

#### Rarity-Specific Trends:
```{r rarity-conclusion-table, echo=FALSE}
rarity_jt_results %>%
  mutate(p_value = format.pval(p_value, digits = 3)) %>%
  dplyr::select(rarity, conclusion, p_value) %>%
  knitr::kable()
```

**Critical Finding:** `r ifelse(normal_creep, "Power creep affects NORMAL (competitive) Pokémon — this has real implications for game balance!", "Power creep may be concentrated in Legendary/Mythical Pokémon, with less impact on competitive play.")`

### 6. Effect Sizes (Practical Significance)

**Maximum effect: Gen 7 with d = `r round(max_d, 2)`** — A "`r max_d_mag`" effect by Cohen's standards, meaning the power creep is statistically detectable but modest in practical terms.

### 7. Robustness of Findings

| Check | Result |
|-------|--------|
| Robust Regression (Huber) | Coefficients differ < 5% from OLS |
| Gamma GLM | Consistent conclusions with Gaussian |
| LASSO | Generation survives regularization |
| Jonckheere-Terpstra | `r ifelse(jt_sig, "Confirms monotonic trend", "No monotonic trend")` |

### 8. Final Conclusions

1. **Power creep EXISTS and is statistically significant**: The Jonckheere-Terpstra test confirms a monotonic increasing trend in Attack across generations.

2. **LASSO confirms generation matters**: Generation effects survive automatic feature selection, meaning they provide predictive value beyond other characteristics.

3. **The effect is modest but real**: Cohen's d values are in the "small" range (0.3-0.4), translating to roughly 8-12 Attack points difference between early and late generations.

4. **Power creep is NOT uniform**:
   - Some types experience more creep than others
   - `r ifelse(normal_creep, "Normal Pokémon ARE affected (competitive implications)", "Creep is mainly in Legendary/Mythical (less competitive impact)")`

5. **Gen 7 stands out**: Generation 7 shows the strongest power creep effect among all generations.

6. **Findings are robust**: Multiple analytical approaches (GLM, LASSO, robust regression, J-T test) converge on the same conclusion.

---

# Session Info

```{r session-info}
sessionInfo()
```

