---
title: "Pokémon Power Creep Analysis"
subtitle: "Using Generalized Linear Models"
author: "Jan Tychtl"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: cosmo
    highlight: tango
    code_folding: show
---

# Introduction

## Main Research Question

**Is there Power Creep in the Pokémon Index?**

- Does generation significantly predict Pokémon strength (Attack)?
- How does attack change with generation after controlling for other factors?

## Methods Applied

Methods applied from SAN project reports analysis:

- OLS Regression, GLMs (Gaussian, Poisson, Gamma, Negative Binomial)
- LASSO/Ridge/ElasticNet regularization
- Backward Stepwise Selection
- ANOVA and Likelihood Ratio Tests
- Time-split Cross-Validation
- Spline functions and polynomial terms for non-linear effects
- Dispersion tests for overdispersion detection
- Comprehensive ablation study with caret cross-validation
- Diagnostic plots and model comparison (AIC/BIC)

*Inspired by: SAN_assignment_GLM_2.Rmd and glms.Rmd*

---

# 1. Setup and Data Loading

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(glmnet)       # LASSO, Ridge, ElasticNet
library(splines)      # Spline functions
library(car)          # ANOVA, VIF
library(MASS)         # stepAIC, robust regression, glm.nb (Negative Binomial)
library(broom)        # tidy model outputs
library(caret)        # Cross-validation with train()
library(ggplot2)      # Visualization
library(corrplot)     # Correlation plots
library(lmtest)       # Likelihood ratio tests
library(AER)          # dispersiontest() for overdispersion
library(rcompanion)   # compareGLM() for model comparison
library(gridExtra)    # Side-by-side plots
library(DescTools)    # Jonckheere-Terpstra test
library(moments)      # Skewness and kurtosis calculations
```

```{r load-data}
# Set working directory and load data
setwd("/home/jenda/Desktop/SAN/pokedex_analysis")
pokemon <- read.csv("data/data_clean.csv")
pokemon_clean <- read.csv("data/data_no_outliers.csv")

cat("Full dataset:", nrow(pokemon), "Pokémon\n")
cat("Outlier-free dataset:", nrow(pokemon_clean), "Pokémon\n")
cat("Removed as outliers:", nrow(pokemon) - nrow(pokemon_clean), "Pokémon\n")
```

**Note:** We load both the full dataset and the outlier-free version. A third dataset (`pokemon_normal`) containing only Normal-rarity Pokémon will be created after preprocessing to isolate the power creep analysis from legendary/mythical effects.

---

# 2. Data Exploration and Preprocessing

## 2.1 Data Overview

```{r data-overview}
cat("Dataset dimensions:", dim(pokemon), "\n")
cat("Number of generations:", length(unique(pokemon$gen)), "\n")
cat("Pokémon per generation:\n")
print(table(pokemon$gen))

# Check structure
str(pokemon)
summary(pokemon)
```

**Findings:** Dataset contains 1024 Pokémon across 9 generations. Generation sizes vary from 72 (Gen 6) to 156 (Gen 5), which may affect statistical power for certain comparisons.

## 2.2 Convert Categorical Variables to Factors

```{r convert-factors}
# Convert factors for FULL DATA
pokemon <- pokemon %>%
  mutate(
    gen = as.factor(gen),
    rarity = as.factor(rarity),
    class_primary = as.factor(class_primary),
    class_secondary = as.factor(class_secondary),
    has_mega_evolution = as.factor(has_mega_evolution),
    has_gigantamax = as.factor(has_gigantamax)
  )

# Define rarity levels (from Information.txt)
# 0 = normal, 1 = sublegendary, 2 = legendary, 3 = mythical
levels(pokemon$rarity) <- c("Normal", "SubLegendary", "Legendary", "Mythical")

# Convert factors for OUTLIER-FREE DATA (same transformations)
pokemon_clean <- pokemon_clean %>%
  mutate(
    gen = as.factor(gen),
    rarity = as.factor(rarity),
    class_primary = as.factor(class_primary),
    class_secondary = as.factor(class_secondary),
    has_mega_evolution = as.factor(has_mega_evolution),
    has_gigantamax = as.factor(has_gigantamax)
  )
levels(pokemon_clean$rarity) <- c("Normal", "SubLegendary", "Legendary", "Mythical")

cat("\nRarity distribution (Full Data):\n")
print(table(pokemon$rarity))
cat("\nRarity distribution (Outlier-Free):\n")
print(table(pokemon_clean$rarity))

# Create NORMAL-ONLY dataset (excluding legendaries/mythicals)
pokemon_normal <- pokemon %>% filter(rarity == "Normal")

cat("\n--- Dataset Sizes ---\n")
cat("Full dataset:", nrow(pokemon), "Pokémon\n")
cat("Outlier-free:", nrow(pokemon_clean), "Pokémon\n
")
cat("Normal-only:", nrow(pokemon_normal), "Pokémon (", 
    round(100 * nrow(pokemon_normal) / nrow(pokemon), 1), "% of full)\n")
```

**Findings:** Majority of Pokémon are Normal rarity. The `pokemon_normal` dataset excludes legendaries/mythicals to test if power creep exists among "regular" Pokémon.

## 2.3 Create Derived Features

```{r derived-features}
# Create derived features for FULL DATA
pokemon <- pokemon %>%
  mutate(
    # Total stats (overall power indicator)
    total_stats = hp + attack + defense + sp_attack + sp_defense + speed,
    
    # Physical vs Special orientation
    physical_power = attack + defense,
    special_power = sp_attack + sp_defense,
    offensive_power = attack + sp_attack,
    defensive_power = defense + sp_defense,
    
    # Is legendary (any form of rare)
    is_legendary = as.factor(ifelse(rarity != "Normal", 1, 0)),
    
    # BMI-like measure
    bmi = weight_kg / (height_m^2),
    
    # Log transformations for skewed variables
    log_weight = log1p(weight_kg),
    log_capture_rate = log1p(capture_rate),
    log_egg_steps = log1p(base_egg_steps),
    
    # Numeric generation for trend analysis
    gen_numeric = as.numeric(as.character(gen))
  )

# Create derived features for OUTLIER-FREE DATA (same transformations)
pokemon_clean <- pokemon_clean %>%
  mutate(
    total_stats = hp + attack + defense + sp_attack + sp_defense + speed,
    physical_power = attack + defense,
    special_power = sp_attack + sp_defense,
    offensive_power = attack + sp_attack,
    defensive_power = defense + sp_defense,
    is_legendary = as.factor(ifelse(rarity != "Normal", 1, 0)),
    bmi = weight_kg / (height_m^2),
    log_weight = log1p(weight_kg),
    log_capture_rate = log1p(capture_rate),
    log_egg_steps = log1p(base_egg_steps),
    gen_numeric = as.numeric(as.character(gen))
  )

# Create derived features for NORMAL-ONLY DATA
# Note: pokemon_normal was filtered from pokemon after factor conversion,
# so it already has rarity as factor. We just need derived features.
pokemon_normal <- pokemon_normal %>%
  mutate(
    total_stats = hp + attack + defense + sp_attack + sp_defense + speed,
    physical_power = attack + defense,
    special_power = sp_attack + sp_defense,
    offensive_power = attack + sp_attack,
    defensive_power = defense + sp_defense,
    is_legendary = as.factor(ifelse(rarity != "Normal", 1, 0)),
    bmi = weight_kg / (height_m^2),
    log_weight = log1p(weight_kg),
    log_capture_rate = log1p(capture_rate),
    log_egg_steps = log1p(base_egg_steps),
    gen_numeric = as.numeric(as.character(gen))
  )

cat("--- All Datasets Ready ---\n")
cat("pokemon:", nrow(pokemon), "rows,", ncol(pokemon), "cols\n")
cat("pokemon_clean:", nrow(pokemon_clean), "rows,", ncol(pokemon_clean), "cols\n")
cat("pokemon_normal:", nrow(pokemon_normal), "rows,", ncol(pokemon_normal), "cols\n")
```

---

# 3. Exploratory Data Analysis (EDA)

## 3.1 Attack by Generation

```{r eda-summary}
# Summary statistics of attack by generation - FULL DATA
attack_summary <- pokemon %>%
  group_by(gen) %>%
  summarise(
    n = n(),
    mean_attack = mean(attack, na.rm = TRUE),
    sd_attack = sd(attack, na.rm = TRUE),
    median_attack = median(attack, na.rm = TRUE),
    min_attack = min(attack, na.rm = TRUE),
    max_attack = max(attack, na.rm = TRUE)
  )
#cat("--- Full Dataset ---\n")
print(attack_summary)

# Summary statistics - OUTLIER-FREE DATA
attack_summary_clean <- pokemon_clean %>%
  group_by(gen) %>%
  summarise(
    n = n(),
    mean_attack = mean(attack, na.rm = TRUE),
    sd_attack = sd(attack, na.rm = TRUE),
    median_attack = median(attack, na.rm = TRUE),
    min_attack = min(attack, na.rm = TRUE),
    max_attack = max(attack, na.rm = TRUE)
  )
cat("\n--- Outlier-Free Dataset ---\n")
print(attack_summary_clean)

# Summary statistics - NORMAL RARITY ONLY
attack_summary_normal <- pokemon_normal %>%
  group_by(gen) %>%
  summarise(
    n = n(),
    mean_attack = mean(attack, na.rm = TRUE),
    sd_attack = sd(attack, na.rm = TRUE),
    median_attack = median(attack, na.rm = TRUE),
    min_attack = min(attack, na.rm = TRUE),
    max_attack = max(attack, na.rm = TRUE)
  )
cat("\n--- Normal Rarity Only ---\n")
print(attack_summary_normal)

# Quick comparison of all three datasets
cat("\n--- Mean Attack Comparison (All Datasets) ---\n")
comparison_eda <- data.frame(
  gen = 1:9,
  full = attack_summary$mean_attack,
  clean = attack_summary_clean$mean_attack,
  normal = attack_summary_normal$mean_attack
) %>% mutate(
  diff_clean = round(full - clean, 1),
  diff_normal = round(full - normal, 1)
)
print(comparison_eda)
```

**Findings:** Mean attack ranges from 68.3 (Gen 2) to 84.8 (Gen 7) in full data. The pattern is NOT monotonic — Gen 2 and Gen 6 dip below Gen 1. Normal-only means follow a similar pattern (66.3 to 81.4), confirming power creep isn't solely driven by legendaries.

```{r eda-boxplot, fig.width=15, fig.height=5}
# Plot: Attack distribution by generation (boxplot) - THREE-WAY COMPARISON
p1_full <- ggplot(pokemon, aes(x = gen, y = attack, fill = gen)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.2, size = 0.8) +
  labs(title = "Full Dataset", x = "Generation", y = "Attack") +
  theme_minimal() +
  theme(legend.position = "none") +
  ylim(0, 200)

p1_clean <- ggplot(pokemon_clean, aes(x = gen, y = attack, fill = gen)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.2, size = 0.8) +
  labs(title = "Outlier-Free", x = "Generation", y = "Attack") +
  theme_minimal() +
  theme(legend.position = "none") +
  ylim(0, 200)

p1_normal <- ggplot(pokemon_normal, aes(x = gen, y = attack, fill = gen)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.2, size = 0.8) +
  labs(title = "Normal Rarity Only", x = "Generation", y = "Attack") +
  theme_minimal() +
  theme(legend.position = "none") +
  ylim(0, 200)

gridExtra::grid.arrange(p1_full, p1_clean, p1_normal, ncol = 3,
                        top = "Attack Distribution by Generation: Full vs Outlier-Free vs Normal-Only")
```

**Findings:** Visual inspection shows later generations (4, 5, 7, 8, 9) have higher median attack. The Normal-only plot removes legendary influence — if the pattern persists, power creep affects regular Pokémon too.

```{r eda-total-stats, fig.width=10, fig.height=6}
# Plot: Total stats by generation
p2_total <- ggplot(pokemon, aes(x = gen, y = total_stats, fill = gen)) +
  geom_boxplot(alpha = 0.7) +
  labs(title = "Total Stats Distribution by Generation",
       x = "Generation", y = "Total Stats") +
  theme_minimal() +
  theme(legend.position = "none")

p1_clean_total <- ggplot(pokemon_clean, aes(x = gen, y = total_stats, fill = gen)) +
  geom_boxplot(alpha = 0.7) +
  labs(title = "Total Stats Distribution (Clean) by Generation",
       x = "Generation", y = "Total Stats") +
  theme_minimal() +
  theme(legend.position = "none")

p3_normal_total <- ggplot(pokemon_normal, aes(x = gen, y = total_stats, fill = gen)) +
  geom_boxplot(alpha = 0.7) +
  labs(title = "Total Stats Distribution (Normal) by Generation",
       x = "Generation", y = "Total Stats") +
  theme_minimal() +
  theme(legend.position = "none")

gridExtra::grid.arrange( p2_total, p1_clean_total, p3_normal_total,ncol = 3,
                        top = "Attack Distribution by Generation: Full vs Outlier-Free")
```

**Findings:** So far it seems that there is a slight upwards trend in the generations. In all versions.
We ca also observe small interesting thing, in the normal dataset we can see that the sum of all the stats seems to have a set celling for stats sum for majority of the generations except 3, 8, 9.

## 3.2 Correlation Matrix

```{r correlation-matrix, fig.width=8, fig.height=8}
# Correlation matrix for numeric variables
numeric_vars <- pokemon %>%
  dplyr::select(attack, hp, defense, sp_attack, sp_defense, speed, 
                height_m, weight_kg, capture_rate, num_abilities, evo_length)

cor_matrix <- cor(numeric_vars, use = "complete.obs")
cat("\nCorrelation matrix (combat stats):\n")
print(round(cor_matrix, 2))

corrplot(cor_matrix, method = "color", type = "upper", 
         addCoef.col = "black", tl.col = "black", tl.srt = 45,
         title = "Correlation Matrix: Combat Stats & Features")
```

**Findings:** Attack correlates moderately with HP (0.42), Defense (0.44), and Speed (0.38). These correlations justify including them as controls — they explain shared variance but don't create severe multicollinearity (all < 0.7).

### 3.2.1 Generation-Specific Correlation Analysis

Do the correlations between combat stats change across generations? If multicollinearity patterns shift, it could indicate changes in Pokémon design philosophy over time.

```{r gen-specific-correlations, fig.width=14, fig.height=12}
# Calculate correlation matrices for each generation
combat_vars <- c("attack", "hp", "defense", "sp_attack", "sp_defense", "speed")

# Function to calculate correlation matrix for a generation
calc_gen_cor <- function(data, gen_num) {
  gen_data <- data %>% filter(gen == gen_num) %>% dplyr::select(all_of(combat_vars))
  cor(gen_data, use = "complete.obs")
}

# Calculate correlations for all generations
gen_cors <- lapply(1:9, function(g) calc_gen_cor(pokemon, g))
names(gen_cors) <- paste0("Gen", 1:9)

# Extract key correlations (attack vs other stats) for each generation
attack_cors <- data.frame(
  Generation = 1:9,
  attack_hp = sapply(gen_cors, function(m) m["attack", "hp"]),
  attack_defense = sapply(gen_cors, function(m) m["attack", "defense"]),
  attack_sp_attack = sapply(gen_cors, function(m) m["attack", "sp_attack"]),
  attack_sp_defense = sapply(gen_cors, function(m) m["attack", "sp_defense"]),
  attack_speed = sapply(gen_cors, function(m) m["attack", "speed"])
)

cat("--- Attack Correlations by Generation ---\n")
print(round(attack_cors, 3))

# Visualize correlation matrices by generation (3x3 grid)
par(mfrow = c(3, 3), mar = c(2, 2, 3, 1))
for(g in 1:9) {
  corrplot(gen_cors[[g]], method = "color", type = "upper",
           addCoef.col = "black", tl.col = "black", tl.srt = 45,
           number.cex = 0.6, tl.cex = 0.7,
           title = paste("Gen", g, "(n =", sum(pokemon$gen == g), ")"),
           mar = c(0, 0, 2, 0))
}
```

```{r gen-correlation-trends, fig.width=12, fig.height=6}
# Plot how attack correlations change over generations
attack_cors_long <- attack_cors %>%
  pivot_longer(cols = -Generation, names_to = "correlation", values_to = "value") %>%
  mutate(correlation = gsub("attack_", "Attack vs ", correlation))

p_cor_trends <- ggplot(attack_cors_long, aes(x = Generation, y = value, color = correlation)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  geom_hline(yintercept = 0.7, linetype = "dashed", color = "red", alpha = 0.5) +
  geom_hline(yintercept = -0.7, linetype = "dashed", color = "red", alpha = 0.5) +
  annotate("text", x = 1.5, y = 0.72, label = "High collinearity threshold", 
           color = "red", size = 3, alpha = 0.7) +
  scale_x_continuous(breaks = 1:9) +
  labs(title = "How Attack Correlations Change Across Generations",
       subtitle = "Dashed lines = ±0.7 (high collinearity threshold)",
       x = "Generation", y = "Correlation with Attack",
       color = "Stat Pair") +
  theme_minimal() +
  theme(legend.position = "bottom")
print(p_cor_trends)
```

```{r gen-multicollinearity-summary}
# Check for any generation with problematic multicollinearity (|cor| > 0.7)
cat("\n--- Multicollinearity Check by Generation ---\n")
cat("Checking for correlations |r| > 0.7 (high collinearity)\n\n")

high_cor_summary <- data.frame()
for(g in 1:9) {
  cor_mat <- gen_cors[[g]]
  # Get upper triangle (excluding diagonal)
  cor_mat[lower.tri(cor_mat, diag = TRUE)] <- NA
  high_cors <- which(abs(cor_mat) > 0.7, arr.ind = TRUE)
  
  if(nrow(high_cors) > 0) {
    for(i in 1:nrow(high_cors)) {
      row_name <- rownames(cor_mat)[high_cors[i, 1]]
      col_name <- colnames(cor_mat)[high_cors[i, 2]]
      cor_val <- cor_mat[high_cors[i, 1], high_cors[i, 2]]
      high_cor_summary <- rbind(high_cor_summary, 
                                 data.frame(Generation = g, 
                                            Var1 = row_name, 
                                            Var2 = col_name,
                                            Correlation = round(cor_val, 3)))
    }
  }
}

if(nrow(high_cor_summary) > 0) {
  cat("Generations with high multicollinearity (|r| > 0.7):\n")
  print(high_cor_summary)
} else {
  cat("No generation has correlations exceeding |r| > 0.7\n")
}

# Summary statistics
cat("\n--- Correlation Stability Summary ---\n")
cor_ranges <- attack_cors %>%
  summarise(
    hp_range = paste0("[", round(min(attack_hp), 2), ", ", round(max(attack_hp), 2), "]"),
    defense_range = paste0("[", round(min(attack_defense), 2), ", ", round(max(attack_defense), 2), "]"),
    sp_attack_range = paste0("[", round(min(attack_sp_attack), 2), ", ", round(max(attack_sp_attack), 2), "]"),
    sp_defense_range = paste0("[", round(min(attack_sp_defense), 2), ", ", round(max(attack_sp_defense), 2), "]"),
    speed_range = paste0("[", round(min(attack_speed), 2), ", ", round(max(attack_speed), 2), "]")
  )
cat("Range of attack correlations across generations:\n")
cat("  Attack vs HP:", cor_ranges$hp_range, "\n")
cat("  Attack vs Defense:", cor_ranges$defense_range, "\n")
cat("  Attack vs Sp.Attack:", cor_ranges$sp_attack_range, "\n")
cat("  Attack vs Sp.Defense:", cor_ranges$sp_defense_range, "\n")
cat("  Attack vs Speed:", cor_ranges$speed_range, "\n")
```

**Findings:** The generation-specific correlation analysis reveals:

1. **Correlation stability:** The correlation structure between combat stats is relatively stable across generations. No generation has dramatically different multicollinearity patterns.

2. **No severe multicollinearity:** No generation exhibits correlations exceeding |r| > 0.7 between predictors, confirming that multicollinearity is not a generation-specific problem.

3. **Notable patterns:** 
   - Attack-Defense correlation tends to be positive (~0.3–0.5) across all generations
   - Attack-Speed correlation varies more (some generations have near-zero correlation)
   - Sp.Attack and Sp.Defense show the most variable relationship with Attack across generations

4. **Implication for modeling:** The consistent correlation structure justifies using a single pooled model rather than generation-specific models. The predictor relationships are stable enough that coefficients estimated from pooled data should be reliable.

## 3.3 Attack Distribution per Generation

```{r attack-distribution, fig.width=12, fig.height=8}
library(ggplot2)
library(ggridges)
library(dplyr)

p_ridge <- pokemon %>%
  ggplot(aes(x = attack, y = as.factor(gen), fill = as.factor(gen))) +
  geom_density_ridges(alpha = 0.7, color = "white", scale = 1.2) +
  labs(
    title = "Attack Distribution per Generation",
    subtitle = "Ridge plot view for comparing distribution shapes",
    x = "Attack",
    y = "Generation",
    fill = "Generation"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

print(p_ridge)
```

```{r attack_normal-distribution, fig.width=12, fig.height=8}
p_dist <- pokemon_normal %>%
  ggplot(aes(x = attack)) +
  geom_histogram(bins = 20, fill = "coral", color = "white", alpha = 0.7) +
  facet_wrap(~gen, ncol = 3) +
  labs(title = "Attack Distribution per Generation (Normal only)",
       subtitle = "Checking if distribution shape changes across generations",
       x = "Attack", y = "Count") +
  theme_minimal()
print(p_dist)
```

```{r total_stats-distribution, fig.width=12, fig.height=8}
p_dist <- pokemon %>%
  ggplot(aes(x = total_stats)) +
  geom_histogram(bins = 20, fill = "steelblue", color = "white", alpha = 0.7) +
  facet_wrap(~gen, ncol = 3) +
  labs(title = "Total stats Distribution per Generation",
       subtitle = "Checking if distribution shape changes across generations",
       x = "Attack", y = "Count") +
  theme_minimal()
print(p_dist)
```

We can see slight shift in the total stats through generations. It was normal distribution like at the first generation, then it gets more skewed to heavy end tails.

## 3.4 Distribution of All Combat Stats (GLM Prerequisites)

Before applying GLMs, we need to verify that our response variables are approximately normally distributed. This section examines the distribution of all six combat stats.

```{r all-stats-distributions, fig.width=14, fig.height=10}
# Reshape data to long format for faceted plotting
combat_stats_long <- pokemon %>%
  dplyr::select(attack, hp, defense, sp_attack, sp_defense, speed) %>%
  pivot_longer(cols = everything(), names_to = "stat", values_to = "value") %>%
  mutate(stat = factor(stat, 
                       levels = c("attack", "hp", "defense", "sp_attack", "sp_defense", "speed"),
                       labels = c("Attack", "HP", "Defense", "Sp. Attack", "Sp. Defense", "Speed")))

# Plot distributions with density overlay
p_all_stats <- ggplot(combat_stats_long, aes(x = value)) +
  geom_histogram(aes(y = after_stat(density)), bins = 30, 
                 fill = "steelblue", color = "white", alpha = 0.7) +
  geom_density(color = "darkred", linewidth = 1) +
  facet_wrap(~ stat, scales = "free", ncol = 3) +
  labs(title = "Distribution of All Combat Stats",
       subtitle = "Histograms with density overlay — checking for normality",
       x = "Stat Value", y = "Density") +
  theme_minimal() +
  theme(strip.text = element_text(face = "bold", size = 11))
print(p_all_stats)
```

```{r stats-summary-table}
# Comprehensive summary statistics for all combat stats
stats_summary <- pokemon %>%
  dplyr::select(attack, hp, defense, sp_attack, sp_defense, speed, total_stats) %>%
  pivot_longer(cols = everything(), names_to = "stat", values_to = "value") %>%
  group_by(stat) %>%
  summarise(
    n = n(),
    mean = round(mean(value), 1),
    median = round(median(value), 1),
    sd = round(sd(value), 1),
    min = min(value),
    max = max(value),
    skewness = round(moments::skewness(value), 2),
    kurtosis = round(moments::kurtosis(value), 2),
    .groups = "drop"
  ) %>%
  mutate(stat = factor(stat, 
                       levels = c("attack", "hp", "defense", "sp_attack", "sp_defense", "speed", "total_stats"),
                       labels = c("Attack", "HP", "Defense", "Sp.Attack", "Sp.Defense", "Speed", "Total Stats"))) %>%
  arrange(stat)

cat("--- Summary Statistics for All Combat Stats ---\n")
print(stats_summary)

# Interpretation
cat("\n--- Normality Assessment ---\n")
cat("Skewness interpretation: |skew| < 0.5 = approximately symmetric\n")
cat("                         |skew| 0.5-1 = moderately skewed\n")
cat("                         |skew| > 1   = highly skewed\n\n")

for(i in 1:nrow(stats_summary)) {
  skew <- abs(stats_summary$skewness[i])
  assessment <- ifelse(skew < 0.5, "approximately symmetric", 
                       ifelse(skew < 1, "moderately skewed", "highly skewed"))
  cat(paste0(stats_summary$stat[i], ": skewness = ", stats_summary$skewness[i], 
             " (", assessment, ")\n"))
}
```

```{r qq-plots-all-stats, fig.width=14, fig.height=10}
# Q-Q plots to assess normality
p_qq <- ggplot(combat_stats_long, aes(sample = value)) +
  stat_qq(color = "steelblue", alpha = 0.5) +
  stat_qq_line(color = "darkred", linewidth = 1) +
  facet_wrap(~ stat, scales = "free", ncol = 3) +
  labs(title = "Q-Q Plots for All Combat Stats",
       subtitle = "Points following the red line indicate normal distribution",
       x = "Theoretical Quantiles", y = "Sample Quantiles") +
  theme_minimal() +
  theme(strip.text = element_text(face = "bold", size = 11))
print(p_qq)
```

```{r shapiro-tests}
# Formal normality tests (Shapiro-Wilk)
# Note: With large samples, even small deviations become significant
cat("\n--- Shapiro-Wilk Normality Tests ---\n")
cat("Note: With n = 1024, even minor deviations from normality become detectable.\n")
cat("      Focus on visual assessment (Q-Q plots) for practical evaluation.\n\n")

# Test each stat (use sample of 5000 max due to Shapiro-Wilk limit)
set.seed(42)
for(stat_name in c("attack", "hp", "defense", "sp_attack", "sp_defense", "speed")) {
  values <- pokemon[[stat_name]]
  if(length(values) > 5000) values <- sample(values, 5000)
  test_result <- shapiro.test(values)
  cat(paste0(stat_name, ": W = ", round(test_result$statistic, 4), 
             ", p = ", format(test_result$p.value, digits = 3), "\n"))
}
```

**Findings for GLM Prerequisites:**

All combat stats show:

- **Slight positive skewness** (0.35 to 0.68) — right tails indicating some high-stat Pokémon
- **Approximately symmetric** distributions around the median
- **No extreme departures** from normality in Q-Q plots
- **Kurtosis** close to 3 (normal distribution has kurtosis = 3)

This supports using **Gaussian GLM with identity link** as our primary specification. The slight right skewness justifies using **Gamma GLM** as a robustness check.

## 3.5 Variance Analysis Across Generations (Homoscedasticity Check)

For Gaussian GLMs, we assume constant variance (homoscedasticity). We need to verify that the variance of each combat stat does not systematically change across generations.

```{r variance-by-generation, fig.width=14, fig.height=10}
# Calculate mean and SD for each stat by generation
variance_summary <- pokemon %>%
  group_by(gen) %>%
  summarise(
    n = n(),
    attack_mean = mean(attack), attack_sd = sd(attack),
    hp_mean = mean(hp), hp_sd = sd(hp),
    defense_mean = mean(defense), defense_sd = sd(defense),
    sp_attack_mean = mean(sp_attack), sp_attack_sd = sd(sp_attack),
    sp_defense_mean = mean(sp_defense), sp_defense_sd = sd(sp_defense),
    speed_mean = mean(speed), speed_sd = sd(speed),
    .groups = "drop"
  )

cat("--- Standard Deviations by Generation (All Combat Stats) ---\n\n")
sd_table <- variance_summary %>%
  dplyr::select(gen, n, attack_sd, hp_sd, defense_sd, sp_attack_sd, sp_defense_sd, speed_sd) %>%
  mutate(across(ends_with("_sd"), ~ round(.x, 1)))
print(sd_table)

# Reshape for plotting
variance_long <- variance_summary %>%
  dplyr::select(gen, ends_with("_sd")) %>%
  pivot_longer(cols = -gen, names_to = "stat", values_to = "sd") %>%
  mutate(stat = gsub("_sd", "", stat),
         stat = factor(stat, 
                       levels = c("attack", "hp", "defense", "sp_attack", "sp_defense", "speed"),
                       labels = c("Attack", "HP", "Defense", "Sp. Attack", "Sp. Defense", "Speed")))

# Plot SD by generation for each stat
p_variance <- ggplot(variance_long, aes(x = as.numeric(as.character(gen)), y = sd)) +
  geom_point(size = 3, color = "steelblue") +
  geom_line(color = "steelblue", alpha = 0.7) +
  geom_hline(aes(yintercept = mean(sd)), linetype = "dashed", color = "darkred") +
  facet_wrap(~ stat, scales = "free_y", ncol = 3) +
  scale_x_continuous(breaks = 1:9) +
  labs(title = "Standard Deviation by Generation for All Combat Stats",
       subtitle = "Dashed red line = overall mean SD. Stable lines indicate homoscedasticity.",
       x = "Generation", y = "Standard Deviation") +
  theme_minimal() +
  theme(strip.text = element_text(face = "bold", size = 11))
print(p_variance)
```

```{r variance-vs-mean-plot, fig.width=14, fig.height=10}
# Plot SD vs Mean to check if variance increases with mean
mean_long <- variance_summary %>%
  dplyr::select(gen, ends_with("_mean")) %>%
  pivot_longer(cols = -gen, names_to = "stat", values_to = "mean") %>%
  mutate(stat = gsub("_mean", "", stat))

variance_vs_mean <- variance_long %>%
  mutate(stat_raw = tolower(gsub(" ", "_", gsub("\\. ", "_", stat)))) %>%
  left_join(mean_long %>% rename(stat_raw = stat), by = c("gen", "stat_raw"))

p_var_vs_mean <- ggplot(variance_vs_mean, aes(x = mean, y = sd)) +
  geom_point(aes(color = factor(gen)), size = 3) +
  geom_smooth(method = "lm", se = FALSE, color = "darkred", linetype = "dashed") +
  facet_wrap(~ stat, scales = "free", ncol = 3) +
  labs(title = "Variance vs Mean: Checking for Heteroscedasticity",
       subtitle = "Flat trend = constant variance (Gaussian OK). Positive slope = variance increases with mean (consider Gamma)",
       x = "Mean", y = "Standard Deviation", color = "Generation") +
  theme_minimal() +
  theme(strip.text = element_text(face = "bold", size = 11),
        legend.position = "bottom")
print(p_var_vs_mean)
```

```{r simple-model-diagnostics, fig.width=14, fig.height=5}
# For simple model (attack ~ gen), gen is a factor so fitted values are discrete
# Standard residual plots show vertical lines - not useful for visual inspection
# Instead, we examine residual distribution BY GENERATION using boxplots

model_simple <- lm(attack ~ gen, data = pokemon)

cat("--- Simple Model: Attack ~ Generation ---\n")
cat("Note: Since generation is a categorical factor, we use boxplots\n")
cat("      to visualize residual spread within each generation.\n\n")

# Add residuals to data
pokemon_resid <- pokemon %>%
  mutate(residuals = residuals(model_simple),
         fitted = fitted(model_simple))

# Boxplot of residuals by generation (better for factor predictor)
p_resid_box <- ggplot(pokemon_resid, aes(x = gen, y = residuals, fill = gen)) +
  geom_boxplot(alpha = 0.7) +
  geom_hline(yintercept = 0, color = "darkred", linetype = "dashed", linewidth = 1) +
  labs(title = "Residuals by Generation (Simple Model: Attack ~ Gen)",
       subtitle = "Similar spread across generations = homoscedasticity",
       x = "Generation", y = "Residuals") +
  theme_minimal() +
  theme(legend.position = "none")

# SD of residuals by generation
resid_sd <- pokemon_resid %>%
  group_by(gen) %>%
  summarise(n = n(), sd_residuals = round(sd(residuals), 2), .groups = "drop")

p_resid_sd <- ggplot(resid_sd, aes(x = gen, y = sd_residuals, fill = gen)) +
  geom_col(alpha = 0.7) +
  geom_hline(yintercept = mean(resid_sd$sd_residuals), color = "darkred", 
             linetype = "dashed", linewidth = 1) +
  geom_text(aes(label = sd_residuals), vjust = -0.3, size = 3) +
  labs(title = "Standard Deviation of Residuals by Generation",
       subtitle = "Dashed line = overall mean SD. Similar bars = constant variance.",
       x = "Generation", y = "SD of Residuals") +
  theme_minimal() +
  theme(legend.position = "none")

gridExtra::grid.arrange(p_resid_box, p_resid_sd, ncol = 2)

cat("Residual SD by generation:\n")
print(resid_sd)
```

```{r simple-model-qqplot, fig.width=8, fig.height=6}
# Q-Q plot for normality of residuals (simple model)
p_qq_simple <- ggplot(pokemon_resid, aes(sample = residuals)) +
  stat_qq(color = "steelblue", alpha = 0.5) +
  stat_qq_line(color = "darkred", linewidth = 1) +
  labs(title = "Q-Q Plot: Residuals from Simple Model (Attack ~ Gen)",
       subtitle = "Points following line = normal residuals",
       x = "Theoretical Quantiles", y = "Sample Quantiles") +
  theme_minimal()
print(p_qq_simple)
```

```{r breusch-pagan-simple}
# Breusch-Pagan tests on SIMPLE models (stat ~ gen only)
cat("\n--- Breusch-Pagan Tests: SIMPLE Models (stat ~ gen) ---\n")
cat("H₀: Constant variance (homoscedasticity)\n")
cat("H₁: Variance changes with fitted values (heteroscedasticity)\n\n")

bp_simple <- data.frame(
  Stat = character(),
  BP_statistic = numeric(),
  df = numeric(),
  p_value = numeric(),
  Conclusion = character(),
  stringsAsFactors = FALSE
)

for(stat_name in c("attack", "hp", "defense", "sp_attack", "sp_defense", "speed")) {
  formula_str <- as.formula(paste(stat_name, "~ gen"))
  model <- lm(formula_str, data = pokemon)
  bp_test <- bptest(model)
  
  conclusion <- ifelse(bp_test$p.value < 0.05, 
                       "Heteroscedastic", "Homoscedastic")
  
  bp_simple <- rbind(bp_simple, data.frame(
    Stat = stat_name,
    BP_statistic = round(bp_test$statistic, 3),
    df = bp_test$parameter,
    p_value = round(bp_test$p.value, 4),
    Conclusion = conclusion
  ))
}

print(bp_simple)
```

```{r full-model-diagnostics, fig.width=12, fig.height=10}
# Now test on FULL model with all controls (more realistic)
# With continuous predictors (hp, defense, speed, etc.), fitted values have proper spread
# so standard diagnostic plots are now meaningful

cat("\n--- Standard Diagnostic Plots: FULL Model ---\n")
cat("Model: attack ~ gen + hp + defense + sp_attack + sp_defense + speed + rarity\n")
cat("With continuous controls, fitted values have proper spread for diagnostics.\n\n")

model_full_diag <- lm(attack ~ gen + hp + defense + sp_attack + sp_defense + speed + rarity, 
                      data = pokemon)

par(mfrow = c(2, 2))
plot(model_full_diag)

# Summary of model
cat("\n--- Full Model R-squared ---\n")
cat("R² =", round(summary(model_full_diag)$r.squared, 4), "\n")
cat("Adjusted R² =", round(summary(model_full_diag)$adj.r.squared, 4), "\n")
```

```{r breusch-pagan-full}
# Breusch-Pagan tests on FULL models (stat ~ gen + controls)
cat("\n--- Breusch-Pagan Tests: FULL Models (with controls) ---\n")
cat("This is more relevant since we will use full models in our analysis.\n\n")

bp_full <- data.frame(
  Stat = character(),
  Model = character(),
  BP_statistic = numeric(),
  df = numeric(),
  p_value = numeric(),
  Conclusion = character(),
  stringsAsFactors = FALSE
)

# Define control variables (excluding the target stat itself)
for(stat_name in c("attack", "hp", "defense", "sp_attack", "sp_defense", "speed")) {
  # Get other stats as controls
  other_stats <- setdiff(c("hp", "defense", "sp_attack", "sp_defense", "speed"), stat_name)
  controls <- paste(c("gen", other_stats, "rarity"), collapse = " + ")
  
  formula_str <- as.formula(paste(stat_name, "~", controls))
  model <- lm(formula_str, data = pokemon)
  bp_test <- bptest(model)
  
  conclusion <- ifelse(bp_test$p.value < 0.05, 
                       "Heteroscedastic", "Homoscedastic")
  
  bp_full <- rbind(bp_full, data.frame(
    Stat = stat_name,
    Model = "Full",
    BP_statistic = round(bp_test$statistic, 3),
    df = bp_test$parameter,
    p_value = round(bp_test$p.value, 4),
    Conclusion = conclusion
  ))
}

print(bp_full)

# Summary comparison
cat("\n--- Summary: Simple vs Full Model Heteroscedasticity ---\n")
comparison <- data.frame(
  Stat = bp_simple$Stat,
  Simple_p = bp_simple$p_value,
  Full_p = bp_full$p_value,
  Simple_Result = bp_simple$Conclusion,
  Full_Result = bp_full$Conclusion
)
print(comparison)

cat("\nSimple models homoscedastic:", sum(bp_simple$p_value >= 0.05), "out of 6\n")
cat("Full models homoscedastic:", sum(bp_full$p_value >= 0.05), "out of 6\n")
```

```{r residual-vs-fitted-all, fig.width=14, fig.height=10}
# Residual vs Fitted plots for FULL models (more relevant)
cat("\n--- Residual Plots for Full Models ---\n")

residual_plots <- list()

for(stat_name in c("attack", "hp", "defense", "sp_attack", "sp_defense", "speed")) {
  other_stats <- setdiff(c("hp", "defense", "sp_attack", "sp_defense", "speed"), stat_name)
  controls <- paste(c("gen", other_stats, "rarity"), collapse = " + ")
  
  formula_str <- as.formula(paste(stat_name, "~", controls))
  model <- lm(formula_str, data = pokemon)
  
  plot_data <- data.frame(
    fitted = fitted(model),
    residuals = residuals(model),
    std_residuals = rstandard(model),
    stat = stat_name
  )
  
  residual_plots[[stat_name]] <- plot_data
}

residual_data <- bind_rows(residual_plots) %>%
  mutate(stat = factor(stat, 
                       levels = c("attack", "hp", "defense", "sp_attack", "sp_defense", "speed"),
                       labels = c("Attack", "HP", "Defense", "Sp. Attack", "Sp. Defense", "Speed")))

# Residuals vs Fitted
p_residuals <- ggplot(residual_data, aes(x = fitted, y = residuals)) +
  geom_point(alpha = 0.3, color = "steelblue") +
  geom_hline(yintercept = 0, color = "darkred", linetype = "dashed") +
  geom_smooth(method = "loess", se = FALSE, color = "orange", linewidth = 1) +
  facet_wrap(~ stat, scales = "free", ncol = 3) +
  labs(title = "Residuals vs Fitted Values (Full Models with Controls)",
       subtitle = "Horizontal spread = homoscedasticity. Funnel shape = heteroscedasticity.",
       x = "Fitted Values", y = "Residuals") +
  theme_minimal() +
  theme(strip.text = element_text(face = "bold", size = 11))
print(p_residuals)
```

```{r scale-location-plot, fig.width=14, fig.height=10}
# Scale-Location plots (sqrt of standardized residuals vs fitted)
# This is the standard diagnostic for heteroscedasticity
p_scale_location <- ggplot(residual_data, aes(x = fitted, y = sqrt(abs(std_residuals)))) +
  geom_point(alpha = 0.3, color = "steelblue") +
  geom_smooth(method = "loess", se = FALSE, color = "orange", linewidth = 1) +
  facet_wrap(~ stat, scales = "free", ncol = 3) +
  labs(title = "Scale-Location Plot (Full Models)",
       subtitle = "Flat line = constant variance. Upward slope = variance increases with fitted values.",
       x = "Fitted Values", y = expression(sqrt("|Standardized Residuals|"))) +
  theme_minimal() +
  theme(strip.text = element_text(face = "bold", size = 11))
print(p_scale_location)
```

**Findings for Variance Analysis:**

- **Visual inspection** of SD by generation shows relatively stable variance across generations for all stats
- **SD vs Mean plots** show no strong positive relationship — variance does not systematically increase with mean
- **Breusch-Pagan tests** for most stats show p > 0.05, supporting the homoscedasticity assumption
- **Residual plots** show roughly horizontal spread without obvious funnel patterns

These findings support using **Gaussian GLM** as our primary specification. The constant variance assumption is reasonably satisfied for all combat stats.

## 3.6 Rarity Analysis

```{r rarity-analysis}
rarity_summary <- pokemon %>%
  group_by(rarity) %>%
  summarise(
    n = n(),
    mean_attack = mean(attack),
    sd_attack = sd(attack),
    mean_total_stats = mean(total_stats)
  )
print(rarity_summary)
```

```{r rarity-summary-plot, fig.width=12, fig.height=5}
# Plot rarity summary: Attack and Total Stats side by side
rarity_long <- rarity_summary %>%
  pivot_longer(cols = c(mean_attack, mean_total_stats), 
               names_to = "stat", values_to = "value") %>%
  mutate(stat = ifelse(stat == "mean_attack", "Mean Attack", "Mean Total Stats"))

p_rarity_summary <- ggplot(rarity_long, aes(x = rarity, y = value, fill = rarity)) +
  geom_col(alpha = 0.8) +
  geom_text(aes(label = round(value, 1)), vjust = -0.3, size = 3.5) +
  facet_wrap(~ stat, scales = "free_y") +
  labs(title = "Mean Attack and Total Stats by Rarity",
       subtitle = "Comparing stat levels across rarity categories",
       x = "Rarity", y = "Value") +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 15, hjust = 1))
print(p_rarity_summary)
```

```{r gen-total-stats-summary}
# Mean total stats across each generation
gen_stats_summary <- pokemon %>%
  group_by(gen) %>%
  summarise(
    n = n(),
    mean_attack = mean(attack),
    mean_total_stats = mean(total_stats),
    sd_total_stats = sd(total_stats),
    median_total_stats = median(total_stats)
  )

cat("--- Mean Total Stats by Generation ---\n")
print(gen_stats_summary)
```

```{r gen-total-stats-plot, fig.width=12, fig.height=5}
# Visualize mean total stats and mean attack by generation
gen_stats_long <- gen_stats_summary %>%
  pivot_longer(cols = c(mean_attack, mean_total_stats), 
               names_to = "stat", values_to = "value") %>%
  mutate(stat = ifelse(stat == "mean_attack", "Mean Attack", "Mean Total Stats"))

p_gen_stats <- ggplot(gen_stats_long, aes(x = as.numeric(as.character(gen)), y = value, color = stat)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  geom_text(aes(label = round(value, 1)), vjust = -0.8, size = 3, show.legend = FALSE) +
  scale_x_continuous(breaks = 1:9) +
  labs(title = "Mean Attack and Total Stats by Generation",
       subtitle = "Tracking overall power trends across generations",
       x = "Generation", y = "Mean Value", color = "Statistic") +
  theme_minimal() +
  theme(legend.position = "bottom")
print(p_gen_stats)

# Calculate trend correlations
cor_attack <- cor(1:9, gen_stats_summary$mean_attack)
cor_total <- cor(1:9, gen_stats_summary$mean_total_stats)
cat("\n--- Trend Correlations ---\n")
cat("Correlation (gen vs mean_attack):", round(cor_attack, 3), "\n")
cat("Correlation (gen vs mean_total_stats):", round(cor_total, 3), "\n")
```

```{r total-stats-column-chart, fig.width=10, fig.height=6}
# Column chart: Mean Total Stats by Generation
p_total_stats_col <- ggplot(gen_stats_summary, aes(x = gen, y = mean_total_stats)) +
  geom_col(aes(fill = gen), alpha = 0.85, color = "white", width = 0.7) +
  geom_text(aes(label = round(mean_total_stats, 1)), vjust = -0.5, size = 3.5) +
  geom_smooth(aes(x = as.numeric(as.character(gen)), y = mean_total_stats),
              method = "lm", formula = y ~ x, se = FALSE, color = "darkred", 
              linetype = "dashed", linewidth = 1) +
  scale_fill_brewer(palette = "Blues") +
  labs(title = "Mean Total Stats by Generation",
       subtitle = paste0("Trend correlation: r = ", round(cor_total, 3), 
                        " | Dashed line = linear trend"),
       x = "Generation", y = "Mean Total Stats") +
  theme_minimal() +
  theme(legend.position = "none",
        plot.title = element_text(face = "bold", size = 14),
        plot.subtitle = element_text(color = "gray40"))
print(p_total_stats_col)

# Also show median total stats as a complementary view
median_stats <- pokemon %>%
  group_by(gen) %>%
  summarise(median_total_stats = median(total_stats), .groups = "drop")

p_median_col <- ggplot(median_stats, aes(x = gen, y = median_total_stats)) +
  geom_col(aes(fill = gen), alpha = 0.85, color = "white", width = 0.7) +
  geom_text(aes(label = round(median_total_stats, 0)), vjust = -0.5, size = 3.5) +
  scale_fill_brewer(palette = "Greens") +
  labs(title = "Median Total Stats by Generation",
       subtitle = "Less sensitive to outliers (legendaries)",
       x = "Generation", y = "Median Total Stats") +
  theme_minimal() +
  theme(legend.position = "none",
        plot.title = element_text(face = "bold", size = 14),
        plot.subtitle = element_text(color = "gray40"))
print(p_median_col)
```

**Findings:** 

- **Rarity effect:** Legendary/Mythical Pokémon have substantially higher attack (~100-115) and total stats (~580-600) vs Normal (~75 attack, ~410 total). This is a potential confound — if later generations have more legendaries, that could explain apparent power creep.

- **Generation trends:** Both mean attack and mean total stats show upward trends across generations. The correlation with generation is strong for both metrics (attack ~0.76, total stats ~0.84), suggesting broad-based power creep affecting overall Pokémon strength, not just attack.

```{r rarity-boxplot, fig.width=12, fig.height=6}
# Visualize attack by rarity and generation
p_rarity <- pokemon %>%
  ggplot(aes(x = gen, y = attack, fill = rarity)) +
  geom_boxplot(alpha = 0.7) +
  labs(title = "Attack by Generation and Rarity",
       subtitle = "Do legendary Pokémon drive power creep?",
       x = "Generation", y = "Attack", fill = "Rarity") +
  theme_minimal() +
  theme(legend.position = "bottom")
print(p_rarity)

# 3. Create a summary table of counts for each Generation and Rarity
rarity_counts <- pokemon %>%
  count(gen, rarity) %>%           # Count by Gen AND Rarity
  pivot_wider(names_from = rarity, # Spread rarity into columns
              values_from = n, 
              values_fill = 0)     # Fill missing combinations with 0

# Print the table
print(rarity_counts)
```

```{r rarity-counts-viz, fig.width=14, fig.height=6}
# Visualize rarity distribution by generation
rarity_counts_long <- pokemon %>%
  count(gen, rarity)

# Option 1: Stacked bar chart (shows total + composition)
p_stacked <- ggplot(rarity_counts_long, aes(x = gen, y = n, fill = rarity)) +
  geom_col(alpha = 0.85) +
  geom_text(aes(label = ifelse(n > 5, n, "")), 
            position = position_stack(vjust = 0.5), size = 3, color = "white") +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Pokémon Count by Generation and Rarity (Stacked)",
       subtitle = "Total Pokémon per generation with rarity breakdown",
       x = "Generation", y = "Count", fill = "Rarity") +
  theme_minimal() +
  theme(legend.position = "bottom")

# Option 2: Grouped bar chart (easier to compare categories)
p_grouped <- ggplot(rarity_counts_long, aes(x = gen, y = n, fill = rarity)) +
  geom_col(position = position_dodge(width = 0.8), alpha = 0.85) +
  geom_text(aes(label = n), 
            position = position_dodge(width = 0.8), vjust = -0.3, size = 2.5) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Pokémon Count by Generation and Rarity (Grouped)",
       subtitle = "Direct comparison of rarity categories within each generation",
       x = "Generation", y = "Count", fill = "Rarity") +
  theme_minimal() +
  theme(legend.position = "bottom")

gridExtra::grid.arrange(p_stacked, p_grouped, ncol = 2)
```

```{r rarity-proportion-viz, fig.width=10, fig.height=5}
# Option 3: Proportion plot (best for seeing relative composition)
p_proportion <- ggplot(rarity_counts_long, aes(x = gen, y = n, fill = rarity)) +
  geom_col(position = "fill", alpha = 0.85) +
  scale_fill_brewer(palette = "Set2") +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(title = "Rarity Composition by Generation (Proportions)",
       subtitle = "What percentage of each generation is Normal vs Legendary/Mythical?",
       x = "Generation", y = "Proportion", fill = "Rarity") +
  theme_minimal() +
  theme(legend.position = "bottom")
print(p_proportion)

# Calculate proportion of rare Pokémon per generation
rare_proportion <- pokemon %>%
  mutate(is_rare = rarity != "Normal") %>%
  group_by(gen) %>%
  summarise(
    total = n(),
    n_rare = sum(is_rare),
    pct_rare = round(100 * n_rare / total, 1)
  )
cat("\n--- Proportion of Rare Pokémon by Generation ---\n")
print(rare_proportion)
```

**Findings:** The visualizations reveal important patterns in rarity distribution:

- **Gen 7 stands out** with the highest number of SubLegendary and Legendary Pokémon (visible in the stacked/grouped charts)
- **Normal Pokémon dominate** all generations (~85-95%), but the proportion of rare Pokémon varies
- **Rarity inflation?** If later generations have higher proportions of rare Pokémon, this could partially explain apparent power creep (since rares have higher stats)

```{r rarity-bar, fig.width=8, fig.height=6}
# Visualize mean attack per rarity (similar to SAN_assignment_GLM_2.Rmd bar chart)
p_rarity_bar <- pokemon %>%
  group_by(rarity) %>%
  summarise(mean_attack = mean(attack)) %>%
  ggplot(aes(x = rarity, y = mean_attack, fill = rarity)) +
  geom_col() +
  geom_text(aes(label = round(mean_attack, 1)), vjust = -0.5) +
  labs(title = "Mean Attack by Rarity Category",
       x = "Rarity", y = "Mean Attack") +
  theme_minimal() +
  theme(legend.position = "none")
print(p_rarity_bar)
```

**Findings:** Here we can see that the rarity has some consequences on the stats. This is why we will also test things with just the normal pokemons to verify if power creep happens because of rarity, or if there is more to that. The biggest amount of sublegendary, legendary pokemons is by far in generation 7.


## 3.7 Scatter Plot: Attack vs HP

Similar to visualization in glms.Rmd.

```{r scatter-plot, fig.width=10, fig.height=6}
p_scatter <- pokemon %>%
  ggplot(aes(x = hp, y = attack, color = gen)) +
  geom_jitter(alpha = 0.5, size = 2) +
  geom_smooth(method = "lm", se = FALSE, color = "black", linetype = "dashed") +
  labs(title = "Attack vs HP by Generation",
       x = "HP", y = "Attack", color = "Generation") +
  theme_minimal()
print(p_scatter)
```

---

# 4. Linear Regression Models

## 4.1 Basic OLS: Attack ~ Generation

```{r lm-basic}
# FULL DATA
model_basic <- lm(attack ~ gen, data = pokemon)
cat("--- Model 1: Basic OLS (Attack ~ Generation) - FULL DATA ---\n")
summary(model_basic)

# OUTLIER-FREE DATA
model_basic_clean <- lm(attack ~ gen, data = pokemon_clean)
cat("\n--- Model 1: Basic OLS (Attack ~ Generation) - OUTLIER-FREE ---\n")
summary(model_basic_clean)

# NORMAL RARITY ONLY
model_basic_normal <- lm(attack ~ gen, data = pokemon_normal)
cat("\n--- Model 1: Basic OLS (Attack ~ Generation) - NORMAL ONLY ---\n")
summary(model_basic_normal)

# Quick comparison
cat("\n--- R² Comparison ---\n")
cat("Full data R²:", round(summary(model_basic)$r.squared, 4), "\n")
cat("Outlier-free R²:", round(summary(model_basic_clean)$r.squared, 4), "\n")
cat("Normal-only R²:", round(summary(model_basic_normal)$r.squared, 4), "\n")
```

**Findings:** Generation alone explains only 3.2% of variance (R² = 0.032). Normal-only R² = 0.031, nearly identical — confirming power creep exists even among regular Pokémon, not just legendaries.

```{r lm-basic-anova}
# F-test for generation effect - ALL THREE DATASETS
cat("--- F-test for generation effect (Full Data) ---\n")
anova(model_basic)

cat("\n--- F-test for generation effect (Outlier-Free) ---\n")
anova(model_basic_clean)

cat("\n--- F-test for generation effect (Normal Only) ---\n")
anova(model_basic_normal)
```

**Findings:** ANOVA confirms generation is a significant predictor overall. Compare p-values across all three datasets — significance in Normal-only data confirms power creep isn't just legendary inflation.

## 4.2 Multiple Regression: Attack ~ Generation + Controls

```{r lm-full}
# FULL DATA
model_full <- lm(attack ~ gen + hp + defense + sp_attack + sp_defense + speed +
                   height_m + weight_kg + rarity + capture_rate + num_abilities + evo_length,
                 data = pokemon)
cat("--- Model 2: Full OLS (Attack ~ Gen + Controls) - FULL DATA ---\n")
summary(model_full)

# OUTLIER-FREE DATA
model_full_clean <- lm(attack ~ gen + hp + defense + sp_attack + sp_defense + speed +
                         height_m + weight_kg + rarity + capture_rate + num_abilities + evo_length,
                       data = pokemon_clean)
cat("\n--- Model 2: Full OLS (Attack ~ Gen + Controls) - OUTLIER-FREE ---\n")
summary(model_full_clean)

# NORMAL RARITY ONLY (no rarity predictor needed since all are Normal)
model_full_normal <- lm(attack ~ gen + hp + defense + sp_attack + sp_defense + speed +
                          height_m + weight_kg + capture_rate + num_abilities + evo_length,
                        data = pokemon_normal)
cat("\n--- Model 2: Full OLS (Attack ~ Gen + Controls) - NORMAL ONLY ---\n")
summary(model_full_normal)

# Coefficient comparison for generation
cat("\n--- Generation Coefficients Comparison (All Datasets) ---\n")
gen_coef_full <- coef(model_full)[grep("^gen", names(coef(model_full)))]
gen_coef_clean <- coef(model_full_clean)[grep("^gen", names(coef(model_full_clean)))]
gen_coef_normal <- coef(model_full_normal)[grep("^gen", names(coef(model_full_normal)))]
coef_comp <- data.frame(
  Gen = paste0("gen", 2:9),
  Full = round(gen_coef_full, 2),
  Clean = round(gen_coef_clean, 2),
  Normal = round(gen_coef_normal, 2)
)
print(coef_comp)

cat("\n--- R² Comparison ---\n")
cat("Full data R²:", round(summary(model_full)$r.squared, 4), "\n")
cat("Outlier-free R²:", round(summary(model_full_clean)$r.squared, 4), "\n")
cat("Normal-only R²:", round(summary(model_full_normal)$r.squared, 4), "\n")
```

**Findings:** After controlling for other stats, R² jumps to 0.473 (full), 0.478 (clean), and 0.445 (normal-only). Generation coefficients are similar across datasets — the Normal-only model confirms power creep exists independently of legendary inflation.

```{r vif-check}
# Check VIF for multicollinearity
cat("\nVariance Inflation Factors (VIF) - Full Data:\n")
vif(model_full)
```

**Findings:** All VIFs < 2.3, indicating no problematic multicollinearity. The model coefficients are reliable.

## 4.3 Model with Interaction: Gen × Rarity

```{r lm-interaction}
model_interaction <- lm(attack ~ gen * rarity + hp + defense + sp_attack + 
                          sp_defense + speed + height_m + weight_kg,
                        data = pokemon)
cat("--- Model 3: OLS with Gen × Rarity Interaction ---\n")
summary(model_interaction)
```

## 4.4 Model Comparison

```{r model-comparison}
# Compare models using ANOVA
cat("--- Model Comparison (ANOVA) ---\n")
anova(model_basic, model_full, model_interaction)

# AIC/BIC comparison
cat("\nModel Comparison (AIC/BIC):\n")
cat("Basic model: AIC =", AIC(model_basic), ", BIC =", BIC(model_basic), "\n")
cat("Full model: AIC =", AIC(model_full), ", BIC =", BIC(model_full), "\n")
cat("Interaction model: AIC =", AIC(model_interaction), ", BIC =", BIC(model_interaction), "\n")
```

**Findings:** The full model is significantly better than basic (p < 2e-16). The interaction model does NOT significantly improve over the full model (p = 0.33), meaning the generation effect doesn't vary much by rarity category. AIC/BIC both favor the full model.

## 4.5 Diagnostic Plots

```{r diagnostics, fig.width=12, fig.height=9}
par(mfrow = c(2, 2))
plot(model_full, main = "OLS Diagnostic Plots")
plot(model_full_clean, main = "OLS Diagnostic Plots (clean)")
plot(model_full_normal, main = "OLS Diagnostic Plots (normal)")

```

---

# 5. Backward Stepwise Selection

```{r stepwise}
# Start with full model
model_step_full <- lm(attack ~ gen + hp + defense + sp_attack + sp_defense + speed +
                        height_m + log_weight + rarity + num_abilities + evo_length +
                        capture_rate + has_mega_evolution,
                      data = pokemon)

# Backward selection using AIC
model_step <- stepAIC(model_step_full, direction = "backward", trace = FALSE)
cat("--- Backward Stepwise Selection Result ---\n")
summary(model_step)
cat("\nFinal model formula:\n")
formula(model_step)
```

**Findings:** Stepwise selection retains generation in the model, confirming it has predictive value. The final model typically keeps gen, hp, defense, sp_defense, speed, and weight. Features like num_abilities and evo_length are often dropped.

---

# 6. LASSO, Ridge, and Elastic Net Regression

## 6.1 Prepare Data

```{r regularization-prep}
# Create design matrix with generation dummies - FULL DATA
X <- model.matrix(~ gen + hp + defense + sp_attack + sp_defense + speed +
                    height_m + weight_kg + capture_rate + num_abilities + evo_length +
                    rarity - 1, data = pokemon)
y <- pokemon$attack

# OUTLIER-FREE DATA
X_clean <- model.matrix(~ gen + hp + defense + sp_attack + sp_defense + speed +
                          height_m + weight_kg + capture_rate + num_abilities + evo_length +
                          rarity - 1, data = pokemon_clean)
y_clean <- pokemon_clean$attack
```

## 6.2 Ridge Regression (α = 0)

```{r ridge}
set.seed(42)
cv_ridge <- cv.glmnet(X, y, alpha = 0, nfolds = 10)
cat("--- Ridge Regression ---\n")
cat("Optimal lambda (min):", cv_ridge$lambda.min, "\n")
cat("Optimal lambda (1se):", cv_ridge$lambda.1se, "\n")

ridge_model <- glmnet(X, y, alpha = 0, lambda = cv_ridge$lambda.min)
cat("\nRidge coefficients:\n")
print(coef(ridge_model))
```

## 6.3 LASSO Regression (α = 1)

```{r lasso}
# FULL DATA
set.seed(42)
cv_lasso <- cv.glmnet(X, y, alpha = 1, nfolds = 10)
cat("--- LASSO Regression (Full Data) ---\n")
cat("Optimal lambda (min):", cv_lasso$lambda.min, "\n")

lasso_model <- glmnet(X, y, alpha = 1, lambda = cv_lasso$lambda.min)
cat("\nLASSO coefficients (non-zero only = selected features):\n")
lasso_coef <- coef(lasso_model)
print(lasso_coef[lasso_coef[,1] != 0, ])

n_selected <- sum(lasso_coef[,1] != 0) - 1
cat("\nNumber of features selected by LASSO:", n_selected, "\n")

# OUTLIER-FREE DATA
set.seed(42)
cv_lasso_clean <- cv.glmnet(X_clean, y_clean, alpha = 1, nfolds = 10)
cat("\n--- LASSO Regression (Outlier-Free) ---\n")
cat("Optimal lambda (min):", cv_lasso_clean$lambda.min, "\n")

lasso_model_clean <- glmnet(X_clean, y_clean, alpha = 1, lambda = cv_lasso_clean$lambda.min)
cat("\nLASSO coefficients (non-zero only = selected features):\n")
lasso_coef_clean <- coef(lasso_model_clean)
print(lasso_coef_clean[lasso_coef_clean[,1] != 0, ])

n_selected_clean <- sum(lasso_coef_clean[,1] != 0) - 1
cat("\nNumber of features selected by LASSO:", n_selected_clean, "\n")

# Compare generation coefficients
cat("\n--- LASSO Generation Coefficients Comparison ---\n")

# Get all coefficient names from both models
coef_full_mat <- as.matrix(lasso_coef)
coef_clean_mat <- as.matrix(lasso_coef_clean)

# Find generation-related coefficients (matching "gen" in name)
gen_idx_full <- grep("^gen", rownames(coef_full_mat))
gen_idx_clean <- grep("^gen", rownames(coef_clean_mat))

# Get the generation names that exist in both
gen_names_full <- rownames(coef_full_mat)[gen_idx_full]
gen_names_clean <- rownames(coef_clean_mat)[gen_idx_clean]
all_gen_names <- union(gen_names_full, gen_names_clean)

# Safely extract coefficients (0 if not present)
get_coef <- function(coef_mat, name) {
  if (name %in% rownames(coef_mat)) coef_mat[name, 1] else 0
}

lasso_gen_full <- sapply(all_gen_names, function(n) get_coef(coef_full_mat, n))
lasso_gen_clean <- sapply(all_gen_names, function(n) get_coef(coef_clean_mat, n))

lasso_comparison <- data.frame(
  Gen = all_gen_names,
  Full = round(lasso_gen_full, 3),
  Clean = round(lasso_gen_clean, 3),
  Both_Selected = ifelse(lasso_gen_full != 0 & lasso_gen_clean != 0, "Yes", 
                         ifelse(lasso_gen_full != 0 | lasso_gen_clean != 0, "One only", "Neither"))
)
rownames(lasso_comparison) <- NULL
print(lasso_comparison)
```

**Findings:** Compare which generation dummies LASSO selects in full vs outlier-free data. If the same generations are retained, the feature selection is robust to outliers.

## 6.4 Elastic Net (α = 0.5)

```{r elastic-net}
set.seed(42)
cv_enet <- cv.glmnet(X, y, alpha = 0.5, nfolds = 10)
cat("--- Elastic Net Regression ---\n")
cat("Optimal lambda:", cv_enet$lambda.min, "\n")

enet_model <- glmnet(X, y, alpha = 0.5, lambda = cv_enet$lambda.min)
cat("\nElastic Net coefficients:\n")
print(coef(enet_model))
```

## 6.5 Regularization Paths

```{r regularization-plots, fig.width=12, fig.height=4}
par(mfrow = c(1, 3))
plot(cv_ridge, main = "Ridge CV")
plot(cv_lasso, main = "LASSO CV")
plot(cv_enet, main = "Elastic Net CV")
```

---

# 7. Generalized Linear Models (GLMs)

## 7.1 Gaussian GLM

Equivalent to OLS for continuous target.

```{r glm-gaussian}
# FULL DATA
glm_gaussian <- glm(attack ~ gen + hp + defense + sp_attack + sp_defense + speed +
                      rarity + weight_kg + height_m,
                    data = pokemon, family = gaussian(link = "identity"))
cat("--- GLM: Gaussian (Attack ~ predictors) - FULL DATA ---\n")
summary(glm_gaussian)

# OUTLIER-FREE DATA
glm_gaussian_clean <- glm(attack ~ gen + hp + defense + sp_attack + sp_defense + speed +
                            rarity + weight_kg + height_m,
                          data = pokemon_clean, family = gaussian(link = "identity"))
cat("\n--- GLM: Gaussian (Attack ~ predictors) - OUTLIER-FREE ---\n")
summary(glm_gaussian_clean)

# NORMAL RARITY ONLY (no rarity predictor)
glm_gaussian_normal <- glm(attack ~ gen + hp + defense + sp_attack + sp_defense + speed +
                             weight_kg + height_m,
                           data = pokemon_normal, family = gaussian(link = "identity"))
cat("\n--- GLM: Gaussian (Attack ~ predictors) - NORMAL ONLY ---\n")
summary(glm_gaussian_normal)

# AIC comparison
cat("\n--- AIC Comparison ---\n")
cat("Full data AIC:", round(AIC(glm_gaussian), 1), "\n")
cat("Outlier-free AIC:", round(AIC(glm_gaussian_clean), 1), "\n")
cat("Normal-only AIC:", round(AIC(glm_gaussian_normal), 1), "(not directly comparable - different N)\n")
```

**Findings:** Gaussian GLM shows which generations are significant. The Normal-only model tests power creep without legendary influence — significant generation effects here confirm the trend exists among regular Pokémon.

## 7.2 Gamma GLM

For positive continuous data, handles heteroscedasticity.

```{r glm-gamma}
glm_gamma <- glm(attack ~ gen + hp + defense + sp_attack + sp_defense + speed +
                   rarity + weight_kg,
                 data = pokemon, family = Gamma(link = "log"))
cat("--- GLM: Gamma with log link ---\n")
summary(glm_gamma)
```

## 7.3 Poisson GLM for Count Data

Using `num_abilities` as the response variable.

```{r glm-poisson}
glm_poisson <- glm(num_abilities ~ gen + rarity + hp + attack + defense + speed,
                   data = pokemon, family = poisson(link = "log"))
cat("--- GLM: Poisson (Number of Abilities) ---\n")
summary(glm_poisson)
```

**Findings:** Interestingly, later generations (7, 9) have *fewer* abilities than Gen 1. Rarer Pokémon also have fewer abilities. This is the opposite of power creep for this metric — newer Pokémon are simpler in terms of ability count.

### Overdispersion Check

```{r overdispersion-check}
# Check mean vs variance (Poisson assumption)
cat("\nPoisson Assumption Check (mean ≈ variance):\n")
cat("Mean num_abilities:", mean(pokemon$num_abilities), "\n")
cat("Variance num_abilities:", var(pokemon$num_abilities), "\n")

# Check for overdispersion using AER package (from SAN_assignment_GLM_2.Rmd)
cat("\n--- Dispersion Test (from AER package) ---\n")
disp_test <- dispersiontest(glm_poisson)
print(disp_test)

# Manual dispersion calculation
dispersion <- sum(residuals(glm_poisson, type = "pearson")^2) / glm_poisson$df.residual
cat("\nManual Dispersion parameter:", dispersion, "\n")

if (disp_test$p.value < 0.05 || dispersion > 1.5) {
  cat("WARNING: Overdispersion detected! Consider Negative Binomial or Quasi-Poisson\n")
} else {
  cat("No significant overdispersion detected.\n")
}
```

**Findings:** Variance (0.55) is actually *less* than mean (2.37), indicating underdispersion rather than overdispersion. The Poisson model is appropriate here. This is unusual but makes sense — ability counts are constrained to a narrow range (typically 2-3).

## 7.4 Quasi-Poisson (for Overdispersion)

```{r glm-quasi}
glm_quasi <- glm(num_abilities ~ gen + rarity + hp + attack + defense + speed,
                 data = pokemon, family = quasipoisson(link = "log"))
cat("--- GLM: Quasi-Poisson (correcting for overdispersion) ---\n")
summary(glm_quasi)
```

## 7.5 Negative Binomial GLM

From SAN_assignment_GLM_2.Rmd - handles overdispersion better than Quasi-Poisson.

```{r glm-nb}
glm_nb <- MASS::glm.nb(num_abilities ~ gen + rarity + hp + attack + defense + speed,
                       data = pokemon)
cat("--- GLM: Negative Binomial (better for overdispersion) ---\n")
summary(glm_nb)
```

## 7.6 Compare GLMs

```{r compare-glms}
# Compare Poisson vs Negative Binomial using AIC
cat("\nAIC Comparison (Poisson vs Negative Binomial):\n")
AIC(glm_poisson, glm_nb)

# Compare using rcompanion (from SAN_assignment_GLM_2.Rmd)
cat("\n--- Compare GLMs using rcompanion ---\n")
glm_comparison <- compareGLM(glm_poisson, glm_quasi, glm_nb)
print(glm_comparison)
```

---

# 8. Comprehensive Ablation Study with Cross-Validation

Inspired by SAN_assignment_GLM_2.Rmd.

```{r ablation-setup, message=FALSE, warning=FALSE}
# Setup 10-fold cross-validation
train_control <- trainControl(method = "cv", number = 10, verboseIter = FALSE)
set.seed(42)

# Prepare data (ensure factors are properly set)
pokemon_cv <- pokemon %>%
  mutate(
    gen = as.factor(gen),
    rarity = as.factor(rarity),
    class_primary = as.factor(class_primary)
  ) %>%
  dplyr::select(attack, gen, hp, defense, sp_attack, sp_defense, speed, 
                weight_kg, height_m, rarity, num_abilities, evo_length, gen_numeric) %>%
  na.omit()

cat("Running ablation study with 10-fold CV...\n")
```

## 8.1 Train All Models

```{r ablation-models, message=FALSE, warning=FALSE}
# Model 1: Null model (intercept only)
m_null <- train(
  x = data.frame(intercept = rep(1, nrow(pokemon_cv))),
  y = pokemon_cv$attack,
  method = "lm",
  trControl = train_control
)

# Model 2: Simple LM (attack ~ gen only)
m_gen_only <- train(
  attack ~ gen,
  data = pokemon_cv,
  method = "lm",
  trControl = train_control
)

# Model 3: LM with gen + basic stats
m_lm_basic <- train(
  attack ~ gen + hp + defense + speed,
  data = pokemon_cv,
  method = "lm",
  trControl = train_control
)

# Model 4: LM with all predictors
m_lm_full <- train(
  attack ~ gen + hp + defense + sp_attack + sp_defense + speed + rarity + weight_kg,
  data = pokemon_cv,
  method = "lm",
  trControl = train_control
)

# Model 5: GLM Gaussian
m_glm_gaussian <- train(
  attack ~ gen + hp + defense + sp_attack + sp_defense + speed + rarity,
  data = pokemon_cv,
  method = "glm",
  family = "gaussian",
  trControl = train_control
)

# Model 6: GLM with polynomial term (from SAN_assignment_GLM_2.Rmd)
m_glm_poly <- train(
  attack ~ gen + hp + defense + sp_attack + sp_defense + poly(speed, 2) + rarity,
  data = pokemon_cv,
  method = "glm",
  family = "gaussian",
  trControl = train_control
)

# Model 7: GLM with spline (from SAN_assignment_GLM_2.Rmd)
m_glm_spline <- train(
  attack ~ gen + ns(hp, df = 3) + ns(defense, df = 3) + sp_attack + sp_defense + 
    ns(speed, df = 3) + rarity,
  data = pokemon_cv,
  method = "glm",
  family = "gaussian",
  trControl = train_control
)

# Model 8: GLM with gen as numeric (linear trend)
m_glm_trend <- train(
  attack ~ gen_numeric + hp + defense + sp_attack + sp_defense + speed + rarity,
  data = pokemon_cv,
  method = "glm",
  family = "gaussian",
  trControl = train_control
)

# Model 9: GLM with gen × rarity interaction
m_glm_interaction <- train(
  attack ~ gen * rarity + hp + defense + sp_attack + sp_defense + speed,
  data = pokemon_cv,
  method = "glm",
  family = "gaussian",
  trControl = train_control
)
```

## 8.2 Compare Models

```{r ablation-compare}
# Compile and compare all models using resamples() (from SAN_assignment_GLM_2.Rmd)
model_list <- resamples(list(
  NULL_MODEL = m_null,
  GEN_ONLY = m_gen_only,
  LM_BASIC = m_lm_basic,
  LM_FULL = m_lm_full,
  GLM_GAUSSIAN = m_glm_gaussian,
  GLM_POLY = m_glm_poly,
  GLM_SPLINE = m_glm_spline,
  GLM_TREND = m_glm_trend,
  GLM_INTERACTION = m_glm_interaction
))

# Print summary of cross-validation results
cat("--- Cross-Validation Results Summary ---\n")
summary(model_list)
```

**Findings:** GLM_SPLINE achieves the best CV performance (RMSE ≈ 21, R² ≈ 0.51). Generation-only model barely beats the null model, confirming that generation alone has limited predictive power. The key improvement comes from adding combat stats, not from modeling generation non-linearly.

```{r ablation-dotplot, fig.width=10, fig.height=8}
# Create dotplot for visual comparison (from SAN_assignment_GLM_2.Rmd)
dotplot(model_list, main = "Model Comparison: 10-Fold CV (Lower RMSE/MAE = Better)")
```

## 8.3 Model Rankings

```{r model-rankings}
# Create comparison table
cv_results <- summary(model_list)

cat("--- Model Rankings by RMSE ---\n")
rmse_means <- cv_results$statistics$RMSE[, "Mean"]
rmse_ranking <- sort(rmse_means)
print(round(rmse_ranking, 3))

cat("\n--- Model Rankings by MAE ---\n")
mae_means <- cv_results$statistics$MAE[, "Mean"]
mae_ranking <- sort(mae_means)
print(round(mae_ranking, 3))

cat("\n--- Model Rankings by R² ---\n")
r2_means <- cv_results$statistics$Rsquared[, "Mean"]
r2_ranking <- sort(r2_means, decreasing = TRUE)
print(round(r2_ranking, 3))
```

## 8.4 AIC Comparison

```{r aic-comparison}
cat("--- AIC Comparison of Final Models ---\n")
final_models_aic <- data.frame(
  Model = c("NULL_MODEL", "GEN_ONLY", "LM_BASIC", "LM_FULL", "GLM_GAUSSIAN",
            "GLM_POLY", "GLM_SPLINE", "GLM_TREND", "GLM_INTERACTION"),
  AIC = c(
    AIC(m_null$finalModel),
    AIC(m_gen_only$finalModel),
    AIC(m_lm_basic$finalModel),
    AIC(m_lm_full$finalModel),
    AIC(m_glm_gaussian$finalModel),
    AIC(m_glm_poly$finalModel),
    AIC(m_glm_spline$finalModel),
    AIC(m_glm_trend$finalModel),
    AIC(m_glm_interaction$finalModel)
  )
)
final_models_aic <- final_models_aic %>% arrange(AIC)
print(final_models_aic)
```

---

# 9. Likelihood Ratio Tests for Generation Effect

```{r lr-test}
# Full model with generation
full_model_lr <- glm(attack ~ gen + hp + defense + sp_attack + sp_defense + speed +
                    rarity + weight_kg, data = pokemon, family = gaussian)

# Reduced model without generation
reduced_model <- glm(attack ~ hp + defense + sp_attack + sp_defense + speed +
                       rarity + weight_kg, data = pokemon, family = gaussian)

# Likelihood ratio test
lr_test <- lrtest(reduced_model, full_model_lr)
cat("--- Likelihood Ratio Test: Is Generation Significant? ---\n")
print(lr_test)

if (lr_test$`Pr(>Chisq)`[2] < 0.05) {
  cat("\nConclusion: Generation IS a significant predictor of Attack (p < 0.05)\n")
} else {
  cat("\nConclusion: Generation is NOT a significant predictor after controlling for other variables\n")
}
```

**Findings:** LR test p = 0.0077 (full data) and p = 0.002 (outlier-free) confirm generation adds significant explanatory power beyond other stats. This is the KEY evidence for power creep — even after controlling for HP/Defense/Speed/Rarity/Weight, generation still matters, and this holds after removing outliers.

---

# 10. Non-Linear Effects with Splines and Polynomials

## 10.1 Polynomial Model

```{r poly-model}
model_poly <- lm(attack ~ poly(gen_numeric, 2) + 
                   poly(hp, 2) + poly(defense, 2) + 
                   poly(speed, 2) + rarity,
                 data = pokemon)
cat("--- Model with Polynomial Terms (degree 2) ---\n")
summary(model_poly)
```

## 10.2 Natural Spline Model

```{r spline-model}
model_spline <- lm(attack ~ ns(gen_numeric, df = 3) + 
                     ns(hp, df = 3) + ns(defense, df = 3) + 
                     ns(speed, df = 3) + rarity,
                   data = pokemon)
cat("--- Model with Natural Splines (df=3) ---\n")
summary(model_spline)
```

## 10.3 Linear Baseline

```{r linear-baseline}
model_linear <- lm(attack ~ gen_numeric + hp + defense + speed + rarity,
                   data = pokemon)
```

## 10.4 Model Comparison

```{r nonlinear-comparison}
cat("--- Comparison: Linear vs Polynomial vs Spline ---\n")
cat("\nAIC comparison:\n")
cat("Linear model:     AIC =", round(AIC(model_linear), 2), "\n")
cat("Polynomial model: AIC =", round(AIC(model_poly), 2), "\n")
cat("Spline model:     AIC =", round(AIC(model_spline), 2), "\n")

cat("\nBIC comparison:\n")
cat("Linear model:     BIC =", round(BIC(model_linear), 2), "\n")
cat("Polynomial model: BIC =", round(BIC(model_poly), 2), "\n")
cat("Spline model:     BIC =", round(BIC(model_spline), 2), "\n")

# ANOVA comparison (nested models)
cat("\nANOVA: Linear vs Spline\n")
anova(model_linear, model_spline)
```

**Findings:** Polynomial and spline models have lower AIC than linear, suggesting non-linear relationships exist (especially for HP and Defense). However, the generation effect itself is approximately linear — the non-linearity comes from combat stats, not from generation.

## 10.5 Visualize Non-Linear Effects

```{r nonlinear-viz, fig.width=10, fig.height=6, warning=FALSE, message=FALSE}
# Create prediction data for visualization
gen_seq <- seq(1, 9, length.out = 100)
pred_data <- data.frame(
  gen_numeric = gen_seq,
  hp = mean(pokemon$hp),
  defense = mean(pokemon$defense),
  speed = mean(pokemon$speed),
  rarity = factor("Normal", levels = levels(pokemon$rarity))
)

pred_data$pred_linear <- predict(model_linear, newdata = pred_data)
pred_data$pred_poly <- predict(model_poly, newdata = pred_data)
pred_data$pred_spline <- predict(model_spline, newdata = pred_data)

# Plot comparison of linear vs polynomial vs spline
p_nonlinear <- ggplot() +
  geom_jitter(data = pokemon, aes(x = gen_numeric, y = attack), 
              alpha = 0.3, width = 0.2, color = "gray50") +
  geom_line(data = pred_data, aes(x = gen_numeric, y = pred_linear, color = "Linear"), 
            size = 1.2) +
  geom_line(data = pred_data, aes(x = gen_numeric, y = pred_poly, color = "Polynomial"), 
            size = 1.2) +
  geom_line(data = pred_data, aes(x = gen_numeric, y = pred_spline, color = "Spline"), 
            size = 1.2) +
  # Add loess smooth for reference (from glms.Rmd)
  geom_smooth(data = pokemon, aes(x = gen_numeric, y = attack), 
              method = "loess", color = "black", linetype = "dashed", 
              alpha = 0.3, se = FALSE) +
  scale_color_manual(values = c("Linear" = "blue", "Polynomial" = "red", "Spline" = "green")) +
  labs(title = "Comparison of Non-linear Models for Generation Effect",
       subtitle = "Dashed line = loess smooth (reference); Points = actual Pokémon",
       x = "Generation", y = "Attack", color = "Model Type") +
  theme_minimal() +
  theme(legend.position = "bottom")
print(p_nonlinear)
```

---

# 11. Time-Split Cross-Validation

Train on Gen ≤5, Test on Gen ≥6.

```{r time-split-data}
# Split data and retain full factor levels to avoid "new levels" issues
train_data <- pokemon %>%
  filter(gen_numeric <= 5) %>%
  mutate(
    gen = factor(gen, levels = levels(pokemon$gen)),
    rarity = factor(rarity, levels = levels(pokemon$rarity))
  )

test_data <- pokemon %>%
  filter(gen_numeric >= 6) %>%
  mutate(
    gen = factor(gen, levels = levels(pokemon$gen)),
    rarity = factor(rarity, levels = levels(pokemon$rarity))
  )

cat("Training set (Gen 1-5):", nrow(train_data), "Pokémon\n")
cat("Test set (Gen 6-9):", nrow(test_data), "Pokémon\n")
```

```{r time-split-models}
# Train models on Gen 1-5
# Use numeric generation to avoid factor level issues when predicting on newer gens
cv_model_lm <- lm(attack ~ gen_numeric + hp + defense + sp_attack + sp_defense + speed +
                    rarity + weight_kg, data = train_data)

cv_model_glm <- glm(attack ~ gen_numeric + hp + defense + sp_attack + sp_defense + speed +
                      rarity + weight_kg, data = train_data, family = Gamma(link = "log"))

# Predict on Gen 6-9
pred_lm <- predict(cv_model_lm, newdata = test_data)
pred_glm <- predict(cv_model_glm, newdata = test_data, type = "response")

# Compute metrics
rmse_lm <- sqrt(mean((test_data$attack - pred_lm)^2))
mae_lm <- mean(abs(test_data$attack - pred_lm))
r2_lm <- 1 - sum((test_data$attack - pred_lm)^2) / sum((test_data$attack - mean(test_data$attack))^2)

rmse_glm <- sqrt(mean((test_data$attack - pred_glm)^2))
mae_glm <- mean(abs(test_data$attack - pred_glm))
r2_glm <- 1 - sum((test_data$attack - pred_glm)^2) / sum((test_data$attack - mean(test_data$attack))^2)

cat("--- Time-Split CV Results ---\n")
cat("OLS Model:\n")
cat("  RMSE:", rmse_lm, "\n")
cat("  MAE:", mae_lm, "\n")
cat("  R²:", r2_lm, "\n")

cat("\nGamma GLM:\n")
cat("  RMSE:", rmse_glm, "\n")
cat("  MAE:", mae_glm, "\n")
cat("  R²:", r2_glm, "\n")
```

**Findings:** OLS generalizes better to future generations (RMSE 24) than Gamma GLM (RMSE 30). The degradation from 10-fold CV (~21) to time-split (~24) suggests the generation effect is not perfectly linear — future generations don't follow the same pattern as past ones. This is a caveat for the power creep claim.

```{r time-split-viz, fig.width=10, fig.height=6}
# Visualize predictions vs actual
pred_comparison <- data.frame(
  actual = test_data$attack,
  pred_lm = pred_lm,
  pred_glm = pred_glm,
  gen = test_data$gen
)

p_cv <- ggplot(pred_comparison, aes(x = actual)) +
  geom_point(aes(y = pred_lm, color = "OLS"), alpha = 0.5) +
  geom_point(aes(y = pred_glm, color = "Gamma GLM"), alpha = 0.5) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "black") +
  labs(title = "Time-Split CV: Predicted vs Actual Attack (Gen 6-9)",
       subtitle = "Model trained on Gen 1-5, tested on Gen 6-9",
       x = "Actual Attack", y = "Predicted Attack",
       color = "Model") +
  theme_minimal()
print(p_cv)
```

---

# 12. Robust Regression (Huber)

```{r robust-regression}
# Fit robust regression (Huber M-estimator)
model_robust <- rlm(attack ~ gen + hp + defense + sp_attack + sp_defense + speed +
                      rarity + weight_kg, data = pokemon)
cat("--- Robust Regression (Huber) ---\n")
summary(model_robust)
```

```{r robust-comparison}
# Compare with OLS
cat("\nComparison of coefficients (OLS vs Robust):\n")
coef_comparison <- data.frame(
  OLS = coef(model_full)[names(coef(model_robust))],
  Robust = coef(model_robust)
)
print(round(coef_comparison, 4))
```

**Findings:** Robust regression coefficients are similar to OLS, indicating outliers don't drastically affect the generation effect estimates. This strengthens confidence in the OLS results.

---

# 13. Final Analysis: Is There Power Creep?

## 13.1 Generation Coefficients

```{r gen-coefficients}
# Extract generation coefficients from models - including all three datasets
gen_coefs <- data.frame(
  Generation = paste0("Gen", 2:9),
  OLS_Full = coef(model_full)[grep("^gen", names(coef(model_full)))],
  OLS_Clean = coef(model_full_clean)[grep("^gen", names(coef(model_full_clean)))],
  OLS_Normal = coef(model_full_normal)[grep("^gen", names(coef(model_full_normal)))],
  Robust = coef(model_robust)[grep("^gen", names(coef(model_robust)))]
)
rownames(gen_coefs) <- NULL

cat("--- Generation Coefficients (relative to Gen 1) ---\n")
cat("Columns: OLS_Full = full data, OLS_Clean = outlier-free, OLS_Normal = normal rarity only, Robust = Huber\n\n")
gen_coefs %>%
  mutate(across(where(is.numeric), ~ round(.x, 3))) %>%
  knitr::kable()
```

**Findings:** Compare coefficients across all datasets. The OLS_Normal column shows the generation effect among regular Pokémon only — similar values to OLS_Full indicate power creep isn't driven by legendaries.

## 13.2 Linear Trend Analysis

```{r trend-analysis}
# Test linear trend in generation - FULL DATA
trend_model <- lm(attack ~ gen_numeric + hp + defense + sp_attack + sp_defense + 
                    speed + rarity, data = pokemon)
cat("--- Linear Trend of Generation on Attack (Full Data) ---\n")
cat("Coefficient of gen_numeric:", round(coef(trend_model)["gen_numeric"], 3), "\n")
cat("p-value:", format(summary(trend_model)$coefficients["gen_numeric", "Pr(>|t|)"], digits = 4), "\n")

# OUTLIER-FREE DATA
trend_model_clean <- lm(attack ~ gen_numeric + hp + defense + sp_attack + sp_defense + 
                          speed + rarity, data = pokemon_clean)
cat("\n--- Linear Trend of Generation on Attack (Outlier-Free) ---\n")
cat("Coefficient of gen_numeric:", round(coef(trend_model_clean)["gen_numeric"], 3), "\n")
cat("p-value:", format(summary(trend_model_clean)$coefficients["gen_numeric", "Pr(>|t|)"], digits = 4), "\n")

# NORMAL RARITY ONLY (no rarity predictor)
trend_model_normal <- lm(attack ~ gen_numeric + hp + defense + sp_attack + sp_defense + 
                           speed, data = pokemon_normal)
cat("\n--- Linear Trend of Generation on Attack (Normal Only) ---\n")
cat("Coefficient of gen_numeric:", round(coef(trend_model_normal)["gen_numeric"], 3), "\n")
cat("p-value:", format(summary(trend_model_normal)$coefficients["gen_numeric", "Pr(>|t|)"], digits = 4), "\n")

cat("\n--- Interpretation ---\n")
cat("Full data: Each generation increases attack by ~", round(coef(trend_model)["gen_numeric"], 1), " points\n")
cat("Outlier-free: Each generation increases attack by ~", round(coef(trend_model_clean)["gen_numeric"], 1), " points\n")
cat("Normal-only: Each generation increases attack by ~", round(coef(trend_model_normal)["gen_numeric"], 1), " points\n")
```

**Findings:** The gen_numeric coefficient represents the average attack increase per generation. All three datasets show a positive coefficient (~1-2 attack points per generation), confirming a consistent upward trend across full, outlier-free, and normal-only data.

## 13.3 Mean Attack by Generation

```{r mean-by-gen}
# Overall mean attack by generation - BOTH DATASETS
mean_by_gen <- pokemon %>%
  group_by(gen) %>%
  summarise(mean_attack = mean(attack)) %>%
  arrange(gen)

mean_by_gen_clean <- pokemon_clean %>%
  group_by(gen) %>%
  summarise(mean_attack = mean(attack)) %>%
  arrange(gen)

mean_by_gen_normal <- pokemon_normal %>%
  group_by(gen) %>%
  summarise(mean_attack = mean(attack)) %>%
  arrange(gen)

cat("--- Mean Attack by Generation (All Three Datasets) ---\n")
comparison_mean <- data.frame(
  gen = 1:9,
  full = mean_by_gen$mean_attack,
  clean = mean_by_gen_clean$mean_attack,
  normal = mean_by_gen_normal$mean_attack
) %>% mutate(
  diff_clean = round(full - clean, 1),
  diff_normal = round(full - normal, 1)
)
print(comparison_mean)

# Calculate trends
trend <- cor(as.numeric(mean_by_gen$gen), mean_by_gen$mean_attack)
trend_clean <- cor(as.numeric(mean_by_gen_clean$gen), mean_by_gen_clean$mean_attack)
trend_normal <- cor(as.numeric(mean_by_gen_normal$gen), mean_by_gen_normal$mean_attack)

cat("\n--- Correlation (gen vs mean_attack) ---\n")
cat("Full data:", round(trend, 3), "\n")
cat("Outlier-free:", round(trend_clean, 3), "\n")
cat("Normal-only:", round(trend_normal, 3), "\n")
```

**Findings:** Correlations (gen vs mean_attack): Full = 0.757, Outlier-free = 0.725, Normal-only = 0.732. All three are strong positive correlations. The Normal-only value (0.732) is very similar to full data, confirming power creep exists among regular Pokémon — legendaries aren't driving the trend.

## 13.4 Summary Visualization: Full vs Outlier-Free Trends

```{r trend-comparison-plot, fig.width=12, fig.height=6}
# Create comparison plot of mean attack trends - ALL THREE DATASETS
trend_data <- data.frame(
  gen = rep(1:9, 3),
  mean_attack = c(mean_by_gen$mean_attack, mean_by_gen_clean$mean_attack, mean_by_gen_normal$mean_attack),
  dataset = rep(c("Full Data", "Outlier-Free", "Normal Only"), each = 9)
)

p_trend_compare <- ggplot(trend_data, aes(x = gen, y = mean_attack, color = dataset, shape = dataset)) +
  geom_point(size = 4) +
  geom_line(linewidth = 1) +
  geom_smooth(method = "lm", se = FALSE, linetype = "dashed", alpha = 0.5) +
  scale_x_continuous(breaks = 1:9) +
  labs(title = "Mean Attack by Generation: All Three Datasets",
       subtitle = paste0("Correlation: Full = ", round(trend, 3), 
                        ", Outlier-Free = ", round(trend_clean, 3),
                        ", Normal-Only = ", round(trend_normal, 3)),
       x = "Generation", y = "Mean Attack",
       color = "Dataset", shape = "Dataset") +
  theme_minimal() +
  theme(legend.position = "bottom")
print(p_trend_compare)
```

**Findings:** This visualization compares generation trends across all three datasets. The Normal-only line shows power creep among regular Pokémon — if it tracks the Full data line, legendaries aren't inflating the trend.

---

# 14. Additional Verification Checks for Power Creep

These checks address potential confounds and strengthen (or weaken) the power creep claim.

## 14.1 Monotonic Trend Test (Jonckheere-Terpstra)

Is there a consistent *ordering* of attack across generations, not just correlation?

```{r monotonic-test, warning=FALSE, message=FALSE}
# Jonckheere-Terpstra test for ordered alternatives
# H0: No trend; H1: Monotonically increasing trend
# FULL DATA
jt_test <- JonckheereTerpstraTest(pokemon$attack, as.numeric(pokemon$gen), 
                                   alternative = "increasing")
cat("--- Jonckheere-Terpstra Test for Monotonic Trend (Full Data) ---\n")
print(jt_test)

# OUTLIER-FREE DATA
jt_test_clean <- JonckheereTerpstraTest(pokemon_clean$attack, as.numeric(pokemon_clean$gen), 
                                         alternative = "increasing")
cat("\n--- Jonckheere-Terpstra Test for Monotonic Trend (Outlier-Free) ---\n")
print(jt_test_clean)

cat("\n--- Comparison ---\n")
cat("Full data p-value:", format(jt_test$p.value, digits = 4), 
    ifelse(jt_test$p.value < 0.05, "(significant)", "(not significant)"), "\n")
cat("Outlier-free p-value:", format(jt_test_clean$p.value, digits = 4),
    ifelse(jt_test_clean$p.value < 0.05, "(significant)", "(not significant)"), "\n")
```

**Findings:** Jonckheere-Terpstra test p = 2.09e-05 (full) and p = 1.61e-05 (outlier-free) confirm a significant monotonically increasing trend. This non-parametric test is more appropriate than Pearson correlation because it tests for *ordered* increase, not just linear association. Both datasets show highly significant results.

## 14.2 Rarity Distribution by Generation

Is power creep driven by more legendaries in later generations?

```{r rarity-by-gen}
# Check if rarity distribution changes across generations
rarity_by_gen <- pokemon %>%
  group_by(gen, rarity) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(gen) %>%
  mutate(prop = n / sum(n)) %>%
  filter(rarity != "Normal")

cat("--- Proportion of Rare Pokémon by Generation ---\n")
print(rarity_by_gen %>% pivot_wider(names_from = rarity, values_from = prop, values_fill = 0))

# Chi-square test: Is rarity distribution independent of generation?
rarity_table <- table(pokemon$gen, pokemon$rarity)
chi_test <- chisq.test(rarity_table)
cat("\nChi-square test (Rarity ~ Generation):\n")
print(chi_test)
```

```{r rarity-by-gen-plot, fig.width=10, fig.height=5}
# Visualize
p_rarity_gen <- pokemon %>%
  group_by(gen, rarity) %>%
  summarise(n = n(), .groups = "drop") %>%
  ggplot(aes(x = gen, y = n, fill = rarity)) +
  geom_col(position = "fill") +
  labs(title = "Rarity Composition by Generation",
       subtitle = "Does the proportion of legendaries increase over time?",
       x = "Generation", y = "Proportion", fill = "Rarity") +
  theme_minimal()
print(p_rarity_gen)
```

**Findings:** If later generations have higher proportions of legendary/mythical Pokémon, this could explain the apparent power creep without a true baseline increase.

## 14.3 Power Creep Among Normal Pokémon Only

Test if power creep exists when excluding legendaries. Uses the pre-built `pokemon_normal` dataset.

```{r normal-only}
# pokemon_normal was created in Section 2 - contains only Normal rarity Pokémon
cat("--- Attack by Generation (Normal Pokémon Only) ---\n")
cat("Dataset size:", nrow(pokemon_normal), "Pokémon\n\n")

normal_summary <- pokemon_normal %>%
  group_by(gen) %>%
  summarise(n = n(), mean_attack = mean(attack), sd_attack = sd(attack))
print(normal_summary)

# Test generation effect on normal Pokémon (simpler model - no rarity needed)
model_normal_check <- lm(attack ~ gen + hp + defense + speed, data = pokemon_normal)
cat("\n--- OLS on Normal Pokémon Only ---\n")
summary(model_normal_check)

# Trend correlation for normal only (already computed in 13.3, but shown here for clarity)
trend_normal_check <- cor(as.numeric(normal_summary$gen), normal_summary$mean_attack)
cat("\nCorrelation (gen vs mean_attack) for Normal Pokémon:", round(trend_normal_check, 3), "\n")

# Compare with full dataset
cat("\n--- Comparison with Full Dataset ---\n")
cat("Full data correlation:", round(trend, 3), "\n")
cat("Normal-only correlation:", round(trend_normal_check, 3), "\n")
cat("Difference:", round(trend - trend_normal_check, 3), "\n")
```

**Findings:** Normal-only correlation = 0.732, very close to full data (0.757). Power creep persists among Normal-rarity Pokémon — this is a genuine baseline shift, not driven by legendary inflation. The 0.025 difference is negligible.

## 14.4 Total Stats Analysis (Not Just Attack)

Does power creep affect overall strength, not just attack?

```{r total-stats-analysis}
cat("--- Total Stats by Generation ---\n")
total_summary <- pokemon %>%
  group_by(gen) %>%
  summarise(
    mean_total = mean(total_stats),
    mean_offensive = mean(offensive_power),
    mean_defensive = mean(defensive_power)
  )
print(total_summary)

# Correlation with generation
cat("\nCorrelations with generation:\n")
cat("  Total stats:", round(cor(as.numeric(total_summary$gen), total_summary$mean_total), 3), "\n")
cat("  Offensive power:", round(cor(as.numeric(total_summary$gen), total_summary$mean_offensive), 3), "\n")
cat("  Defensive power:", round(cor(as.numeric(total_summary$gen), total_summary$mean_defensive), 3), "\n")

# Model for total stats
model_total <- lm(total_stats ~ gen + rarity, data = pokemon)
cat("\n--- OLS: Total Stats ~ Generation + Rarity ---\n")
summary(model_total)
```

**Findings:** If total_stats also increases with generation, power creep is broad-based. If only attack increases, the "creep" might be a shift toward offensive designs rather than overall power inflation.

## 14.5 Generation Effect with Confidence Intervals

Visualize the uncertainty in generation effects.

```{r gen-ci-plot, fig.width=10, fig.height=6}
# Extract generation coefficients with CIs from full model
gen_effects <- tidy(model_full, conf.int = TRUE) %>%
  filter(grepl("^gen", term)) %>%
  mutate(
    generation = as.numeric(gsub("gen", "", term)),
    term = factor(term, levels = paste0("gen", 2:9))
  )

# Add Gen 1 as reference (0 effect)
gen_effects <- bind_rows(
  data.frame(term = "gen1", estimate = 0, std.error = 0, 
             conf.low = 0, conf.high = 0, generation = 1,
             statistic = NA, p.value = NA),
  gen_effects
)

p_gen_ci <- ggplot(gen_effects, aes(x = generation, y = estimate)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  geom_pointrange(aes(ymin = conf.low, ymax = conf.high), 
                  color = "steelblue", size = 0.8) +
  geom_line(color = "steelblue", alpha = 0.5) +
  labs(title = "Generation Effect on Attack (with 95% CI)",
       subtitle = "Controlling for HP, Defense, Speed, Rarity, etc.",
       x = "Generation", y = "Effect on Attack (vs Gen 1)") +
  scale_x_continuous(breaks = 1:9) +
  theme_minimal()
print(p_gen_ci)
```

**Findings:** This plot shows which generations are *significantly* different from Gen 1 (CIs not crossing zero). Wide CIs indicate uncertainty — the power creep claim should be tempered if many CIs overlap zero.

## 14.6 Summary of Verification Checks

```{r verification-summary}
cat("================================================================================\n")
cat("                    POWER CREEP VERIFICATION SUMMARY\n")
cat("================================================================================\n\n")

# Collect results
cat("1. MONOTONIC TREND (Jonckheere-Terpstra):\n")
cat("   p-value:", format(jt_test$p.value, digits = 4), "\n")
cat("   Interpretation:", ifelse(jt_test$p.value < 0.05, 
                                  "Significant monotonic increase", 
                                  "No consistent ordering"), "\n\n")

cat("2. RARITY CONFOUND:\n")
cat("   Chi-square p-value:", format(chi_test$p.value, digits = 4), "\n")
cat("   Interpretation:", ifelse(chi_test$p.value < 0.05,
                                  "Rarity distribution differs by generation (potential confound)",
                                  "Rarity distribution similar across generations"), "\n\n")

cat("3. NORMAL POKÉMON ONLY:\n")
cat("   Trend correlation:", round(trend_normal, 3), "\n")
cat("   Interpretation:", ifelse(abs(trend_normal) > 0.5,
                                  "Power creep exists even among normal Pokémon",
                                  "Power creep may be driven by legendaries"), "\n\n")

cat("4. TOTAL STATS (not just attack):\n")
total_trend <- cor(as.numeric(total_summary$gen), total_summary$mean_total)
cat("   Trend correlation:", round(total_trend, 3), "\n")
cat("   Interpretation:", ifelse(abs(total_trend) > 0.5,
                                  "Broad power creep (all stats)",
                                  "Creep specific to attack/offensive stats"), "\n\n")

cat("================================================================================\n")
```

---

# 15. Robustness Check: Analysis Without Outliers

This section repeats key analyses on the outlier-free dataset to verify that power creep findings are not driven by extreme values.

## 15.1 Outlier-Free Data Summary

```{r prep-clean-data}
# pokemon_clean was already preprocessed in Section 2 (factors and derived features)
# Here we just summarize the differences

cat("--- Outlier-Free Dataset Summary ---\n")
cat("Total Pokémon (Full):", nrow(pokemon), "\n")
cat("Total Pokémon (Clean):", nrow(pokemon_clean), "\n")
cat("Pokémon removed:", nrow(pokemon) - nrow(pokemon_clean), "\n")
cat("Percentage removed:", round(100 * (nrow(pokemon) - nrow(pokemon_clean)) / nrow(pokemon), 1), "%\n\n")

# Which generations lost the most Pokémon?
gen_comparison <- data.frame(
  gen = 1:9,
  full = as.numeric(table(pokemon$gen)),
  clean = as.numeric(table(pokemon_clean$gen))
) %>%
  mutate(removed = full - clean, pct_removed = round(100 * removed / full, 1))

cat("Pokémon removed by generation:\n")
print(gen_comparison)
```

**Findings:** This shows which generations had the most outliers removed. If later generations lose more outliers, it could affect the power creep analysis.

## 15.2 Descriptive Comparison: Attack by Generation

```{r clean-descriptive}
# Compare mean attack between full and clean datasets
attack_comparison <- data.frame(
  gen = 1:9,
  mean_full = pokemon %>% group_by(gen) %>% summarise(m = mean(attack)) %>% pull(m),
  mean_clean = pokemon_clean %>% group_by(gen) %>% summarise(m = mean(attack)) %>% pull(m)
) %>%
  mutate(
    difference = mean_full - mean_clean,
    pct_change = round(100 * difference / mean_full, 1)
  )

cat("--- Mean Attack Comparison (Full vs Outlier-Free) ---\n")
print(attack_comparison)

# Correlations
trend_full <- cor(1:9, attack_comparison$mean_full)
trend_clean <- cor(1:9, attack_comparison$mean_clean)

cat("\nCorrelation (gen vs mean_attack):\n")
cat("  Full dataset:", round(trend_full, 3), "\n")
cat("  Outlier-free:", round(trend_clean, 3), "\n")
```

**Findings:** If the correlation remains similar after removing outliers, the power creep signal is robust. Large changes suggest outliers were driving the effect.

## 15.3 Visualization: Full vs Outlier-Free

```{r clean-boxplot, fig.width=12, fig.height=5}
# Side-by-side boxplots
p_full <- ggplot(pokemon, aes(x = gen, y = attack, fill = gen)) +
  geom_boxplot(alpha = 0.7) +
  labs(title = "Full Dataset", x = "Generation", y = "Attack") +
  theme_minimal() +
  theme(legend.position = "none") +
  ylim(0, 200)

p_clean <- ggplot(pokemon_clean, aes(x = gen, y = attack, fill = gen)) +
  geom_boxplot(alpha = 0.7) +
  labs(title = "Outlier-Free Dataset", x = "Generation", y = "Attack") +
  theme_minimal() +
  theme(legend.position = "none") +
  ylim(0, 200)

gridExtra::grid.arrange(p_full, p_clean, ncol = 2, 
                        top = "Attack Distribution: Full vs Outlier-Free")
```

**Findings:** Visual comparison shows how outlier removal affects the distribution shape and median values per generation.

## 15.4 OLS Regression: Full vs Outlier-Free

```{r clean-ols}
# Full model on outlier-free data
model_clean <- lm(attack ~ gen + hp + defense + sp_attack + sp_defense + speed +
                    height_m + weight_kg + rarity,
                  data = pokemon_clean)

cat("--- OLS on Outlier-Free Data ---\n")
summary(model_clean)

# Compare key statistics
cat("\n--- Model Comparison: Full vs Outlier-Free ---\n")
cat("R² (Full):", round(summary(model_full)$r.squared, 3), "\n")
cat("R² (Clean):", round(summary(model_clean)$r.squared, 3), "\n")
```

**Findings:** Compare R² and generation coefficients between models. Similar results indicate robustness; large differences suggest outlier sensitivity.

## 15.5 Generation Coefficients Comparison

```{r clean-coef-comparison}
# Extract generation coefficients from both models
gen_coef_full <- coef(model_full)[grep("^gen", names(coef(model_full)))]
gen_coef_clean <- coef(model_clean)[grep("^gen", names(coef(model_clean)))]

coef_comparison <- data.frame(
  Generation = paste0("Gen ", 2:9),
  Full_Data = round(gen_coef_full, 2),
  Outlier_Free = round(gen_coef_clean, 2),
  Difference = round(gen_coef_full - gen_coef_clean, 2)
)
rownames(coef_comparison) <- NULL

cat("--- Generation Coefficients (vs Gen 1 baseline) ---\n")
print(coef_comparison)

# Average absolute difference
cat("\nMean absolute difference in gen coefficients:", 
    round(mean(abs(coef_comparison$Difference)), 2), "\n")
```

**Findings:** Small differences (< 2-3 attack points) indicate the generation effect is robust to outliers. Large differences suggest specific outliers were influential.

## 15.6 Likelihood Ratio Test: Outlier-Free Data

```{r clean-lr-test}
# Full model with generation (outlier-free)
full_model_clean <- glm(attack ~ gen + hp + defense + sp_attack + sp_defense + speed +
                          rarity + weight_kg, data = pokemon_clean, family = gaussian)

# Reduced model without generation (outlier-free)
reduced_model_clean <- glm(attack ~ hp + defense + sp_attack + sp_defense + speed +
                             rarity + weight_kg, data = pokemon_clean, family = gaussian)

# Likelihood ratio test
lr_test_clean <- lrtest(reduced_model_clean, full_model_clean)
cat("--- LR Test: Is Generation Significant? (Outlier-Free Data) ---\n")
print(lr_test_clean)

cat("\n--- Comparison of LR Test Results ---\n")
cat("Full data p-value:", format(lr_test$`Pr(>Chisq)`[2], digits = 4), "\n")
cat("Outlier-free p-value:", format(lr_test_clean$`Pr(>Chisq)`[2], digits = 4), "\n")

if (lr_test_clean$`Pr(>Chisq)`[2] < 0.05) {
  cat("\nConclusion: Generation remains significant after removing outliers.\n")
} else {
  cat("\nConclusion: Generation is NOT significant in outlier-free data — outliers may have driven the effect.\n")
}
```

**Findings:** LR test p = 0.0077 (full) and p = 0.002 (outlier-free). Both are significant at α = 0.05. This is the KEY robustness check — the generation effect is even MORE significant after removing outliers (p = 0.002 < 0.0077), indicating outliers were actually dampening the signal, not inflating it.

## 15.7 Monotonic Trend Test: Outlier-Free

```{r clean-monotonic}
# Reference the JT test results computed in Section 14.1
cat("--- Jonckheere-Terpstra Test Results (from Section 14.1) ---\n")
cat("\nFull data:\n")
cat("  JT statistic:", jt_test$statistic, "\n")
cat("  p-value:", format(jt_test$p.value, digits = 4), "\n")

cat("\nOutlier-free:\n")
cat("  JT statistic:", jt_test_clean$statistic, "\n")
cat("  p-value:", format(jt_test_clean$p.value, digits = 4), "\n")

cat("\n--- Interpretation ---\n")
if (jt_test$p.value < 0.05 && jt_test_clean$p.value < 0.05) {
  cat("Both datasets show significant monotonic trend — ROBUST.\n")
} else if (jt_test$p.value < 0.05 && jt_test_clean$p.value >= 0.05) {
  cat("Full data significant but outlier-free not — trend may be driven by outliers.\n")
} else {
  cat("Neither dataset shows significant monotonic trend.\n")
}
```

**Findings:** Compares whether the monotonic increasing trend holds after outlier removal.

## 15.8 Summary: Outlier Robustness

```{r clean-summary}
cat("================================================================================\n")
cat("              OUTLIER ROBUSTNESS SUMMARY\n")
cat("================================================================================\n\n")

cat("1. DATA REDUCTION:\n")
cat("   Pokémon removed:", nrow(pokemon) - nrow(pokemon_clean), 
    "(", round(100*(nrow(pokemon) - nrow(pokemon_clean))/nrow(pokemon), 1), "%)\n\n")

cat("2. CORRELATION (gen vs mean_attack):\n")
cat("   Full data:", round(trend_full, 3), "\n")
cat("   Outlier-free:", round(trend_clean, 3), "\n")
cat("   Change:", round(trend_clean - trend_full, 3), "\n\n")

cat("3. LR TEST (Is generation significant?):\n")
cat("   Full data: p =", format(lr_test$`Pr(>Chisq)`[2], digits = 4), 
    ifelse(lr_test$`Pr(>Chisq)`[2] < 0.05, "(significant)", "(not significant)"), "\n")
cat("   Outlier-free: p =", format(lr_test_clean$`Pr(>Chisq)`[2], digits = 4),
    ifelse(lr_test_clean$`Pr(>Chisq)`[2] < 0.05, "(significant)", "(not significant)"), "\n\n")

cat("4. MONOTONIC TREND (Jonckheere-Terpstra):\n")
cat("   Full data: p =", format(jt_test$p.value, digits = 4),
    ifelse(jt_test$p.value < 0.05, "(significant)", "(not significant)"), "\n")
cat("   Outlier-free: p =", format(jt_test_clean$p.value, digits = 4),
    ifelse(jt_test_clean$p.value < 0.05, "(significant)", "(not significant)"), "\n\n")

# Overall verdict
both_lr_sig <- lr_test$`Pr(>Chisq)`[2] < 0.05 && lr_test_clean$`Pr(>Chisq)`[2] < 0.05
both_jt_sig <- jt_test$p.value < 0.05 && jt_test_clean$p.value < 0.05

cat("5. ROBUSTNESS VERDICT:\n")
if (both_lr_sig && both_jt_sig) {
  cat("   ROBUST: Power creep finding holds in both full and outlier-free data.\n")
} else if (both_lr_sig) {
  cat("   PARTIALLY ROBUST: LR test significant in both, but monotonic trend differs.\n")
} else if (lr_test$`Pr(>Chisq)`[2] < 0.05 && lr_test_clean$`Pr(>Chisq)`[2] >= 0.05) {
  cat("   NOT ROBUST: Generation effect disappears after removing outliers.\n")
  cat("   Interpretation: Outliers (likely high-attack legendaries) drove the effect.\n")
} else {
  cat("   INCONCLUSIVE: Mixed results between datasets.\n")
}

cat("================================================================================\n")
```

---

# 16. Type-Specific Power Creep: Generation × Type Interactions

This section explores whether power creep affects different Pokémon types differently. Instead of fitting separate models for each type or rarity, we include **interaction terms** between generation and primary type (and rarity) in a regularized model. LASSO will automatically select which type-specific generational effects are significant.

**Motivation:** Your coworker suggested that the generation effect might be stronger for certain types (e.g., Fighting, Dragon) or rarities (e.g., legendary). By including `gen × type` interactions, we can detect differential power creep without fitting many separate models.

**Note:** Primary types are nominal (not ordinal) categories — there's no inherent ordering between Fire, Water, Grass, etc. We treat them as unordered factors.

**Type Encoding Reference:** class_primary integers map to: 0=Grass, 1=Fire, 2=Water, 3=Bug, 4=Normal, 5=Poison, 6=Electric, 7=Ground, 8=Fairy, 9=Fighting, 10=Psychic, 11=Rock, 12=Ghost, 13=Ice, 14=Dragon, 15=Dark, 16=Steel, 17=Flying.

## 16.1 Prepare Interaction Data

```{r interaction-prep}
# Create dataset with type interactions
# Note: class_primary has 18 levels (types), gen has 9 levels
# Full interaction would create 18*9 = 162 terms, but we use gen_numeric for simplicity

pokemon_interact <- pokemon %>%
  mutate(
    # Ensure factors are properly set
    class_primary = as.factor(class_primary),
    rarity = as.factor(rarity),
    gen = as.factor(gen)
  ) %>%
  dplyr::select(attack, total_stats, gen, gen_numeric, class_primary, class_secondary,
                rarity, hp, defense, sp_attack, sp_defense, speed, 
                weight_kg, height_m, num_abilities, evo_length,
                has_mega_evolution, has_gigantamax) %>%
  na.omit()

cat("--- Type Distribution ---\n")
print(table(pokemon_interact$class_primary))

cat("\n--- Rarity Distribution ---\n")
print(table(pokemon_interact$rarity))

# Check cross-tabulation: types × generations
cat("\n--- Sample sizes per Type × Generation (first 6 types) ---\n")
type_gen_table <- table(pokemon_interact$class_primary, pokemon_interact$gen)
print(type_gen_table[1:6, ])
```

**Note:** Some type × generation combinations have small sample sizes (< 10), which may limit the power to detect interactions for rare types.

## 16.2 LASSO with Generation × Type Interactions

We create a design matrix that includes:
- Main effects: generation, type, rarity, combat stats
- Interactions: `gen_numeric × class_primary` and `gen_numeric × rarity`

LASSO will select which interactions have non-zero coefficients.

```{r lasso-interaction}
# Create design matrix with interactions
# Using gen_numeric (1-9) instead of factor gen for cleaner interaction interpretation
X_interact <- model.matrix(
  ~ gen_numeric * class_primary + gen_numeric * rarity + 
    hp + defense + sp_attack + sp_defense + speed + weight_kg - 1,
  data = pokemon_interact
)

y_interact <- pokemon_interact$attack

cat("Design matrix dimensions:", dim(X_interact), "\n")
cat("Number of interaction terms:", sum(grepl(":", colnames(X_interact))), "\n")

# Fit LASSO with cross-validation
set.seed(42)
cv_lasso_interact <- cv.glmnet(X_interact, y_interact, alpha = 1, nfolds = 10)

cat("\n--- LASSO with Interactions ---\n")
cat("Optimal lambda (min):", cv_lasso_interact$lambda.min, "\n")
cat("Optimal lambda (1se):", cv_lasso_interact$lambda.1se, "\n")

# Get coefficients at optimal lambda
lasso_interact_model <- glmnet(X_interact, y_interact, alpha = 1, 
                                lambda = cv_lasso_interact$lambda.min)
lasso_interact_coef <- coef(lasso_interact_model)

# Extract non-zero coefficients
nonzero_coef <- lasso_interact_coef[lasso_interact_coef[,1] != 0, ]
cat("\nNumber of features selected:", length(nonzero_coef) - 1, "out of", ncol(X_interact), "\n")
```

**Findings:** LASSO selected 28 out of 48 potential features. The regularization successfully identified which main effects and interactions contribute to predicting attack, shrinking irrelevant terms to zero.

## 16.3 Selected Interaction Terms

```{r selected-interactions}
# Separate main effects from interactions
coef_df <- data.frame(
  term = names(nonzero_coef),
  coefficient = as.numeric(nonzero_coef)
) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    is_interaction = grepl(":", term),
    type = case_when(
      grepl("class_primary", term) & is_interaction ~ "Gen × Type",
      grepl("rarity", term) & is_interaction ~ "Gen × Rarity",
      grepl("class_primary", term) ~ "Type (main)",
      grepl("rarity", term) ~ "Rarity (main)",
      grepl("gen_numeric", term) & !is_interaction ~ "Generation (main)",
      TRUE ~ "Combat stat"
    )
  ) %>%
  arrange(desc(abs(coefficient)))

cat("--- Selected Coefficients by Category ---\n")

# Main effects (combat stats)
cat("\n1. Combat Stats (main effects):\n")
coef_df %>% filter(type == "Combat stat") %>% print()

# Generation main effect
cat("\n2. Generation Main Effect:\n")
coef_df %>% filter(type == "Generation (main)") %>% print()

# Type main effects
cat("\n3. Type Main Effects (selected by LASSO):\n")
coef_df %>% filter(type == "Type (main)") %>% print()

# Rarity main effects
cat("\n4. Rarity Main Effects:\n")
coef_df %>% filter(type == "Rarity (main)") %>% print()

# KEY: Interaction terms
cat("\n5. Gen × Type INTERACTIONS (selected by LASSO):\n")
gen_type_interact <- coef_df %>% filter(type == "Gen × Type")
if(nrow(gen_type_interact) > 0) {
  print(gen_type_interact)
} else {
  cat("   No Gen × Type interactions selected (all shrunk to zero)\n")
}

cat("\n6. Gen × Rarity INTERACTIONS (selected by LASSO):\n")
gen_rarity_interact <- coef_df %>% filter(type == "Gen × Rarity")
if(nrow(gen_rarity_interact) > 0) {
  print(gen_rarity_interact)
} else {
  cat("   No Gen × Rarity interactions selected (all shrunk to zero)\n")
}
```

**Findings:** LASSO selected 4 Gen × Type interactions and 1 Gen × Rarity interaction:

- **gen_numeric:class_primary5 (Poison):** –0.42 → Power creep is *weaker* for Poison types
- **gen_numeric:class_primary4 (Normal):** +0.39 → Power creep is *stronger* for Normal (basic) types
- **gen_numeric:class_primary8 (Fairy):** –0.13 → Power creep is *weaker* for Fairy types
- **gen_numeric:class_primary17 (Flying):** +0.02 → Slightly stronger for Flying types
- **gen_numeric:raritySubLegendary:** +0.46 → SubLegendary Pokémon experience *stronger* power creep

**Interpretation:** The selection of only 4/18 type interactions suggests power creep is relatively uniform across most types. The notable exceptions are Normal types (experiencing above-average power creep) and Poison/Fairy types (experiencing below-average power creep). SubLegendary Pokémon show the strongest differential effect.

## 16.4 Visualize Type-Specific Trends

Let's visualize how mean attack evolves by type across generations to contextualize the LASSO findings.

```{r type-trends, fig.width=14, fig.height=10}
# Calculate mean attack by type and generation
type_trends <- pokemon_interact %>%
  group_by(class_primary, gen) %>%
  summarise(
    n = n(),
    mean_attack = mean(attack),
    se_attack = sd(attack) / sqrt(n()),
    .groups = "drop"
  )

# Select top 9 most common types for visualization
top_types <- pokemon_interact %>%
  count(class_primary, sort = TRUE) %>%
  head(9) %>%
  pull(class_primary)

type_trends_top <- type_trends %>%
  filter(class_primary %in% top_types)

# Plot: Mean attack by generation, faceted by type
p_type_trends <- ggplot(type_trends_top, 
                         aes(x = as.numeric(as.character(gen)), y = mean_attack)) +
  geom_point(aes(size = n), alpha = 0.7, color = "steelblue") +
  geom_line(alpha = 0.5, color = "steelblue") +
  geom_smooth(method = "lm", se = FALSE, linetype = "dashed", color = "red", size = 0.8) +
  facet_wrap(~ class_primary, ncol = 3, scales = "free_y") +
  scale_x_continuous(breaks = 1:9) +
  labs(title = "Mean Attack by Generation for Top 9 Primary Types",
       subtitle = "Red dashed line = linear trend; Point size = sample size",
       x = "Generation", y = "Mean Attack", size = "N") +
  theme_minimal() +
  theme(legend.position = "bottom")
print(p_type_trends)
```

```{r type-slopes}
# Calculate per-type slopes (correlation between gen and mean_attack)
type_slopes <- type_trends %>%
  group_by(class_primary) %>%
  summarise(
    n_total = sum(n),
    n_gens = n(),
    correlation = cor(as.numeric(as.character(gen)), mean_attack),
    .groups = "drop"
  ) %>%
  filter(n_gens >= 5) %>%  # Only types present in at least 5 generations
  arrange(desc(correlation))

cat("--- Power Creep Strength by Type (correlation: gen vs mean_attack) ---\n")
cat("Positive = attack increases with generation; Negative = attack decreases\n\n")
print(type_slopes)

cat("\n--- Types with STRONGEST power creep (cor > 0.5): ---\n")
strong_creep <- type_slopes %>% filter(correlation > 0.5)
if(nrow(strong_creep) > 0) print(strong_creep) else cat("None\n")

cat("\n--- Types with REVERSE power creep (cor < -0.3): ---\n")
reverse_creep <- type_slopes %>% filter(correlation < -0.3)
if(nrow(reverse_creep) > 0) print(reverse_creep) else cat("None\n")
```

**Findings:** The type-specific correlation analysis reveals clear patterns:

- **Strongest power creep (cor > 0.5):** Fighting (0.76), Ice (0.68), Grass (0.62), Fairy (0.60) — these types show consistent attack increases across generations
- **Moderate power creep:** Water (0.50), Steel (0.49), Ground (0.47), Electric (0.45)
- **Weakest/flat trends:** Rock (–0.20), Dragon (–0.18), Psychic (–0.09), Fire (–0.004) — these types show no meaningful power creep or slight decreases
- **No reverse power creep:** No types have correlation < –0.3, meaning no type is systematically getting weaker

Interestingly, **Dragon** — often considered a "power" type — shows slight *negative* trend (–0.18), possibly because early-gen Dragons set a high baseline that later generations haven't exceeded.

## 16.5 Rarity-Specific Power Creep

Does power creep differ by rarity? Are legendaries experiencing more/less power creep than normal Pokémon?

```{r rarity-trends, fig.width=12, fig.height=5}
# Calculate mean attack by rarity and generation
rarity_trends <- pokemon_interact %>%
  group_by(rarity, gen) %>%
  summarise(
    n = n(),
    mean_attack = mean(attack),
    mean_total_stats = mean(total_stats),
    .groups = "drop"
  )

# Plot: Attack trends by rarity
p_rarity_attack <- ggplot(rarity_trends, 
                           aes(x = as.numeric(as.character(gen)), y = mean_attack, 
                               color = rarity, shape = rarity)) +
  geom_point(aes(size = n), alpha = 0.7) +
  geom_line(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, linetype = "dashed", size = 0.8) +
  scale_x_continuous(breaks = 1:9) +
  labs(title = "Mean Attack by Generation and Rarity",
       subtitle = "Dashed lines = linear trends",
       x = "Generation", y = "Mean Attack", 
       color = "Rarity", shape = "Rarity", size = "N") +
  theme_minimal() +
  theme(legend.position = "bottom")

# Plot: Total stats trends by rarity
p_rarity_total <- ggplot(rarity_trends, 
                          aes(x = as.numeric(as.character(gen)), y = mean_total_stats, 
                              color = rarity, shape = rarity)) +
  geom_point(aes(size = n), alpha = 0.7) +
  geom_line(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, linetype = "dashed", size = 0.8) +
  scale_x_continuous(breaks = 1:9) +
  labs(title = "Mean Total Stats by Generation and Rarity",
       subtitle = "Dashed lines = linear trends",
       x = "Generation", y = "Mean Total Stats", 
       color = "Rarity", shape = "Rarity", size = "N") +
  theme_minimal() +
  theme(legend.position = "bottom")

gridExtra::grid.arrange(p_rarity_attack, p_rarity_total, ncol = 2)
```

```{r rarity-slopes}
# Calculate per-rarity slopes
rarity_slopes <- rarity_trends %>%
  group_by(rarity) %>%
  summarise(
    n_total = sum(n),
    cor_attack = cor(as.numeric(as.character(gen)), mean_attack),
    cor_total = cor(as.numeric(as.character(gen)), mean_total_stats),
    .groups = "drop"
  )

cat("--- Power Creep by Rarity ---\n")
cat("cor_attack = correlation of gen with mean attack\n")
cat("cor_total = correlation of gen with mean total stats\n\n")
print(rarity_slopes)
```

**Findings:** The rarity-specific analysis reveals a striking pattern:

| Rarity | Attack Trend | Total Stats Trend | Interpretation |
|--------|-------------|-------------------|----------------|
| Normal | cor = 0.73 | cor = 0.82 | **Strong power creep** — baseline is rising |
| SubLegendary | cor = 0.71 | cor = –0.64 | Attack ↑ but total stats ↓ → more offense-focused |
| Legendary | cor = 0.09 | cor = –0.44 | Flat attack, declining total → no power creep |
| Mythical | cor = –0.16 | cor = –0.63 | Slightly weaker over time |

**Key insight:** Power creep is concentrated in **Normal** and **SubLegendary** categories. Legendary and Mythical Pokémon show flat or negative trends — they're not getting stronger. This confirms power creep is a *baseline* shift affecting regular Pokémon, not "legendary inflation." SubLegendaries are becoming more offense-focused (attack increasing while total stats decrease).

## 16.6 GLM with Explicit Interactions (OLS)

For interpretability, let's fit an OLS model with explicit interaction terms for the most important types.

```{r glm-interactions}
# Focus on types with potential differential power creep
# Use numeric generation for cleaner interpretation
model_interact_ols <- lm(
  attack ~ gen_numeric * class_primary + gen_numeric * rarity +
    hp + defense + sp_attack + sp_defense + speed,
  data = pokemon_interact
)

cat("--- OLS with Gen × Type and Gen × Rarity Interactions ---\n")
cat("(Showing only significant interactions at p < 0.1)\n\n")

# Extract summary and filter significant interactions
interact_summary <- tidy(model_interact_ols) %>%
  filter(grepl(":", term)) %>%  # Only interaction terms
  filter(p.value < 0.1) %>%     # Marginally significant
  arrange(p.value)

if(nrow(interact_summary) > 0) {
  print(interact_summary)
} else {
  cat("No significant interactions at p < 0.1\n")
}

# ANOVA: Test if interactions add significant explanatory power
cat("\n--- ANOVA: Do Type Interactions Improve the Model? ---\n")
model_no_type_interact <- lm(
  attack ~ gen_numeric + class_primary + gen_numeric * rarity +
    hp + defense + sp_attack + sp_defense + speed,
  data = pokemon_interact
)
anova_type <- anova(model_no_type_interact, model_interact_ols)
print(anova_type)

cat("\n--- ANOVA: Do Rarity Interactions Improve the Model? ---\n")
model_no_rarity_interact <- lm(
  attack ~ gen_numeric * class_primary + rarity +
    hp + defense + sp_attack + sp_defense + speed,
  data = pokemon_interact
)
anova_rarity <- anova(model_no_rarity_interact, model_interact_ols)
print(anova_rarity)
```

**Findings:** The ANOVA tests reveal:

- **Type Interactions:** F = 0.998, p = 0.457 → Adding all 17 type interaction terms does NOT significantly improve the model. While LASSO selected 4 individual interactions, the *overall* set of type interactions doesn't add meaningful explanatory power.
- **Rarity Interactions:** The SubLegendary interaction selected by LASSO represents the most meaningful moderating effect.

**Interpretation:** Power creep is relatively **uniform** across types at the population level. The type-specific deviations identified by LASSO (Normal +0.39, Poison –0.42) represent subtle modulations rather than dramatic differences. This is actually a stronger finding — it means power creep is a systematic, broad-based phenomenon affecting the entire Pokémon roster, not driven by specific types becoming overpowered.

## 16.7 Summary: Type-Specific Power Creep

```{r interaction-summary, results='asis'}
# Compute summary statistics
n_type_interactions_selected <- nrow(gen_type_interact)
n_rarity_interactions_selected <- nrow(gen_rarity_interact)
strongest_type <- type_slopes$class_primary[1]
strongest_type_cor <- round(type_slopes$correlation[1], 3)
weakest_type <- type_slopes$class_primary[nrow(type_slopes)]
weakest_type_cor <- round(type_slopes$correlation[nrow(type_slopes)], 3)

cat("
================================================================================
              TYPE-SPECIFIC POWER CREEP SUMMARY
================================================================================

1. LASSO FEATURE SELECTION:
   - Gen × Type interactions selected: ", n_type_interactions_selected, "
   - Gen × Rarity interactions selected: ", n_rarity_interactions_selected, "
   - Interpretation: ", ifelse(n_type_interactions_selected == 0, 
     "Power creep is relatively UNIFORM across types (no type-specific effects)",
     paste0("Certain types have differential power creep: ", 
            paste(gen_type_interact$term, collapse = ", "))), "

2. TYPE-SPECIFIC TRENDS:
   - Strongest power creep: ", strongest_type, " (cor = ", strongest_type_cor, ")
   - Weakest/reverse creep: ", weakest_type, " (cor = ", weakest_type_cor, ")
   - These correlations measure how strongly mean attack increases with generation.

3. RARITY-SPECIFIC TRENDS:
   - Normal Pokémon: cor = ", round(rarity_slopes$cor_attack[rarity_slopes$rarity == "Normal"], 3), "
   - Legendaries: cor = ", round(mean(rarity_slopes$cor_attack[rarity_slopes$rarity != "Normal"]), 3), " (avg of rare categories)
   - Interpretation: ", ifelse(
     abs(rarity_slopes$cor_attack[rarity_slopes$rarity == "Normal"]) > 0.5,
     "Power creep exists among NORMAL Pokémon — not just legendary inflation",
     "Power creep signal is weak among Normal Pokémon"), "

4. KEY INSIGHT:
   ", ifelse(n_type_interactions_selected == 0 && n_rarity_interactions_selected == 0,
     "Neither type nor rarity significantly moderates the generation effect.
   Power creep, to the extent it exists, is a broad phenomenon affecting 
   all types and rarities similarly.",
     "Some types/rarities experience differential power creep.
   Consider stratified analyses for the most affected groups."), "

================================================================================
")
```

**Conclusion:** This analysis tests whether power creep is uniform across types and rarities or concentrated in specific groups. The LASSO feature selection approach is more efficient than fitting separate models — it automatically identifies which interaction terms are worth keeping. If no interactions are selected, power creep is a broad phenomenon; if specific interactions emerge, those types/rarities are driving the trend.

---

# Conclusions

```{r conclusions, results='asis'}
# Compute additional metrics for conclusion
jt_sig <- jt_test$p.value < 0.05
jt_clean_sig <- jt_test_clean$p.value < 0.05
lr_sig <- lr_test$`Pr(>Chisq)`[2] < 0.05
lr_clean_sig <- lr_test_clean$`Pr(>Chisq)`[2] < 0.05
rarity_confound <- chi_test$p.value < 0.05
normal_trend_strong <- abs(trend_normal) > 0.5
total_trend <- cor(as.numeric(total_summary$gen), total_summary$mean_total)
outlier_robust <- lr_sig && lr_clean_sig

conclusion_text <- glue::glue(
"================================================================================
                              CONCLUSIONS
================================================================================

1. DESCRIPTIVE ANALYSIS:
   - Dataset contains {nrow(pokemon)} Pokémon across {length(unique(pokemon$gen))} generations
   - Mean attack ranges from {round(min(mean_by_gen$mean_attack), 1)} (Gen 2) to {round(max(mean_by_gen$mean_attack), 1)} (Gen 7)
   - Correlation (gen vs mean_attack): {round(trend, 3)}
   - Pattern is NOT monotonic: Gen 2 and Gen 6 dip below Gen 1

2. REGRESSION ANALYSIS:
   - OLS R² (full model): {round(summary(model_full)$r.squared, 3)}
   - Generation effect p-value (LR test): {format(lr_test$`Pr(>Chisq)`[2], scientific = TRUE)}
   - After controls, only Gen 5 and Gen 7 remain clearly elevated

3. ABLATION STUDY (10-fold CV):
   - Best model: {names(rmse_ranking)[1]} (RMSE: {round(rmse_ranking[1], 3)})
   - Generation-only model barely beats null (R² ≈ 0.03)
   - Combat stats (HP, Defense, Speed) drive most predictive power

4. REGULARIZATION (LASSO):
   - Features selected: {n_selected}
   - LASSO retained generation dummies: {sum(grepl('gen', names(lasso_coef[lasso_coef[,1] != 0, ])))}

5. TIME-SPLIT CV (Train Gen 1-5, Test Gen 6-9):
   - OLS RMSE on future generations: {round(rmse_lm, 2)}
   - Performance degrades from CV (~21) to time-split (~24)
   - Future generations don't perfectly follow past patterns

6. VERIFICATION CHECKS:
   - Monotonic trend test: {ifelse(jt_sig, 'SIGNIFICANT', 'NOT significant')}
   - Rarity confound: {ifelse(rarity_confound, 'YES - rarity distribution differs by generation', 'NO')}
   - Normal Pokémon trend: {round(trend_normal, 3)} ({ifelse(normal_trend_strong, 'strong', 'weak')})
   - Total stats trend: {round(total_trend, 3)}

7. OUTLIER ROBUSTNESS:
   - Pokémon removed as outliers: {nrow(pokemon) - nrow(pokemon_clean)} ({round(100*(nrow(pokemon) - nrow(pokemon_clean))/nrow(pokemon), 1)}%)
   - LR test (full data): p = {format(lr_test$`Pr(>Chisq)`[2], digits = 4)} ({ifelse(lr_sig, 'significant', 'not significant')})
   - LR test (outlier-free): p = {format(lr_test_clean$`Pr(>Chisq)`[2], digits = 4)} ({ifelse(lr_clean_sig, 'significant', 'not significant')})
   - Robustness: {ifelse(outlier_robust, 'ROBUST - effect holds without outliers', 'NOT ROBUST - outliers drive the effect')}

8. TYPE-SPECIFIC POWER CREEP (Section 16):
   - Gen × Type interactions selected by LASSO: {n_type_interactions_selected}
   - Gen × Rarity interactions selected by LASSO: {n_rarity_interactions_selected}
   - Strongest type-specific trend: {strongest_type} (cor = {strongest_type_cor})
   - Interpretation: {ifelse(n_type_interactions_selected == 0, 
     'Power creep is UNIFORM across types',
     'Some types experience differential power creep')}

9. POWER CREEP VERDICT:
{if (jt_sig && lr_sig && normal_trend_strong && outlier_robust) {
  '   STRONG EVIDENCE FOR POWER CREEP:
   - Statistically significant after controlling for other factors
   - Monotonic trend confirmed
   - Effect persists among Normal-rarity Pokémon
   - Effect is ROBUST to outlier removal'
} else if (lr_sig && outlier_robust) {
  '   MODERATE EVIDENCE FOR POWER CREEP:
   - Generation is statistically significant (p < 0.05)
   - Effect is robust to outlier removal
   - BUT: Effect is not perfectly monotonic (Gen 2, 6 are lower)
   - BUT: Much of the effect is mediated by other combat stats
   - Interpretation: Later generations tend to have higher attack,
     but this is partly because they have higher HP/Defense/Speed too'
} else if (lr_sig && !outlier_robust) {
  '   WEAK EVIDENCE FOR POWER CREEP:
   - Generation is significant in full data (p < 0.05)
   - BUT: Effect disappears after removing outliers
   - Interpretation: Power creep may be driven by extreme Pokémon (legendaries, etc.),
     not a general trend across all Pokémon'
} else {
  '   NO STRONG EVIDENCE: After controlling for type, rarity, and stats,
   generation does not significantly predict attack.'
}}

================================================================================")

cat(conclusion_text)
```

---

# 17. Final Summary: Power Creep Across All Datasets

This section provides a consolidated summary of findings across all three datasets.

```{r final-summary, results='asis'}
# Compute final summary statistics
summary_table <- data.frame(
  Metric = c(
    "Sample Size",
    "Mean Attack Correlation (gen vs mean)",
    "Basic OLS R² (gen only)",
    "Full OLS R² (with controls)",
    "LR Test p-value",
    "Jonckheere-Terpstra p-value",
    "Power Creep Detected?"
  ),
  Full_Data = c(
    nrow(pokemon),
    round(trend, 3),
    round(summary(model_basic)$r.squared, 4),
    round(summary(model_full)$r.squared, 4),
    format(lr_test$`Pr(>Chisq)`[2], digits = 4),
    format(jt_test$p.value, digits = 4),
    ifelse(lr_test$`Pr(>Chisq)`[2] < 0.05 && jt_test$p.value < 0.05, "YES", "NO")
  ),
  Outlier_Free = c(
    nrow(pokemon_clean),
    round(trend_clean, 3),
    round(summary(model_basic_clean)$r.squared, 4),
    round(summary(model_full_clean)$r.squared, 4),
    format(lr_test_clean$`Pr(>Chisq)`[2], digits = 4),
    format(jt_test_clean$p.value, digits = 4),
    ifelse(lr_test_clean$`Pr(>Chisq)`[2] < 0.05 && jt_test_clean$p.value < 0.05, "YES", "NO")
  ),
  Normal_Only = c(
    nrow(pokemon_normal),
    round(trend_normal, 3),
    round(summary(model_basic_normal)$r.squared, 4),
    round(summary(model_full_normal)$r.squared, 4),
    "N/A (no rarity var)",
    "N/A",
    ifelse(abs(trend_normal) > 0.5, "YES (strong trend)", "WEAK")
  )
)

cat("## Summary Table: Power Creep Evidence Across Datasets\n\n")
knitr::kable(summary_table, align = "lrrr")
```

## 17.1 Key Conclusions

```{r key-conclusions, results='asis'}
cat("
### Evidence FOR Power Creep:

1. **Statistically Significant**: LR test confirms generation is a significant predictor 
   (p = 0.0077 full, p = 0.002 outlier-free) even after controlling for HP, Defense, 
   Speed, Rarity, and Weight.

2. **Monotonic Trend Confirmed**: Jonckheere-Terpstra test shows a significant 
   increasing trend (p = 2.09e-05 full, p = 1.61e-05 outlier-free).

3. **Robust to Outliers**: The effect is STRONGER after removing outliers 
   (p = 0.002 < 0.0077), meaning outliers were dampening, not inflating, the signal.

4. **Not Driven by Legendaries**: Normal-only correlation (0.732) is nearly identical 
   to full data (0.757). Power creep exists among regular Pokémon.

5. **Consistent Across Datasets**: All three datasets (full, outlier-free, normal-only) 
   show the same pattern of increasing attack across generations.

6. **Type-Specific Analysis (Section 16)**: Using LASSO with Gen × Type interactions, 
   we tested whether certain types experience stronger power creep. If LASSO selected 
   few/no interaction terms, power creep is uniform across types rather than driven 
   by specific types like Dragon or Fighting.

### Caveats and Limitations:

1. **Non-Monotonic Pattern**: Gen 2 and Gen 6 have LOWER mean attack than Gen 1. 
   Power creep is a general trend, not a strict generation-by-generation increase.

2. **Small Effect Size**: Generation alone explains only ~3% of variance. 
   Combat stats (HP, Defense, Speed) are far more predictive.

3. **Rarity Confound**: Later generations have different rarity distributions 
   (chi-square test significant). However, the Normal-only analysis addresses this.

4. **Mediation**: Much of the generation effect is mediated by other combat stats — 
   later Pokémon have higher attack partly because they also have higher HP/Defense/Speed.

### Final Verdict:

**MODERATE-TO-STRONG EVIDENCE FOR POWER CREEP**

- Generation is a statistically significant predictor of Attack
- The effect is robust to outlier removal
- The effect persists among Normal-rarity Pokémon (not just legendaries)
- The trend is monotonically increasing (Jonckheere-Terpstra significant)

However, the effect is small in magnitude (~1-2 attack points per generation) and 
the pattern has exceptions (Gen 2, Gen 6 dip below Gen 1). Power creep is real but 
subtle — later generations are slightly stronger on average, but the difference is 
modest compared to within-generation variation.
")
```

---

# Session Info

```{r session-info}
sessionInfo()
```

